### 2021-09-09 README.Rmd
Initial commit

---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-",
  fig.retina = 3
)
```

# Causal inference in nonprofit management and public administration

<!-- badges: start -->
<!-- badges: end -->

### Note on "corrupt caterpillar" project name

Because project titles change all the time with revisions, rewriting, and peer review, we used [**codename**](http://svmiller.com/codename/) to generate an [Ubuntu-style](https://wiki.ubuntu.com/DevelopmentCodeNames) internal-to-us project name that won't change. The seed `1022` was generated with random atmospheric noise at [random.org](https://www.random.org/).

```{r generate-code-name}
library(codename)
codename_message()

codename(type = "ubuntu", seed = 1022)
```

## Licenses

**Text and figures:** All prose and images are licensed under Creative Commons ([CC-BY-4.0](http://creativecommons.org/licenses/by/4.0/)).

**Code:** All code is licensed under the [MIT License](LICENSE.md).

## Contributions and Code of Conduct

We welcome contributions from everyone. Before you get started, please see our [contributor guidelines](CONTRIBUTING.md). Please note that this project is released with a [Contributor Code of Conduct](https://contributor-covenant.org/version/2/0/CODE_OF_CONDUCT.html). By contributing to this project, you agree to abide by its terms.

### 2021-09-09 README.Rmd
{targets}ification

@@
 
-```{r, include = FALSE}
+```{r setup, include=FALSE}
 knitr::opts_chunk$set(
@@
 )
+
+library(emoji)
 ```
@@
 
-### Note on "corrupt caterpillar" project name
+### `r emoji_glue(":devil::bug:")`: Note on "corrupt caterpillar" project name
 

### 2021-09-09 _targets.R

@@ -1,0 +1,29 @@
+library(targets)
+library(tarchetypes)
+
+# General pipeline settings -----------------------------------------------
+
+options(tidyverse.quiet = TRUE,
+        dplyr.summarise.inform = FALSE)
+
+set.seed(4393)  # From random.org
+
+
+# R functions -------------------------------------------------------------
+
+# source("R/whatever.R")
+
+# here::here() returns an absolute path, which then gets stored in tar_meta and
+# becomes computer-specific (i.e. /Users/andrew/Research/blah/thing.Rmd).
+# There's no way to get a relative path directly out of here::here(), but
+# fs::path_rel() works fine with it (see
+# https://github.com/r-lib/here/issues/36#issuecomment-530894167)
+here_rel <- function(...) {fs::path_rel(here::here(...))}
+
+
+# Pipeline ----------------------------------------------------------------
+
+list(
+  # Knit the README
+  tar_render(readme, here_rel("README.Rmd"), output_options = list(html_preview = FALSE))
+)

### 2021-09-09 manuscript/manuscript.Rmd
pandocification

---
title: "Full title goes here"
short-title: "Shorter title goes here"
author:
- name: Andrew Heiss
  affiliation: Georgia State University
  email: aheiss@gsu.edu
  url: https://www.andrewheiss.com/
- name: Meng Ye
  affiliation: Georgia State University
  email: schaudhry@lclark.edu
date: "September 10, 2021"
published: Drafting
code-repo: "Access the code at <https://github.com/andrewheiss/corrupt-caterpillar>"
abstract: >-
  Abstract goes here.
keywords: comma, separated, keywords, here
word-count: ""
thanks: >-
  Acknowledgments here if any
reference-section-title: References
bibliography: bibliography.bib
toc: false
title-page: true
endnotes: false
endfloat: false
mainfont: Cochineal
sansfont: Inter
author-note: >-
  Author note here
correspondence: >-
  Correspondence note here
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
# Figure out output format
is_docx <- knitr::pandoc_to("docx") | knitr::pandoc_to("odt")
is_latex <- knitr::pandoc_to("latex")
is_html <- knitr::pandoc_to("html")

# Word-specific things
table_format <- ifelse(is_docx, "huxtable", "kableExtra")  # Huxtable tables
conditional_dpi <- ifelse(is_docx, 300, 300)  # Higher DPI
conditional_align <- ifelse(is_docx, "default", "center")  # Word doesn't support align

# Knitr options
knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE,
  tidy.opts = list(width.cutoff = 120),  # Code width
  fig.retina = 3, dpi = conditional_dpi,
  fig.width = 7, fig.asp = 0.618,
  fig.align = conditional_align, out.width = "100%",
  fig.path = "output/figures/",
  cache.path = "output/_cache/",
  fig.process = function(x) {  # Remove "-1" from figure names
    x2 = sub('-\\d+([.][a-z]+)$', '\\1', x)
    if (file.rename(x, x2)) x2 else x
  }
)

# R options
options(
  width = 90,  # Output width
  dplyr.summarise.inform = FALSE,  # Turn off dplyr's summarize() auto messages
  knitr.kable.NA = "",  # Make NAs blank in kables
  kableExtra.latex.load_packages = FALSE,  # Don't add LaTeX preamble stuff
  modelsummary_default = table_format  # Set modelsummary backend
)
```

```{r libraries-data, include=FALSE}
library(tidyverse)
library(targets)
library(kableExtra)
library(here)

withr::with_dir(here::here(), {
  # Load stuff from targets
})
```

Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua [@BaBerrettCoupet:2021]. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.

### 2021-09-09 R/funs_knitting.R

@@ -1,0 +1,218 @@
+library(dplyr)
+library(purrr)
+library(stringr)
+library(bib2df)
+library(rvest)
+library(xml2)
+library(stringi)
+
+render_html <- function(input, output, csl, support_folder, ...) {
+  # Add CSS file as a dependency so it goes into the _files directory
+  dep <- htmltools::htmlDependency(
+    name = "ath",
+    version = "1.0.0",
+    "pandoc/css",
+    stylesheet = "ath-clean.css"
+  )
+  extra_dependencies <- list(dep)
+
+  # IMPORTANT
+  # When knitting to docx, bookdown deletes the _files directory for whatever
+  # reason, so if you knit to HTML and then docx, you get a nice *_files
+  # directly that then disappears when the Word file is done. One way around
+  # this is to specify lib_dir here in rmarkdown::render()
+
+  out <- rmarkdown::render(
+    input = input,
+    output_file = output,
+    bookdown::html_document2(
+      template = "pandoc/templates/html.html",
+      pandoc_args = c("--metadata", "link-citations=true",
+                      "--metadata", "linkReferences=true",
+                      paste0("--csl=", csl)),
+      md_extensions = "+raw_tex+smart-autolink_bare_uris+ascii_identifiers",
+      toc = TRUE,
+      number_sections = FALSE,
+      self_contained = FALSE,
+      theme = NULL,
+      extra_dependencies = extra_dependencies,
+      lib_dir = support_folder
+    ),
+    encoding = "UTF-8"
+  )
+
+  return(fs::path_rel(out))
+}
+
+render_pdf <- function(input, output, bibstyle, ...) {
+  out <- rmarkdown::render(
+    input = input,
+    output_file = output,
+    bookdown::pdf_document2(
+      template = "pandoc/templates/xelatex.tex",
+      latex_engine = "xelatex",
+      dev = "cairo_pdf",
+      pandoc_args = c("--top-level-division=section",
+                      "--shift-heading-level-by=0",
+                      "-V", bibstyle,
+                      "-V", "chapterstyle=hikma-article"),
+      md_extensions = "+raw_tex+smart-autolink_bare_uris",
+      toc = FALSE,
+      keep_tex = FALSE,
+      citation_package = "biblatex"
+    ),
+    encoding = "UTF-8"
+  )
+
+  return(fs::path_rel(out))
+}
+
+render_pdf_ms <- function(input, output, bibstyle, ...) {
+  out <- rmarkdown::render(
[…]

### 2021-09-09 _targets.R

@@ -2,6 +2,8 @@ library(targets)
 library(tarchetypes)
 
 # General pipeline settings -----------------------------------------------
+csl <- "pandoc/csl/apa.csl"
+bibstyle <- "bibstyle-apa"
 
 options(tidyverse.quiet = TRUE,
         dplyr.summarise.inform = FALSE)
@@ -12,6 +14,7 @@ set.seed(4393)  # From random.org
 # R functions -------------------------------------------------------------
 
 # source("R/whatever.R")
+source("R/funs_knitting.R")
 
 # here::here() returns an absolute path, which then gets stored in tar_meta and
 # becomes computer-specific (i.e. /Users/andrew/Research/blah/thing.Rmd).
@@ -24,6 +27,65 @@ here_rel <- function(...) {fs::path_rel(here::here(...))}
 # Pipeline ----------------------------------------------------------------
 
 list(
+
+  ## Manuscript targets ------------------------------------------------------
+
+  # tarchetypes::tar_render() automatically detects target dependencies in Rmd
+  # files and knits them, but there's no easy way to pass a custom rendering
+  # script like bookdown::html_document2(), so two things happen here:
+  #   1. Set a file-based target with tar_target_raw() and use tar_knitr_deps()
+  #      to detect the target dependencies in the Rmd file
+  #   2. Use a bunch of other file-based targets to actually render the document
+  #      through different custom functions
+  tar_target(bib_file,
+             here_rel("manuscript", "bibliography.bib"),
+             format = "file"),
+  
+  tar_target_raw("main_manuscript", here_rel("manuscript", "manuscript.Rmd"),
+                 format = "file",
+                 deps = c("bib_file",
+                          tar_knitr_deps(here_rel("manuscript", "manuscript.Rmd")))),
+  tar_target(html,
+             render_html(
+               input = main_manuscript,
+               output = here_rel("manuscript", "output", "manuscript.html"),
+               csl = csl,
+               bib_file,
+               support_folder = "output/html-support"),
+             format = "file"),
+  tar_target(pdf,
+             render_pdf(
+               input = main_manuscript,
+               output = here_rel("manuscript", "output", "manuscript.pdf"),
+               bibstyle = bibstyle,
+               bib_file),
+             format = "file"),
+  tar_target(ms_pdf,
+             render_pdf_ms(
+               input = main_manuscript,
+               output = here_rel("manuscript", "output", "manuscript-ms.pdf"),
+               bibstyle = bibstyle,
[…]

### 2021-09-09 _targets_packages.R

@@ -1,0 +1,25 @@
+# Generated by targets::tar_renv(): do not edit by hand
+library(base)
+library(bib2df)
+library(bs4Dash)
+library(clustermq)
+library(datasets)
+library(future)
+library(graphics)
+library(grDevices)
+library(gt)
+library(markdown)
+library(methods)
+library(pingr)
+library(rstudioapi)
+library(rvest)
+library(shiny)
+library(shinybusy)
+library(shinyWidgets)
+library(stats)
+library(stringi)
+library(tarchetypes)
+library(targets)
+library(utils)
+library(visNetwork)
+library(xml2)

### 2021-09-13 manuscript/manuscript.Rmd
Add some possible ideas for paper structure

@@
 Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
+
+Ideas:
+
+- Be like @Rohrer:2018 or @BlackwellGlynn:2018 and focus mostly on the DAG framework + a simulated example + an applied example from 1–2 previously published NVSQ or Voluntas papers - make it not super mathy, but still introduce $\operatorname{do}(\cdot)$ notation (like explain how $P(y \mid \operatorname{do}(x))$ is not the same as the observational $P(y \mid x)$, and that's why correlation ≠ causation)
+- …or, be like @BaBerrettCoupet:2021 and show a range of methods, like DAGs, RCTs, diff-in-diff, RD, and IV
+- …or do both and focus mostly on DAGs but also include a section that maps traditional econometrics methods to the DAG world?

### 2021-09-15 manuscript/manuscript.Rmd
Add possible outline

@@
   affiliation: Georgia State University
-  email: schaudhry@lclark.edu
+  email: mye2@student.gsu.edu
 date: "September 10, 2021"
@@
 - …or do both and focus mostly on DAGs but also include a section that maps traditional econometrics methods to the DAG world?
+
+More thoughts:
+
+- Take a similar approach to @BaBerrettCoupet:2021 and search top three nonprofit (also PA?) journals for mentions/uses of specific methods - make a graph of it
+- Provide a table of R and Stata commands at the end, like @BaBerrettCoupet:2021?
+- Mostly focus on DAGs and confounding, etc.
+- using DAGs to help with identification - identification is the main challenge of causal inference.
+You can do it with natural experiments, or you can do it with selection of observables driven by careful causal model and closing backdoors to isolate arrow.
+- Give a @Rohrer:2018 -esque introduction to DAGs + identification, then show an example (simulated data?
+real data?), then show how econometrics tools can be mapped onto DAGs for identification (and summarize some of the results from diff-in-diff, RDD, and IV papers over the past decade?)
+
+Tentative outline maybe:
+
+# Causal inference in nonprofit studies
+
+The "causal revolution" / "credibility revolution" has been happening for the past decade in social science disciplines - important because of reasons
+
+How has the discipline of nonprofit studies embraced the causal revolution?
+
+Are there fears of the word "causation" and an emphasis on pure association?
+
+Nonprofit data is very observational - quantitative data from past X years in N journals relies on experimental data Y% of the time, and observational data Z% of the time.
+
+In this paper, we provide a introductory primer / framework for approaching causal questions with observational data - accessible and approachable guide to nonprofit researchers and practitioners.
+The framework here is crucial for general academic research, policy analysis, and program evaluation, among other common types of nonprofit research.
+
+# Intro to DAGs
+
+## Paths and causal structures
+
+## Confounding specifically is the main issue with observational data
+
+Especially if it's unmeasured or unobservable…
+
+# Identification with DAGs
+
+## Adjustment and statistical control: blocking backdoor paths
+
+## How to adjust: stratification, including variable in regression model, matching, IPW
+
+## Stop controlling for everything
+
+Bad controls and colliders
+
+Table 2 fallacy + @KeeleStevensonElwert:2020 thing
+
+# Design-based causal inference with DAGs
+
+## Instrumental variables
+
+## Diff-in-diff
+
+## Regression discontinuity
+
+# Tools for researchers and practitioners
+
+Code in R and Stata? Point to other resources like The Mixtape, The Effect, Pearl's stuff, Morgan and Winship?
+
+# Conclusion

### 2021-09-15 README.Rmd
Add instructions for replicating

@@
 
+## How to replicate
+
+You can either [download the compendium as a ZIP file](/archive/main.zip) or use GitHub to clone or fork the compendium repository (see the green "Code" button at the top of the GitHub page).
+
+We use the [**renv** package](https://rstudio.github.io/renv/articles/renv.html) to create a stable version-specific library of packages, and we use the [**targets** package](https://docs.ropensci.org/targets/) to manage all file dependencies and run the analysis.
+To reproduce the findings and re-run the analysis, do the following:
+
+1. Download and install these fonts:
+    - [Cochineal](https://fontesk.com/cochineal-typeface/)
+    - [Econ Sans](https://github.com/hrbrmstr/hrbrthemes/tree/master/inst/fonts/econ-sans)
+    - [Linux Libertine O](https://www.cufonfonts.com/font/linux-libertine-o) (also [here](https://sourceforge.net/projects/linuxlibertine/))
+    - [Libertinus Math](https://github.com/alerque/libertinus)
+    - [InconsolataGo](https://github.com/ryanoasis/nerd-fonts/tree/master/patched-fonts/InconsolataGo)
+2. [Install R](https://cloud.r-project.org/) (and preferably [RStudio](https://www.rstudio.com/products/rstudio/download/#download)).
+    - If you're using macOS, [install XQuartz too](https://www.xquartz.org/), so that you have access to the Cairo graphics library
+3. Open `corrupt-caterpillar.Rproj` to open an [RStudio Project](https://r4ds.had.co.nz/workflow-projects.html).
+4. Make sure you have a working installation of LaTeX:
+    - *Easy-and-recommended way*: Install the [**tinytex** package](https://yihui.org/tinytex/) by running `install.packages("tinytex")` in the R console, then running `tinytex::install_tinytex()`
+    - *Easy-but-requires-huge-4+-GB-download way*: Download TeX Live ([macOS](http://www.tug.org/mactex/); [Windows](https://miktex.org/))
+5.
+If it's not installed already, R *should* try to install the **renv** package when you open the RStudio Project for the first time.
+If you don't see a message about package installation, install it yourself by running `install.packages("renv")` in the R console.
+6. Run `renv::restore()` in the R console to install all the required packages for this project.
+7. Run `targets::tar_make()` in the R console to automatically download all data files, process the data, run the analysis, and compile the paper and appendix.
+
+Running `targets::tar_make()` will create several helpful outputs:
+
+1. All project data in `data/`
+2. An analysis notebook website in `analysis/_site/index.html`
+3. PDF, HTML, and Word versions of the manuscript in `manuscript/output/`
+
 ## Licenses

### 2021-09-15 manuscript/manuscript.Rmd
Add experiments too
Add instructions for replicating

@@
 mainfont: Cochineal
-sansfont: Inter
+sansfont: Econ Sans Cnd
 author-note: >-
@@
 
+## Experiments
+
 ## Instrumental variables

### 2021-09-15 _targets.R
Temporarily stop knitting to PDF bc fonts

@@ -53,20 +53,20 @@ list(
                bib_file,
                support_folder = "output/html-support"),
              format = "file"),
-  tar_target(pdf,
-             render_pdf(
-               input = main_manuscript,
-               output = here_rel("manuscript", "output", "manuscript.pdf"),
-               bibstyle = bibstyle,
-               bib_file),
-             format = "file"),
-  tar_target(ms_pdf,
-             render_pdf_ms(
-               input = main_manuscript,
-               output = here_rel("manuscript", "output", "manuscript-ms.pdf"),
-               bibstyle = bibstyle,
-               bib_file),
-             format = "file"),
+  # tar_target(pdf,
+  #            render_pdf(
+  #              input = main_manuscript,
+  #              output = here_rel("manuscript", "output", "manuscript.pdf"),
+  #              bibstyle = bibstyle,
+  #              bib_file),
+  #            format = "file"),
+  # tar_target(ms_pdf,
+  #            render_pdf_ms(
+  #              input = main_manuscript,
+  #              output = here_rel("manuscript", "output", "manuscript-ms.pdf"),
+  #              bibstyle = bibstyle,
+  #              bib_file),
+  #            format = "file"),
   tar_target(docx,
              render_docx(
                input = main_manuscript,

### 2021-09-17 R/funs_knitting.R
New and improved colors and math annotations
Made this all work with externally sourced TeX and HTML files + mathjax + mathtools and unicode-math and Libertinus Math

@@ -36,7 +36,8 @@ render_html <- function(input, output, csl, support_folder, ...) {
       self_contained = FALSE,
       theme = NULL,
       extra_dependencies = extra_dependencies,
-      lib_dir = support_folder
+      lib_dir = support_folder,
+      includes = list(before_body = "pandoc/colors-annotations.html")
     ),
     encoding = "UTF-8"
   )
@@ -58,8 +59,9 @@ render_pdf <- function(input, output, bibstyle, ...) {
                       "-V", "chapterstyle=hikma-article"),
       md_extensions = "+raw_tex+smart-autolink_bare_uris",
       toc = FALSE,
-      keep_tex = FALSE,
-      citation_package = "biblatex"
+      keep_tex = TRUE,
+      citation_package = "biblatex",
+      includes = list(in_header = "pandoc/colors-annotations.tex")
     ),
     encoding = "UTF-8"
   )
@@ -81,7 +83,8 @@ render_pdf_ms <- function(input, output, bibstyle, ...) {
       md_extensions = "+raw_tex+smart-autolink_bare_uris",
       toc = FALSE,
       keep_tex = FALSE,
-      citation_package = "biblatex"
+      citation_package = "biblatex",
+      includes = list(in_header = "pandoc/colors-annotations.tex")
     ),
     encoding = "UTF-8"
   )

### 2021-09-17 _targets.R

@@ -53,20 +53,20 @@ list(
                bib_file,
                support_folder = "output/html-support"),
              format = "file"),
-  # tar_target(pdf,
-  #            render_pdf(
-  #              input = main_manuscript,
-  #              output = here_rel("manuscript", "output", "manuscript.pdf"),
-  #              bibstyle = bibstyle,
-  #              bib_file),
-  #            format = "file"),
-  # tar_target(ms_pdf,
-  #            render_pdf_ms(
-  #              input = main_manuscript,
-  #              output = here_rel("manuscript", "output", "manuscript-ms.pdf"),
-  #              bibstyle = bibstyle,
-  #              bib_file),
-  #            format = "file"),
+  tar_target(pdf,
+             render_pdf(
+               input = main_manuscript,
+               output = here_rel("manuscript", "output", "manuscript.pdf"),
+               bibstyle = bibstyle,
+               bib_file),
+             format = "file"),
+  tar_target(ms_pdf,
+             render_pdf_ms(
+               input = main_manuscript,
+               output = here_rel("manuscript", "output", "manuscript-ms.pdf"),
+               bibstyle = bibstyle,
+               bib_file),
+             format = "file"),
   tar_target(docx,
              render_docx(
                input = main_manuscript,

### 2021-09-17 R/graphics.R
Add custom graphics functions

@@ -1,0 +1,94 @@
+# ggplot themes -----------------------------------------------------------
+
+update_geom_defaults(ggdag:::GeomDagText,
+                     list(family = "Assistant", 
+                          fontface = "bold",
+                          color = "black"))
+
+#' theme_np
+#'
+#' A custom ggplot2 theme used throughout this project
+#'
+#' @param base_size base font size (default is 11)
+#' @param base_family base font family (default is Assistant)
+#'
+#' @export
+#'
+#' @examples
+#' library(ggplot2)
+#'
+#' ggplot(mpg, aes(x = displ, y = hwy, color = drv)) +
+#'   geom_point() +
+#'   labs(x = "Engine displacement", y = "Highway miles per gallon",
+#'        title = "A standard graph about cars",
+#'        subtitle = "Heavier cars get lower mileage",
+#'        caption = "Source: A dataset included in ggplot2",
+#'        color = "Drive") +
+#'   theme_np()
+theme_np <- function(base_size = 11, base_family = "Assistant") {
+  theme_bw(base_size, base_family) + 
+    theme(plot.title = element_text(size = rel(1.4), face = "bold",
+                                    family = "Assistant"),
+          plot.subtitle = element_text(size = rel(1), face = "plain",
+                                       family = "Assistant Light"),
+          plot.caption = element_text(size = rel(0.8), color = "grey50", face = "plain",
+                                      family = "Assistant Light",
+                                      margin = margin(t = 10), hjust = 0),
+          plot.tag = element_text(size = rel(1.4), face = "bold",
+                                  family = "Assistant"),
+          panel.border = element_rect(color = "grey50", fill = NA, size = 0.15),
+          panel.spacing = unit(1, "lines"),
+          panel.grid.minor = element_blank(),
+          strip.text = element_text(size = rel(0.9), hjust = 0,
+                                    family = "Assistant", face = "bold"),
+          strip.background = element_rect(fill = "#ffffff", colour = NA),
+          axis.ticks = element_blank(),
+          axis.title = element_text(family = "Assistant SemiBold", face = "plain", 
+                                    size = rel(0.8)),
+          axis.title.x = element_text(margin = margin(t = 5)),
+          axis.text = element_text(family = "Assistant Light", face = "plain"),
+          legend.key = element_blank(),
[…]

### 2021-09-17 _targets.R

@@ -27,6 +27,9 @@ here_rel <- function(...) {fs::path_rel(here::here(...))}
 # Pipeline ----------------------------------------------------------------
 
 list(
+  
+  # Define helper functions
+  tar_target(plot_funs, here_rel("R", "graphics.R"), format = "file"),
 
   ## Manuscript targets ------------------------------------------------------
 

### 2021-09-17 manuscript/manuscript.Rmd
Add plot for three types of associations

@@
 mainfont: Cochineal
-sansfont: Econ Sans Cnd
+sansfont: Assistant
 author-note: >-
@@
 library(tidyverse)
+library(patchwork)
+library(ggdag)
+library(dagitty)
 library(targets)
@@
   # Load stuff from targets
+  source(tar_read(plot_funs))
 })
@@
 Tentative outline maybe:
+
+```{r dag-associations, fig.width=9, fig.height=2.75, fig.asp=NULL, fig.cap="Three types of relationships in DAGs"}
+data_confounder <- dagify(
+  Y ~ Z + X,
+  X ~ Z,
+  coords = list(x = c(X = 1, Y = 3, Z = 2),
+                y = c(X = 1, Y = 1, Z = 2))
+) %>% 
+  tidy_dagitty() %>% 
+  mutate(arrow_color = ifelse(name == "Z", clrs_np$orange, "grey80")) %>% 
+  mutate(node_color = ifelse(name == "Z", TRUE, FALSE))
+
+confounder <- ggplot(data_confounder, aes(x = x, y = y, xend = xend, yend = yend)) +
+  geom_dag_edges(aes(edge_colour = arrow_color), edge_width = 1) +
+  geom_dag_point(aes(color = node_color), size = 12) +
+  geom_dag_text(data = filter(data_confounder, name != "Z"), size = 4) +
+  geom_dag_text(data = filter(data_confounder, name == "Z"), color = "white", size = 4) +
+  scale_color_manual(values = c("grey80", clrs_np$orange), guide = "none") +
+  coord_cartesian(ylim = c(0.95, 2.05)) +
+  labs(title = "Confounder", subtitle = "(Fork)") +
+  theme_np_dag() +
+  theme(plot.title = element_text(hjust = 0.5, color = clrs_np$orange),
+        plot.subtitle = element_text(hjust = 0.5, color = clrs_np$orange))
+
+data_mediator <- dagify(
+  Y ~ X + Z,
+  Z ~ X,
+  coords = list(x = c(X = 1, Y = 3, Z = 2),
+                y = c(X = 1, Y = 1, Z = 2))
+) %>% 
+  tidy_dagitty() %>% 
+  mutate(arrow_color = case_when(
+    name == "Z" & to == "Y" ~ clrs_np$purple, 
+    name == "X" & to == "Z" ~ clrs_np$purple,
+    TRUE ~ "grey80")) %>% 
+  mutate(node_color = ifelse(name == "Z", TRUE, FALSE))
+
+mediator <- ggplot(data_mediator, aes(x = x, y = y, xend = xend, yend = yend)) +
+  geom_dag_edges(aes(edge_colour = arrow_color), edge_width = 1) +
+  geom_dag_point(aes(color = node_color), size = 12) +
+  geom_dag_text(data = filter(data_mediator, name != "Z"), size = 4) +
+  geom_dag_text(data = filter(data_mediator, name == "Z"), color = "white", size = 4) +
+  scale_color_manual(values = c("grey80", clrs_np$purple), guide = "none") +
+  coord_cartesian(ylim = c(0.95, 2.05)) +
+  labs(title = "Mediator", subtitle = "(Chain)") +
+  theme_np_dag() +
+  theme(plot.title = element_text(hjust = 0.5, color = clrs_np$purple),
+        plot.subtitle = element_text(hjust = 0.5, color = clrs_np$purple))
+
+data_collider <- dagify(
+  Y ~ X,
+  Z ~ X + Y,
+  coords = list(x = c(X = 1, Y = 3, Z = 2),
+                y = c(X = 1, Y = 1, Z = 2))
+) %>% 
+  tidy_dagitty() %>% 
+  mutate(arrow_color = ifelse(to == "Z", clrs_np$red, "grey80")) %>% 
+  mutate(node_color = ifelse(name == "Z", TRUE, FALSE))
+
+collider <- ggplot(data_collider, aes(x = x, y = y, xend = xend, yend = yend)) +
+  geom_dag_edges(aes(edge_colour = arrow_color), edge_width = 1) +
+  geom_dag_point(aes(color = node_color), size = 12) +
+  geom_dag_text(data = filter(data_collider, name != "Z"), size = 4) +
+  geom_dag_text(data = filter(data_collider, name == "Z"), color = "white", size = 4) +
+  scale_color_manual(values = c("grey80", clrs_np$red), guide = "none") +
+  coord_cartesian(ylim = c(0.95, 2.05)) +
+  labs(title = "Collider", subtitle = "(Inverted fork)") +
+  theme_np_dag() +
+  theme(plot.title = element_text(hjust = 0.5, color = clrs_np$red),
+        plot.subtitle = element_text(hjust = 0.5, color = clrs_np$red))
+
+combined_associations <- (confounder | plot_spacer() | mediator | plot_spacer() | collider) +
+  plot_layout(widths = c(0.32, 0.02, 0.32, 0.02, 0.32))
+combined_associations
+```
 

### 2021-09-17 manuscript/manuscript.Rmd
Add econometrics DAGs

@@
 
+The language of causal graphs, identification, and adjustment provide a universal grammar for discussing causal effects.
+Commonly used approaches in econometrics and other social science disciplines can be written as causal graphs (see Figure \@ref(fig:design-based-dags))
+
+```{r design-based-dags, fig.width=9, fig.height=6, fig.asp=NULL, fig.cap="Possible DAGs for common design-based experimental and quasi-experimental approaches to causal inference.
+Red arrows represent the identified and isolated relationship between treatment $x$ and outcome $y$.
+Square nodes represent statistical adjustment."}
+data_rct <- dagify(
+  Y ~ Z + X,
+  coords = list(x = c(X = 1, Y = 3, Z = 2),
+                y = c(X = 1, Y = 1, Z = 2)),
+  labels = c(X = "X = x"),
+  exposure = "X",
+  outcome = "Y"
+) %>% 
+  tidy_dagitty() %>% 
+  node_status() %>% 
+  mutate(arrow_color = ifelse(name == "X", clrs_np$red, "grey80"))
+
+rct <- ggplot(data_rct, aes(x = x, y = y, xend = xend, yend = yend)) +
+  geom_dag_edges(aes(edge_colour = arrow_color), edge_width = 1) +
+  geom_dag_point(aes(color = status), size = 12) +
+  geom_dag_text(size = 4) +
+  geom_dag_text(aes(label = label),
+                      nudge_x = 0, nudge_y = 0.2) +
+  scale_color_manual(values = c(clrs_np$green, clrs_np$blue), 
+                     na.value = clrs_np$yellow, guide = "none") +
+  coord_cartesian(ylim = c(0.95, 2.05)) +
+  labs(title = "Randomized trial", 
+       subtitle = "Randomization deletes all arrows into X") +
+  theme_np_dag() +
+  theme(plot.title = element_text(hjust = 0.5),
+        plot.subtitle = element_text(hjust = 0.5))
+
+data_iv <- dagify(Y ~ X + U,
+                  X ~ I + U,
+                  exposure = "X",
+                  outcome = "Y",
+                  latent = "U",
+                  coords = list(x = c(X = 1, Y = 3, U = 2, I = 0),
+                                y = c(X = 1, Y = 1, U = 2, I = 1))) %>% 
+  tidy_dagitty() %>% 
+  adjust_for(var = "I") %>% 
+  node_status() %>% 
+  mutate(arrow_color = ifelse(name == "X", clrs_np$red, "grey80"))
+
+iv <- ggplot(data_iv, aes(x = x, y = y, xend = xend, yend = yend)) +
+  geom_dag_edges(aes(edge_colour = arrow_color), edge_width = 1) +
+  geom_dag_point(aes(color = status, shape = adjusted), size = 12) +
+  geom_dag_text(size = 4) +
+  scale_color_manual(values = c(clrs_np$green, "grey80", clrs_np$blue), 
+                     na.value = clrs_np$yellow, guide = "none") +
+  scale_shape_manual(values = c(15, 19), guide = "none") +
+  coord_cartesian(ylim = c(0.95, 2.05)) +
+  labs(title = "Instrumental variable", 
+       subtitle = "Find effect of instrument (I) on X, then find effect of (X | I) on Y") +
+  theme_np_dag() +
+  theme(plot.title = element_text(hjust = 0.5),
+        plot.subtitle = element_text(hjust = 0.5))
+
+data_diff_diff <- dagify(Y ~ X + Time + Location,
+                         X ~ Time + Location,
+                         exposure = "X",
+                         outcome = "Y",
+                         coords = list(x = c(X = 1, Y = 3, Time = 2, Location = 2),
+                                       y = c(X = 1, Y = 1, Time = 2, Location = 0))) %>% 
+  tidy_dagitty() %>% 
+  adjust_for(var = c("Time", "Location")) %>% 
+  node_status() %>% 
+  mutate(arrow_color = ifelse(name == "X", clrs_np$red, "grey80"))
+
+diff_diff <- ggplot(data_diff_diff, aes(x = x, y = y, xend = xend, yend = yend)) +
+  geom_dag_edges(aes(edge_colour = arrow_color), edge_width = 1) +
+  geom_dag_point(aes(color = status, shape = adjusted), size = 12) +
+  geom_dag_text(data = filter(data_diff_diff, name %in% c("X", "Y")), size = 4) +
+  geom_dag_text(data = filter(data_diff_diff, name == "Location"), 
+                nudge_y = 0.37, size = 4) +
+  geom_dag_text(data = filter(data_diff_diff, name == "Time"), 
+                nudge_y = -0.37, size = 4) +
+  scale_color_manual(values = c(clrs_np$green, clrs_np$blue), 
+                     na.value = clrs_np$yellow, guide = "none") +
+  scale_shape_manual(values = c(15, 19), guide = "none") +
+  coord_cartesian(ylim = c(-0.05, 2.05)) +
+  labs(title = "Difference-in-differences", 
+       subtitle = "Adjust for both time (e.g., year) and location (e.g., country, state)") +
+  theme_np_dag() +
+  theme(plot.title = element_text(hjust = 0.5),
+        plot.subtitle = element_text(hjust = 0.5))
+
+data_rdd <- dagify(Y ~ X + Running + U,
+                   X ~ Threshold + U,
+                   Running ~ U,
+                   Threshold ~ Running,
+                   exposure = "X",
+                   outcome = "Y",
+                   latent = "U",
+                   coords = list(x = c(X = 1, Y = 4, Running = 3, Threshold = 2, U = 3),
+                                 y = c(X = 1, Y = 1, Running = 2, Threshold = 1.75, U = 3))) %>% 
+  tidy_dagitty() %>% 
+  adjust_for(var = c("Threshold", "Running")) %>% 
+  node_status() %>% 
+  mutate(arrow_color = ifelse(name == "X", clrs_np$red, "grey80"),
+         name = recode(name, "Running" = "Running\nvariable"))
+
+rdd <- ggplot(data_rdd, aes(x = x, y = y, xend = xend, yend = yend)) +
+  geom_dag_edges(aes(edge_colour = arrow_color), edge_width = 1) +
+  geom_dag_point(aes(color = status, shape = adjusted), size = 12) +
+  geom_dag_text(data = filter(data_rdd, name %in% c("X", "Y", "U")), size = 4) +
+  geom_dag_text(data = filter(data_rdd, name == "Running\nvariable"),
+                nudge_y = -0.47, size = 4, lineheight = 1) +
+  geom_dag_text(data = filter(data_rdd, name == "Threshold"),
+                nudge_y = -0.35, size = 4) +
+  scale_color_manual(values = c(clrs_np$green, "grey80", clrs_np$blue), 
+                     na.value = clrs_np$yellow, guide = "none") +
+  scale_shape_manual(values = c(15, 19), guide = "none") +
+  coord_cartesian(ylim = c(0.95, 3.05)) +
+  labs(title = "Regression discontinuity", 
+       subtitle = "Adjust for both the running variable and the threshold") +
+  theme_np_dag() +
+  theme(plot.title = element_text(hjust = 0.5),
+        plot.subtitle = element_text(hjust = 0.5))
+
+plot_econometrics <- ((rct | iv) / plot_spacer() / (diff_diff | rdd)) +
+  plot_layout(heights = c(0.49, 0.02, 0.49))
+plot_econometrics
+```
+
 ## Experiments

### 2021-09-18 R/graphics.R
Coordinate math and plot colors

@@ -92,3 +92,37 @@ theme_np_dag <- function(...) {
 clrs_np <- viridisLite::turbo(6, begin = 0, end = 1) %>% 
   purrr::set_names(c("purple", "blue", "green", "yellow", "orange", "red")) %>% 
   as.list()
+
+#' Convert colors to LaTeX RGB definitions
+#'
+#' Take a named list of hex colors and convert them to RGB values.
+#'
+#' The resulting output has the class `clrs` and will print only the LaTeX color
+#' definitions by default using `print.clrs()`
+#'
+#' @param clrs named list of hex colors
+#'
+#' @return A tibble with columns for the color name, hex code, red value, green
+#'   value, blue value, comma-separated RGB value, and a LaTeX color definition
+#'
+#' @export
+#' @md
+#' 
+#' @example 
+#' get_latex_rgb(clrs)
+get_latex_rgb <- function(clrs = clrs_np) {
+  clrs <- unlist(clrs)
+  
+  clrs_table <- tibble::enframe(x, name = "color_name", value = "hex") %>% 
+    bind_cols(as_tibble(t(col2rgb(x)))) %>%
+    mutate(rgb = paste(red, green, blue, sep = ", ")) %>%
+    mutate(latex = glue::glue("\\definecolor{<<color_name>>}{RGB}{<<rgb>>}",
+                              .open = "<<", .close = ">>"))
+  
+  class(clrs_table) <- append("clrs", class(clrs_table))
+  clrs_table
+}
+
+print.clrs <- function(x) {
+  cat(x$latex, sep = "\n")
+}

### 2021-09-18 manuscript/manuscript.Rmd
Use dotted line for U -> X arrow bc it gets deleted

@@
   mutate(arrow_color = ifelse(name == "X", clrs_np$red, "grey80"),
-         name = recode(name, "Running" = "Running\nvariable"))
+         name = recode(name, "Running" = "Running\nvariable"),
+         linetype = ifelse(name == "U" & to == "X", "21", "solid"))
 
 rdd <- ggplot(data_rdd, aes(x = x, y = y, xend = xend, yend = yend)) +
-  geom_dag_edges(aes(edge_colour = arrow_color), edge_width = 1) +
+  geom_dag_edges(aes(edge_colour = arrow_color, edge_linetype = linetype), edge_width = 1) +
   geom_dag_point(aes(color = status, shape = adjusted), size = 12) +

### 2021-09-18 R/graphics.R
Use S3 magic to generate and plot the color palette

@@ -86,12 +86,48 @@ theme_np_dag <- function(...) {
 
 # Colors ------------------------------------------------------------------
 
-# clrs_turbo <- viridisLite::turbo(6, begin = 0, end = 1)
-# clrs_turbo %>% scales::show_col()
+#' Nonprofit palette
+#' 
+#' Generate a palette of 6 colors from the viridis turbo palette
+#'
+#' @return A named list of colors
+#' @export
+#' @md
+#'
+#' @examples
+#' clrs_np <- np_palette()
+#' 
+#' # Access colors by name with $
+#' clrs_np$blue
+#' clrs_np$orange
+np_palette <- function() {
+  clrs <- viridisLite::turbo(6, begin = 0, end = 1) %>% 
+    purrr::set_names(c("purple", "blue", "green", "yellow", "orange", "red")) %>% 
+    as.list()
+  
+  class(clrs) <- append("palette", class(clrs))
+  return(clrs)
+}
 
-clrs_np <- viridisLite::turbo(6, begin = 0, end = 1) %>% 
-  purrr::set_names(c("purple", "blue", "green", "yellow", "orange", "red")) %>% 
-  as.list()
+
+#' Show a palette
+#'
+#' This is a wrapper for `scales::show_col()` that automatically plots a list of
+#' colors generated with `np_palette()`
+#'
+#' @param clrs color palette to show
+#' @param ... arguments passed to `scales::show_col()`
+#'
+#' @return
+#' @export
+#' @md
+#'
+#' @examples
+#' clrs_np <- np_palette()
+#' plot(clrs_np)
+plot.palette <- function(clrs, ...) {
+  scales::show_col(unlist(clrs), ...)
+}
 
 #' Convert colors to LaTeX RGB definitions
 #'
@@ -109,8 +145,8 @@ clrs_np <- viridisLite::turbo(6, begin = 0, end = 1) %>%
 #' @md
 #' 
 #' @example 
-#' get_latex_rgb(clrs)
-get_latex_rgb <- function(clrs = clrs_np) {
+#' get_latex_rgb()
+get_latex_rgb <- function(clrs = np_palette()) {
   clrs <- unlist(clrs)
   
   clrs_table <- tibble::enframe(x, name = "color_name", value = "hex") %>% 
@@ -119,10 +155,10 @@ get_latex_rgb <- function(clrs = clrs_np) {
     mutate(latex = glue::glue("\\definecolor{<<color_name>>}{RGB}{<<rgb>>}",
[…]

### 2021-09-18 manuscript/manuscript.Rmd
Load palette
Use an annotate() instead of a mostly empty geom_text()

@@
 })
+
+clrs_np <- np_palette()
 ```
@@
                 y = c(X = 1, Y = 1, Z = 2)),
-  labels = c(X = "X = x"),
   exposure = "X",
@@
   geom_dag_text(size = 4) +
-  geom_dag_text(aes(label = label),
-                      nudge_x = 0, nudge_y = 0.2) +
+  annotate(geom = "text", x = 1, y = 1.2, label = "X = x") +
   scale_color_manual(values = c(clrs_np$green, clrs_np$blue), 

### 2021-09-18 R/graphics.R

@@ -1,4 +1,9 @@
 # ggplot themes -----------------------------------------------------------
+# Set default fonts for text-based geoms
+update_geom_defaults("text",
+                     list(family = "Assistant", 
+                          fontface = "bold",
+                          color = "black"))
 
 update_geom_defaults(ggdag:::GeomDagText,
                      list(family = "Assistant", 

### 2021-09-18 R/funs_knitting.R
Hide messages

@@ -1,4 +1,4 @@
-library(dplyr)
+suppressPackageStartupMessages(library(dplyr))
 library(purrr)
 library(stringr)
 library(bib2df)

### 2021-09-18 manuscript/manuscript.Rmd
Add a longer outline + citation placeholders

@@
   email: mye2@student.gsu.edu
-date: "September 10, 2021"
+date: "September 18, 2021"
 published: Drafting
@@
 
-Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua [@BaBerrettCoupet:2021].
-Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
-Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur.
-Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
+Introduction stuff…
 
-Ideas:
+The "causal revolution" / "credibility revolution" has been happening for the past decade in social science disciplines - important because of reasons
 
-- Be like @Rohrer:2018 or @BlackwellGlynn:2018 and focus mostly on the DAG framework + a simulated example + an applied example from 1–2 previously published NVSQ or Voluntas papers - make it not super mathy, but still introduce $\operatorname{do}(\cdot)$ notation (like explain how $P(y \mid \operatorname{do}(x))$ is not the same as the observational $P(y \mid x)$, and that's why correlation ≠ causation)
-- …or, be like @BaBerrettCoupet:2021 and show a range of methods, like DAGs, RCTs, diff-in-diff, RD, and IV
-- …or do both and focus mostly on DAGs but also include a section that maps traditional econometrics methods to the DAG world?
+How has the discipline of nonprofit studies embraced the causal revolution?
 
-More thoughts:
+Are there fears of the word "causation" and an emphasis on pure association?
 
-- Take a similar approach to @BaBerrettCoupet:2021 and search top three nonprofit (also PA?) journals for mentions/uses of specific methods - make a graph of it
-- Provide a table of R and Stata commands at the end, like @BaBerrettCoupet:2021?
-- Mostly focus on DAGs and confounding, etc.
-- using DAGs to help with identification - identification is the main challenge of causal inference.
-You can do it with natural experiments, or you can do it with selection of observables driven by careful causal model and closing backdoors to isolate arrow.
-- Give a @Rohrer:2018 -esque introduction to DAGs + identification, then show an example (simulated data?
-real data?), then show how econometrics tools can be mapped onto DAGs for identification (and summarize some of the results from diff-in-diff, RDD, and IV papers over the past decade?)
+Nonprofit data is very observational - quantitative data from past X years in N journals relies on experimental data Y% of the time, and observational data Z% of the time.
 
-Tentative outline maybe:
+Similar to recent work by @BaBerrettCoupet:2021 and @Ma-etal:2021 and @Rohrer:2018, in this paper, we provide a introductory primer / framework for approaching causal questions with observational data - accessible and approachable guide to nonprofit researchers and practitioners.
+The framework here is crucial for general academic research, policy analysis, and program evaluation, among other common types of nonprofit research.
+
+# Why care about causation in nonprofit studies? (MENG)
+
+Graph and analysis of causal stuff in top 3 here
+
+[@Samii:2016]
+
+> 941 - "Causal empiricism is associated with “identification strategy” research designs."
+
+> This is different from quantitative “pseudo-general pseudo-facts” that come from multiple regression
+
+>> In quantitative research, pseudo-facts are statistical results that are interpreted erroneously in terms of their causal implications, and pseudo-general findings are ones that are erroneously described as applying to a more general class of units than is immediately warranted.
+
+Don't make unwarranted causal-ish statements, but also don't be afraid of the "c-word" [@Hernan:2018]
+
+Language can imply causality even if authors explicitly eschew causal language and identification strategies [@HaberWietenRohrer:2021]
+
+# Intro to DAGs (ANDREW)
+
+[@PearlMackenzie:2018; @MorganWinship:2014; @PearlGlymourJewell:2016]
+
+DAGs encode our understanding of the data generating process - what in the natural and social world leads to the treatment/exposure, the outcome, and both simultaneously.
+It represents a type of philosophical model of how the observed world works.
+Importantly, causal graphs also encode our assumptions of how the world works.
+
+In addition to encoding our philosophy and theory of the data generating process, DAGs also serve as an important statistical tool for isolating or identifying causal quantities of interest.
+Identification strategy definition:
+
+> The central role of an identification strategy is to provide a logic for establishing that D is independent of potential values of Y, thus allowing the analyst to interpret observed associations as causal effects.
+[@KeeleStevensonElwert:2020]
+
+## A vocabulary for causal inference
+
+A grammar
+
+Paths and causal structures
+
+- Nodes - nodes can be unmeasurable, or even unobserved
+- Edges - Arrows indicate a relationship, or the passing of *statistical association* between nodes - intervening on one node leads to changes in another node - When two nodes are connected by an arrow, we are stating that there is an assumed causal relationship between those nodes; when there is no arrow between nodes, we are explicitly stating that there is no relationship between the two.
+- d-connection and d-separation - statistical association cannot pass between nodes, either because the arrows are drawn in a way that makes it so information does not link the two, or because conditioning/adjustment blocks the pathway
+- Acyclicity - can't get back to a node - represents flow of time - if things are cyclical, like hiring new staff $\leftrightarrow$ increased capacity, you can make these nodes time-based: hiring new staff_t → increased capacity_t → hiring new staff_t+1 → increased capacity_t+1
 
@@
 
-# Causal inference in nonprofit studies
+This stuff in Figure \@ref(fig:dag-associations): (@Elwert:2013 for forks, chains, inverted forks terminology)
 
-The "causal revolution" / "credibility revolution" has been happening for the past decade in social science disciplines - important because of reasons
+- Confounding - forks
+- Mediation - chains
+- Collision - inverted forks [@ElwertWinship:2014; @KnoxLoweMummolo:2020]
 
-How has the discipline of nonprofit studies embraced the causal revolution?
+Confounding is a major scary issue though - especially if it's unmeasured or unobservable…
 
-Are there fears of the word "causation" and an emphasis on pure association?
+# Identification with DAGs (ANDREW)
 
-Nonprofit data is very observational - quantitative data from past X years in N journals relies on experimental data Y% of the time, and observational data Z% of the time.
+## Experimental causal effects vs. observational causal effects
 
-In this paper, we provide a introductory primer / framework for approaching causal questions with observational data - accessible and approachable guide to nonprofit researchers and practitioners.
-The framework here is crucial for general academic research, policy analysis, and program evaluation, among other common types of nonprofit research.
+Potential outcomes notation
 
-# Intro to DAGs
+$\operatorname{do}(x)$ notation
 
-## Paths and causal structures
+In experiments, the researcher has total control over assignment to treatment, which means all edges/arrows that might influence treatment can be removed.
+There is no confounding to worry about and we can measure the exact causal effect of X on Y.
+In potential outcomes language, you still can't see each individual's yes and no response, but you can average all the yeses and noes and get an average causal effect
 
-## Confounding specifically is the main issue with observational data
+With observational data, we'd like to measure $\mathbf{E}(y \mid \operatorname{do}(x))$ but we can only actually see and measure $\mathbf{E}(y \mid x)$, and as shown in Equation \@ref(eq:correlation-not-causation), these two expressions are not the same.
+This is a formal version of the common phrase "correlation isn't causation":
 
-Especially if it's unmeasured or unobservable…
+\begin{equation}
+{\color{gray} \overbrace{{\color{orange} \underbracket[0.25pt]{{\color{black} \mathbf{E}(y \mid \operatorname{do}(x)) \vphantom{\frac{1}{2}}}}_{\color{orange} \text{Causation}}}}^{\color{gray} \mathclap{\substack{\text{The average} \\ \text{population-level} \\ \text{change in $y$ when} \\ \textit{directly intervening} \\ \text{(or doing) $x$}}}}}
+\quad \neq \quad
+{\color{gray} \overbrace{\color{purple} \underbracket[0.25pt]{{\color{black} \mathbf{E}(y \mid x)} \vphantom{\frac{1}{2}}}_{\color{purple} \text{Correlation}}}^{\color{gray} \mathclap{\substack{\text{The average} \\ \text{population-level} \\ \text{change in $y$ when} \\ \text{accounting for} \\ \textit{observed } x}}}}
+(\#eq:correlation-not-causation)
+\end{equation}
 
-# Identification with DAGs
+What we want to be able to do is transform the $\mathbf{E}(y \mid \operatorname{do}(x))$ expression into something without the $\operatorname{do}(x)$, or something *do*-free.
+A set of three systematic rules for analyzing and decomposing causal graphs known as *do*-calculus provide certain conditions under which we can treat an interventional $\operatorname{do}(\cdot)$ expression like an observed value instead.
+A complete exploration of these three rules of *do*-calculus go beyond the scope of this paper, but lots of resources like Pearl, that one textbook, other things in my blog post, etc.
+[@Pearl:2012; @Pearl:2019]
 
-## Adjustment and statistical control: blocking backdoor paths
+The most common derivation of the rules of *do*-calculus is an approach called "backdoor adjustment".
+By adjusting or controlling for nodes that open up backdoor paths between the treatment and outcome nodes, we can mathematically transform a $\operatorname{do}(\cdot)$ expression into something based solely on observational data.
+Formally, the backdoor adjustment formula is defined in Equation \@ref(eq:backdoor):
 
-## How to adjust: stratification, including variable in regression model, matching, IPW
+\begin{equation}
+{\color{gray} \overbrace{\color{black} \mathbf{E}(y \mid \operatorname{do}(x)) \vphantom{\frac{1}{2}}}^{\substack{\text{Causal effect} \\ \text{of $x$ on $y$}}}}
+=
+{\color{gray} \underbrace{{\color{black} \sum_z}}_{\mathclap{\substack{\text{Sum across} \\ \text{all values of $z$}}}}}
+{\color{gray} \overbrace{\color{black} \mathbf{E} (y \mid x, z) \vphantom{\frac{1}{2}}}^{\mathclap{\substack{\text{Conditional} \\ \text{mean of $y$,} \\ \text{given $x$ and $z$$\dots$}}}}}
+\times 
+{\color{gray} \overbrace{\color{black} \mathbf{P}(z) \vphantom{\frac{1}{2}}}^{\mathclap{\substack{\text{$\dots$ weighted} \\ \text{by $z$}}}}}
+(\#eq:backdoor)
+\end{equation}
+
+Put more simply, \@ref(eq:backdoor) demonstrates that we can remove the interventional $\operatorname{do}(x)$ from the left-hand side of the equation by controlling for (or conditioning on) all the confounders $z$ that open up a backdoor pathway between treatment and outcome.
+As a simplified illustration, suppose that the relationship between treatment and outcome is confounded only by a nonprofit's size, which is measured as either large or small.
+Applying this backdoor adjustment formula would entail finding average value of the outcome conditioned on the treatment among large nonprofits, multiplied by the proportion of large nonprofits, added to the average value of the outcome conditioned on the treatment among small nonprofits, multiplied by the proportion of small nonnprofits.
+The resulting sum would then be the unconfounded causal effect.
+
+In practice, statistical adjustment rarely involves a single binary confounder.
+For instance, in the causal graph in Fig X, X, Y, and Z all open up backdoors between treatment and outcome, and all three would need to be adjusted for.
+We will provide a practical demonstration of more common adjustment strategies when there are multiple confounders in section X.
+At this point, what is important to note is that adjusting for confounding nodes allows us to isolate the single pathway between treatment and outcome.
+Because spurious statistical associations from other nodes have been blocked statistically, the relationship we care about is identified and we can talk about the *causal* effect of the treatment on the outcome.
+
+TODO: Plot of backdoor and frontdoor adjustment DAGs, but using nonprofit situations
+
+A less common derivation of the rules of *do*-calculus is frontdoor adjustment, commonly used when confounding is unobserved and undertheorized and unmeasurable.
+Smoking genetics tar cancer thing - Bellemare paper example [@BellemareBloemWexler:2020] - frontdoor adjustment formula here?
+We do not provide a complete example here—see Bellemare for that—but again, what is most important here is that we can again mathematically transform a quantity with an interventional do(x) into a do-free quantity, meaning that we can make causal claims from observational data.
+
+The backdoor and frontdoor criteria are the most common applications of *do*-calculus because they are readily apparent in causal graphs—it is possible to see forks joining exposure and outcome and identify backdoors, or see measurable mediating nodes that could be used as front doors.
+In more complex DAGs, these backdoor and frontdoor shortcuts might not be readily visible.
+In that case, there are software packages that algorithmically work through the various rules of *do*-calculus to determine the set of nodes that need to be adjusted in order to isolate the x → y relationship.
+Not every DAG is identifiable; but any identifiable DAG can be identified.
 
 ## Stop controlling for everything
+
+In addition to identification, DAGs provide additional statistical insight and guidance regarding control variables or covariates in regression models
 
@@
 
-# Design-based causal inference with DAGs
+Post-treatment control bias - cite that one paper by Brendan Nyhan - DAGs make it obvious which nodes are post-treatment, since they appear after the treatment node in the causal chain in the graph
 
+# How to adjust (MENG AND ANDREW)
+
+TODO: Use simulated data based on a DAG that has a mediator, collider, and a bunch of confounders
+
+TODO: Replicate a study + draw a DAG for it + use IPW to close back doors
+
+## Stratification
+
+## Regression
+
+## Matching
+
+Matching creates entirely new treatment/control populations
+
+## IPW
+
+IPW creates comparable pseudo populations
+
+MAYBE: Make an image with little shaded people showing how matching and IPW work, similar to Torres:2020 (maybe with https://github.com/propublica/weepeople ?)
+
+## Other methods?
+
+Briefly mention other methods, like marginal structural models [@BlackwellGlynn:2018]?
+
+# Design-based causal inference with DAGs (ANDREW AND MENG)
+
+[@AngristPischke:2009; @AngristPischke:2015]
 The language of causal graphs, identification, and adjustment provide a universal grammar for discussing causal effects.
@@
 Red arrows represent the identified and isolated relationship between treatment $x$ and outcome $y$.
-Square nodes represent statistical adjustment."}
+Square nodes represent statistical adjustment.", warning=FALSE}
 data_rct <- dagify(
@@
 
+For each of these:
+
+1. Sort explanation of what they require, how they work
+2. Explanation of how that connects to the DAG and how the graph reveals the identification strategy
+3. Point to some nonprofit-related papers that use these
+
 ## Experiments
+
+No need to control for a ton of things in an RCT precisely because the arrows into X get deleted.
+No need to worry about perfect balance checks because the researcher has control over and understands the data generating process and assignment to treatment.
+Look at that one knitted Rmd on RCT FAQs: https://macartan.github.io/i/notes/rct_faqs.html
 
@@
 
+IVs have to meet the exclusion restriction - the instrument can only influence the outcome through the treatment.
+DAGs make this assumption very clear.
+There cannot be an arrow connecting the instrument to the outcome.
+DAGs also inform the exogeneity assumption—no other nodes in the graph can feed into the instrument node
+
+It's like frontdoor adjustment (https://www.stat.cmu.edu/~cshalizi/402/lectures/23-causal-estimation/lecture-23.pdf) - Really we're finding the causal effect of I on X, then X on Y, generally through 2SLS
+
 ## Diff-in-diff
+
+Time / location, TWFE stuff
 
@@
 
-# Tools for researchers and practitioners
+Threshold/cutpoint
+
+Adjusting for the threshold and only looking at data right around it makes it so that we can treat the sample as if it were random (by assumption), which then means we can delete any arrows going into X just like an RCT
+
+Cite Nick's *The Effect* - refer to his website with the DAGs and animations
+
+# Tools for researchers and practitioners (BOTH)
 
@@
 
-# Conclusion
+[@Huntington-Klein:2021; @Cunningham:2021]
+
+# Conclusion (BOTH)

### 2021-09-18 README.Rmd
Make this more Windows friendly

@@
     - [Cochineal](https://fontesk.com/cochineal-typeface/)
-    - [Econ Sans](https://github.com/hrbrmstr/hrbrthemes/tree/master/inst/fonts/econ-sans)
+    - [Assistant](https://fonts.google.com/specimen/Assistant)
     - [Linux Libertine O](https://www.cufonfonts.com/font/linux-libertine-o) (also [here](https://sourceforge.net/projects/linuxlibertine/))
@@
     - [InconsolataGo](https://github.com/ryanoasis/nerd-fonts/tree/master/patched-fonts/InconsolataGo)
+      - If you're using Windows, make sure you right click on the font files and choose "Install for all users" when installing these fonts
 2. [Install R](https://cloud.r-project.org/) (and preferably [RStudio](https://www.rstudio.com/products/rstudio/download/#download)).
-    - If you're using macOS, [install XQuartz too](https://www.xquartz.org/), so that you have access to the Cairo graphics library
+    - If you're using macOS, [install XQuartz too](https://www.xquartz.org/) so that you have access to the Cairo graphics library
+    - If you're using Windows, [install RTools too](https://cran.r-project.org/bin/windows/Rtools/) and add it to your PATH so that you can install packages from source if needed
 3. Open `corrupt-caterpillar.Rproj` to open an [RStudio Project](https://r4ds.had.co.nz/workflow-projects.html).
 4. Make sure you have a working installation of LaTeX:
-    - *Easy-and-recommended way*: Install the [**tinytex** package](https://yihui.org/tinytex/) by running `install.packages("tinytex")` in the R console, then running `tinytex::install_tinytex()`
+    - *Easy-and-recommended way*: Install the [**tinytex** package](https://yihui.org/tinytex/) by running `install.packages("tinytex")` in the R console, then running `tinytex::install_tinytex()`.
+The first time you compile a PDF might require an extra minute or so while **tinytex** installs some missing LaTeX packages like csquotes, mathools, and others
     - *Easy-but-requires-huge-4+-GB-download way*: Download TeX Live ([macOS](http://www.tug.org/mactex/); [Windows](https://miktex.org/))

### 2021-09-18 manuscript/manuscript.Rmd
Make this more Windows friendly

@@
 sansfont: Assistant
+monofont: '`r if (Sys.info()["sysname"] == "Windows") "InconsolataGo Nerd Font" else "InconsolataGo"`'
+ms_mainfont: '`r if (Sys.info()["sysname"] == "Windows") "Linux Libertine" else "Linux Libertine O"`'
+ms_sansfont: Assistant
+ms_monofont: '`r if (Sys.info()["sysname"] == "Windows") "InconsolataGo Nerd Font" else "InconsolataGo"`'
 author-note: >-

### 2021-09-18 R/graphics.R

@@ -1,3 +1,12 @@
+# Windows fonts
+if (Sys.info()["sysname"] == "Windows") {
+  # Register these fonts if running R on Windows
+  windowsFonts(`Assistant` = windowsFont("Assistant"))
+  windowsFonts(`Assistant Light` = windowsFont("Assistant Light"))
+  windowsFonts(`Assistant SemiBold` = windowsFont("Assistant SemiBold"))
+}
+
+
 # ggplot themes -----------------------------------------------------------
 # Set default fonts for text-based geoms
 update_geom_defaults("text",

### 2021-09-19 R/funs_knitting.R
Prefix the word count with a * bc Windows trims it off?

@@ -217,5 +217,5 @@ count_words <- function(html) {
 }
 
 print.wordcount <- function(x) {
-  cat(scales::comma(x), "words in manuscript\n")
+  cat("*", scales::comma(x), "words in manuscript\n")
 }

### 2021-09-19 README.Rmd
Minor edits

@@
 
-1. Download and install these fonts:
+1. Download and install these fonts (if you're using Windows, make sure you right click on the font files and choose "Install for all users" when installing these fonts):
     - [Cochineal](https://fontesk.com/cochineal-typeface/)
@@
     - [InconsolataGo](https://github.com/ryanoasis/nerd-fonts/tree/master/patched-fonts/InconsolataGo)
-      - If you're using Windows, make sure you right click on the font files and choose "Install for all users" when installing these fonts
 2. [Install R](https://cloud.r-project.org/) (and preferably [RStudio](https://www.rstudio.com/products/rstudio/download/#download)).
@@
     - *Easy-and-recommended way*: Install the [**tinytex** package](https://yihui.org/tinytex/) by running `install.packages("tinytex")` in the R console, then running `tinytex::install_tinytex()`.
-The first time you compile a PDF might require an extra minute or so while **tinytex** installs some missing LaTeX packages like csquotes, mathools, and others
+The first time you compile a PDF might require an extra minute or so while **tinytex** installs some missing LaTeX packages like csquotes, mathools, and others, but after that, it'll go fast from then on.
     - *Easy-but-requires-huge-4+-GB-download way*: Download TeX Live ([macOS](http://www.tug.org/mactex/); [Windows](https://miktex.org/))

### 2021-09-22 manuscript/manuscript.Rmd
Add note about RCTs and controls

@@
 No need to worry about perfect balance checks because the researcher has control over and understands the data generating process and assignment to treatment.
-Look at that one knitted Rmd on RCT FAQs: https://macartan.github.io/i/notes/rct_faqs.html
+Look at that one knitted Rmd on RCT FAQs: https://macartan.github.io/i/notes/rct_faqs.html - CONSORT also says to stop doing balance tests - only really need to control for things that might be predictive (https://twitter.com/statsepi/status/1115902270888128514?s=21), but theoretically anything that influences the allocation to treatment is taken care of by randomization.
 

### 2022-04-13 manuscript/manuscript.Rmd
Create better outline; add more placeholder stuff
Starting to incorporate WCNPD comments

@@
 ---
-title: "Full title goes here"
-short-title: "Shorter title goes here"
+title: "Clarifying Correlation and Causation: A Guide to Modern Quantitative Causal Inference in Nonprofit Studies"
+short-title: "Clarifying Correlation and Causation"
 author:
@@
   email: mye2@student.gsu.edu
-date: "September 18, 2021"
+  url: https://aysps.gsu.edu/phd-student/ye-meng/
+date: "April 13, 2022"
 published: Drafting
@@
 abstract: >-
-  Abstract goes here.
+  Discovering causal relationships and testing theoretical mechanisms is a core endeavor of social science.
+Randomized experiments have long served as a gold standard for making valid causal inferences, but most of the data social scientists work with is observational and non-experimental.
+However, with newer methodological developments in economics, political science, epidemiology, and other disciplines, an increasing number of studies in social science make causal claims with observational data.
+As a newer interdisciplinary field, however, nonprofit studies has lagged behind other disciplines in its use of observational causal inference.
+In this paper, we present a hands-on introduction and guide to design-based observational causal inference methods.
+We first review and categorize all studies making causal claims in top nonprofit studies journals over the past decade to illustrate the field’s current of experimental and observational approaches to causal inference.
+We then introduce a framework for modeling and identifying causal processes using directed acyclic graphs (DAGs) and provide a walk-through of the assumptions and procedures for making inferences with a range of different methods, including matching, inverse probability weighting, difference-in-differences, regression discontinuity designs, and instrumental variables.
+We illustrate each approach with synthetic and empirical examples and provide sample R and Stata code for implementing these methods.
+We conclude by encouraging scholars and practitioners to make more careful and explicit causal claims in their observational empirical research, collectively developing and improving quantitative work in the broader field of nonprofit studies.
 keywords: comma, separated, keywords, here
@@
 bibliography: bibliography.bib
-toc: false
+toc: true
 title-page: true
@@
 
-Introduction stuff…
+Causal questions are at the root of many questions in nonprofit studies.
+Do transparency and accountability improve nonprofit performance?
+Does being listed on a transparency-focused website like GuideStar increase donations?
+Other things
+
+Researchers can investigate causal mechanisms using many different methods.
+Qualitative scholars can do focus groups, interviews, constructivist stuff, case studies, whatever—deep dives into specific mechanisms in particular cases.
+In the age of large, publicly available datasets—as well as nonprofits, foundations, and government agencies swimming in data—quantitative research can interrogate causal mechanisms as well.
+But at the root of all this data lies the statistical adage “correlation does not imply causation”.
+In the absence of experimental data, where researchers can randomly assign nonprofits or donors to treatment and control conditions to eliminate any confounding differences between the two groups, making causal claims from quantitative studies is difficult.
+It is not impossible though.
+Advances in econometric, epidemiological, statistical, and computer scientific methods over the past few decades have led to a “credibility revolution,” where researchers can use carefully constructed research designs to estimate causal effects using observational data
+
+In this paper, we provide an introductory primer to thinking about causal questions using observational data. This will allow nonprofit researchers to make more careful causal claims…
 
@@
 The framework here is crucial for general academic research, policy analysis, and program evaluation, among other common types of nonprofit research.
+
+Importantly, in this article with are concerned only with *quantitative* causal inference, or the use of statistical approaches to identify and isolate causal mechanisms.
+We do not assert that quantitative methods are the only avenue for answering causal questions.
+Qualitative and mixed methods like focus groups, action research, ___, ___, and ___ are designed to explore causal mechanisms with rich detail (@cite1).
+Some qualitative methods are at epistemological odds with the whole endeavor of causal effects (@cite2).
+
+Instead, we address the perennial caveat taught in introductory statistics classes: that correlation does not imply causation.
+What *does* imply causation?
+How can we use statistical tools to explore causal questions using observational data?
 
@@
 
-# Intro to DAGs (ANDREW)
+Weasel words - association, determinants, etc. 
+Determinants and prediction is fine! The focus is on getting the most accurate prediction of the outcome
+When making causal claims, though, the focus is on one of the Xs—one lever that a government agency or nonprofit organization can manipulate to affect some sort of change
+
+# Why is causal inference so difficult?
+
+## Fundamental problem of causal inference
+
+Formulating and testing hypotheses is the foundation of scientific inquiry. In chemistry, researchers can intervene with blah blah H2O
+
+People and individuals, however, do not respond in identical ways. Counterfactuals - what would have happened in the absense of an intervention to the same person or organization.
+
+Rubin notation for individual potential outcomes
+Treatment effect = whatever
+
+However, it is impossible to observe or measure both Yi0 and Yi1.
+A charitable foundation interested in poverty reduction cannot give a grant to a nonprofit and simultaneously not give it to the same nonprofit in order to measure the impact of its donation.
+What the recipient nonprofit receives (or does not receive) from the foundation becomes a realized outcome rather than a potential outcome, and the counterfactual outcome is forever unrealized and unmeasurable.
+
+To get around this, we can take the average (or expectation, often indicated with the mathematical function E) of many units that receive the treatment and compare it with the average of comparable units that don’t receive the treatment
+
+Rubin notation: Potential outcomes notation
+
+To help clarify that blah blah, we'll introduce one more piece of mathematical notation - the do() operator
+
+Pearl notation: $\operatorname{do}(x)$ notation = an intervention.
+
+## Selection bias
+
+In addition to the fact that potential outcomes are unobservable, one more characteristic of observational data makes causal inference more complicated.
+
+Individuals and organizations have agency and choose their interventions
+
+A foundation interested in reducing poverty will typically research a range of candidate nonprofits and select the one that fits their own internal criteria.
+Nonprofits applying for a grant from that foundation will tailor their applications to meet the foundation's preferences.
+Accordingly, there are systematic differences between nonprofits that receive a grant and those that don't.
+It is tempting to measure the average outcome of recipient nonprofits and compare it with the average outcome of non-recipient nonprofits, but this estimate would be incorrect.
+
+Formula showing wrong effect
+
+# A vocabulary for observational causal inference (ANDREW)
+
+In the absence of random assignment, finding causal effects in observational data where units of analysis can self select their treatments and outcomes thus appears impossible.
+
+Goal of much research in recent years has been to create methods that allow researchers to make plausible causal arguments using observed, non-experimental data.
+Judea Pearl, computer science, and epidemiology provide us with a special framework for observational causal inference.
+We introduce the vocabulary and grammar of this framework below.
+
+## Causal diagrams
 
@@
 
-In addition to encoding our philosophy and theory of the data generating process, DAGs also serve as an important statistical tool for isolating or identifying causal quantities of interest.
-Identification strategy definition:
-
-> The central role of an identification strategy is to provide a logic for establishing that D is independent of potential values of Y, thus allowing the analyst to interpret observed associations as causal effects.
-[@KeeleStevensonElwert:2020]
-
-## A vocabulary for causal inference
-
-A grammar
-
-Paths and causal structures
-
 - Nodes - nodes can be unmeasurable, or even unobserved
 - Edges - Arrows indicate a relationship, or the passing of *statistical association* between nodes - intervening on one node leads to changes in another node - When two nodes are connected by an arrow, we are stating that there is an assumed causal relationship between those nodes; when there is no arrow between nodes, we are explicitly stating that there is no relationship between the two.
-- d-connection and d-separation - statistical association cannot pass between nodes, either because the arrows are drawn in a way that makes it so information does not link the two, or because conditioning/adjustment blocks the pathway
 - Acyclicity - can't get back to a node - represents flow of time - if things are cyclical, like hiring new staff $\leftrightarrow$ increased capacity, you can make these nodes time-based: hiring new staff_t → increased capacity_t → hiring new staff_t+1 → increased capacity_t+1
@@
 
-# Identification with DAGs (ANDREW)
+## Causal identification
 
-## Experimental causal effects vs. observational causal effects
+In addition to encoding our philosophy and theory of the data generating process, DAGs also serve as an important statistical tool for isolating or identifying causal quantities of interest.
+Identification strategy definition:
 
-Potential outcomes notation
+> The central role of an identification strategy is to provide a logic for establishing that D is independent of potential values of Y, thus allowing the analyst to interpret observed associations as causal effects.
+[@KeeleStevensonElwert:2020]
 
-$\operatorname{do}(x)$ notation
+- d-connection and d-separation - statistical association cannot pass between nodes, either because the arrows are drawn in a way that makes it so information does not link the two, or because conditioning/adjustment blocks the pathway
+
+## *do*-calculus and statistical adjustment
 
@@
 With observational data, we'd like to measure $\mathbf{E}(y \mid \operatorname{do}(x))$ but we can only actually see and measure $\mathbf{E}(y \mid x)$, and as shown in Equation \@ref(eq:correlation-not-causation), these two expressions are not the same.
-This is a formal version of the common phrase "correlation isn't causation":
+This is a formal statement of the phrase "correlation isn't causation":
 
@@
 
-## Stop controlling for everything
+## Bad controls
 
@@
 
+Good/bad controls: <https://papers.ssrn.com/sol3/papers.cfm?abstract_id=3689437> - <https://twitter.com/analisereal/status/1512596580632707078>
+
 Table 2 fallacy + @KeeleStevensonElwert:2020 thing
@@
 
-# How to adjust (MENG AND ANDREW)
+# Adjustment-based identification (MENG AND ANDREW)
 
@@
 
-## Stratification
+## Backdoor adjustment
 
-## Regression
+The logic of *do*-calculus tells us what nodes or variables been to be adjusted for to isolate the treatment → outcome arrow, but the DAG provides no guidance about how to actually make these adjustments.
 
-## Matching
+There are many different ways—regression, matching, IPW—each with different nuances.
+
+### Regression
+
+### Matching
 
@@
 
-## IPW
+### IPW
 
@@
 
+## Illustration
+
+### Simulated example
+
+### Real example
+
+## Frontdoor adjustment
+
+A less common approach to removing confounding is frontdoor adjustment
+
 ## Other methods?
@@
 
-# Design-based causal inference with DAGs (ANDREW AND MENG)
+# Design-based identification (ANDREW AND MENG)
 
 [@AngristPischke:2009; @AngristPischke:2015]
+
 The language of causal graphs, identification, and adjustment provide a universal grammar for discussing causal effects.
@@
 
+Quasi-experiments (design-based—special situations)
+
+- Diff-in-diff
+- RDD
+- IV
+
 ## Experiments
@@
 Look at that one knitted Rmd on RCT FAQs: https://macartan.github.io/i/notes/rct_faqs.html - CONSORT also says to stop doing balance tests - only really need to control for things that might be predictive (https://twitter.com/statsepi/status/1115902270888128514?s=21), but theoretically anything that influences the allocation to treatment is taken care of by randomization.
-
-## Instrumental variables
-
-IVs have to meet the exclusion restriction - the instrument can only influence the outcome through the treatment.
-DAGs make this assumption very clear.
-There cannot be an arrow connecting the instrument to the outcome.
-DAGs also inform the exogeneity assumption—no other nodes in the graph can feed into the instrument node
-
-It's like frontdoor adjustment (https://www.stat.cmu.edu/~cshalizi/402/lectures/23-causal-estimation/lecture-23.pdf) - Really we're finding the causal effect of I on X, then X on Y, generally through 2SLS
 
@@
 
+## Instrumental variables
+
+IVs have to meet the exclusion restriction - the instrument can only influence the outcome through the treatment.
+DAGs make this assumption very clear.
+There cannot be an arrow connecting the instrument to the outcome.
+DAGs also inform the exogeneity assumption—no other nodes in the graph can feed into the instrument node
+
+It's like frontdoor adjustment (https://www.stat.cmu.edu/~cshalizi/402/lectures/23-causal-estimation/lecture-23.pdf) - Really we're finding the causal effect of I on X, then X on Y, generally through 2SLS
+
 # Tools for researchers and practitioners (BOTH)
@@
 
+## Dealing with time
+
+- Epidemiology and polsci and marginal structural models
+- Economics and TWFE and synthetic controls and event studies - fancier diff-in-diff
+
 # Conclusion (BOTH)
+
+A call for more causation / more careful thinking around causation. Prediction is fine. But don't automatically run away from causal work.

### 2023-12-15 README.qmd
Massive Quarto-ification

--- 2021-09-19 README.Rmd
@@
 ---
-output: github_document
+format: gfm
 ---
 
-<!-- README.md is generated from README.Rmd. Please edit that file -->
+<!-- README.md is generated from README.qmd. Please edit that file -->
 
@@
 
-```{r generate-code-name}
+```{r generate-code-name, eval=FALSE}
 library(codename)
 codename_message()
+#> code name generated by {codename} v.0.5.0. R version 4.3.1 (2023-06-16).
 
 codename(type = "ubuntu", seed = 1022)
+#> [1] "corrupt caterpillar"
 ```

### 2023-12-15 manuscript/manuscript.qmd
Massive Quarto-ification

--- 2022-04-13 manuscript/manuscript.Rmd
@@
 short-title: "Clarifying Correlation and Causation"
+date: "December 14, 2023"
 author:
+- name: Meng Ye
+  affiliation: Georgia State University
+  email: mye2@gsu.edu
+  url: https://aysps.gsu.edu/phd-student/ye-meng/
+  orcid: 0000-0002-6552-8338
 - name: Andrew Heiss
-  affiliation: Georgia State University
+  affiliations: 
+    - id: gsu
+      name: Georgia State University
+      department: Andrew Young School of Policy Studies, Department of Public Management and Policy
+      address: "55 Park Place NE, #464"
+      city: Atlanta
+      region: GA
+      country: USA
+      postal-code: 30303
   email: aheiss@gsu.edu
   url: https://www.andrewheiss.com/
-- name: Meng Ye
-  affiliation: Georgia State University
-  email: mye2@student.gsu.edu
-  url: https://aysps.gsu.edu/phd-student/ye-meng/
-date: "April 13, 2022"
+  orcid: 0000-0002-3948-3914
+  attributes:
+    corresponding: true
+correspondence-prefix: "Correspondence concerning this article should be addressed to"
 published: Drafting
-code-repo: "Access the code at <https://github.com/andrewheiss/corrupt-caterpillar>"
-abstract: >-
+abstract: |
   Discovering causal relationships and testing theoretical mechanisms is a core endeavor of social science.
@@
 We conclude by encouraging scholars and practitioners to make more careful and explicit causal claims in their observational empirical research, collectively developing and improving quantitative work in the broader field of nonprofit studies.
-keywords: comma, separated, keywords, here
+thanks: |
+  Acknowledgments here if any
+epigraph:
+  - text: "Don't let us forget that the causes of human actions are usually immeasurably more complex and varied than our subsequent explanations of them."
+    source: "Fyodor Dostoevsky, *The Idiot*, part IV, chapter 2"
+keywords:
+  - those
+  - go
+  - here
 word-count: ""
-thanks: >-
-  Acknowledgments here if any
-reference-section-title: References
-bibliography: bibliography.bib
-toc: true
-title-page: true
-endnotes: false
-endfloat: false
-mainfont: Cochineal
-sansfont: Assistant
-monofont: '`r if (Sys.info()["sysname"] == "Windows") "InconsolataGo Nerd Font" else "InconsolataGo"`'
-ms_mainfont: '`r if (Sys.info()["sysname"] == "Windows") "Linux Libertine" else "Linux Libertine O"`'
-ms_sansfont: Assistant
-ms_monofont: '`r if (Sys.info()["sysname"] == "Windows") "InconsolataGo Nerd Font" else "InconsolataGo"`'
-author-note: >-
-  Author note here
-correspondence: >-
-  Correspondence note here
-editor_options: 
-  chunk_output_type: console
+bibliography: references.json
+link-citations: true
 ---
@@
 ```{r setup, include=FALSE}
-# Figure out output format
-is_docx <- knitr::pandoc_to("docx") | knitr::pandoc_to("odt")
-is_latex <- knitr::pandoc_to("latex")
-is_html <- knitr::pandoc_to("html")
+if (is.null(knitr::pandoc_to())) {
+  fmt_out <- "interactive"
+} else {
+  fmt_out <- knitr::pandoc_to()
+}
 
-# Word-specific things
-table_format <- ifelse(is_docx, "huxtable", "kableExtra")  # Huxtable tables
-conditional_dpi <- ifelse(is_docx, 300, 300)  # Higher DPI
-conditional_align <- ifelse(is_docx, "default", "center")  # Word doesn't support align
-
-# Knitr options
 knitr::opts_chunk$set(
-  echo = FALSE, warning = FALSE, message = FALSE,
-  tidy.opts = list(width.cutoff = 120),  # Code width
-  fig.retina = 3, dpi = conditional_dpi,
-  fig.width = 7, fig.asp = 0.618,
-  fig.align = conditional_align, out.width = "100%",
-  fig.path = "output/figures/",
-  cache.path = "output/_cache/",
-  fig.process = function(x) {  # Remove "-1" from figure names
-    x2 = sub('-\\d+([.][a-z]+)$', '\\1', x)
-    if (file.rename(x, x2)) x2 else x
-  }
+  echo = FALSE, include = FALSE,
+  warning = FALSE, message = FALSE
 )
 
-# R options
+knitr::opts_chunk$set(
+  fig.align = "center", fig.retina = 3,
+  fig.width = 6, fig.height = (6 * 0.618),
+  out.width = "100%", collapse = TRUE
+)
+
 options(
-  width = 90,  # Output width
-  dplyr.summarise.inform = FALSE,  # Turn off dplyr's summarize() auto messages
-  knitr.kable.NA = "",  # Make NAs blank in kables
-  kableExtra.latex.load_packages = FALSE,  # Don't add LaTeX preamble stuff
-  modelsummary_default = table_format  # Set modelsummary backend
+  digits = 3, width = 120,
+  dplyr.summarise.inform = FALSE,
+  knitr.kable.NA = ""
 )
@@
 library(targets)
-library(kableExtra)
 library(here)
 
-withr::with_dir(here::here(), {
-  # Load stuff from targets
-  source(tar_read(plot_funs))
-})
+# Targets stuff
+tar_config_set(
+  store = here::here("_targets"),
+  script = here::here("_targets.R")
+)
 
-clrs_np <- np_palette()
+# Plotting functions
+invisible(list2env(tar_read(graphic_functions), .GlobalEnv))
 ```
@@
 
-```{r dag-associations, fig.width=9, fig.height=2.75, fig.asp=NULL, fig.cap="Three types of relationships in DAGs"}
+```{r}
+#| label: fig-dag-associations
+#| fig-width: 9
+#| fig-height: 2.75
+#| fig-cap: "Three types of relationships in DAGs"
+#| include: true
+
 data_confounder <- dagify(
@@
   tidy_dagitty() %>% 
-  mutate(arrow_color = ifelse(name == "Z", clrs_np$orange, "grey80")) %>% 
+  mutate(arrow_color = ifelse(name == "Z", clrs$orange, "grey80")) %>% 
   mutate(node_color = ifelse(name == "Z", TRUE, FALSE))
@@
   geom_dag_text(data = filter(data_confounder, name == "Z"), color = "white", size = 4) +
-  scale_color_manual(values = c("grey80", clrs_np$orange), guide = "none") +
+  scale_color_manual(values = c("grey80", clrs$orange), guide = "none") +
   coord_cartesian(ylim = c(0.95, 2.05)) +
@@
   theme_np_dag() +
-  theme(plot.title = element_text(hjust = 0.5, color = clrs_np$orange),
-        plot.subtitle = element_text(hjust = 0.5, color = clrs_np$orange))
+  theme(plot.title = element_text(hjust = 0.5, color = clrs$orange),
+        plot.subtitle = element_text(hjust = 0.5, color = clrs$orange))
 
@@
   mutate(arrow_color = case_when(
-    name == "Z" & to == "Y" ~ clrs_np$purple, 
-    name == "X" & to == "Z" ~ clrs_np$purple,
+    name == "Z" & to == "Y" ~ clrs$purple, 
+    name == "X" & to == "Z" ~ clrs$purple,
     TRUE ~ "grey80")) %>% 
@@
   geom_dag_text(data = filter(data_mediator, name == "Z"), color = "white", size = 4) +
-  scale_color_manual(values = c("grey80", clrs_np$purple), guide = "none") +
+  scale_color_manual(values = c("grey80", clrs$purple), guide = "none") +
   coord_cartesian(ylim = c(0.95, 2.05)) +
@@
   theme_np_dag() +
-  theme(plot.title = element_text(hjust = 0.5, color = clrs_np$purple),
-        plot.subtitle = element_text(hjust = 0.5, color = clrs_np$purple))
+  theme(plot.title = element_text(hjust = 0.5, color = clrs$purple),
+        plot.subtitle = element_text(hjust = 0.5, color = clrs$purple))
 
@@
   tidy_dagitty() %>% 
-  mutate(arrow_color = ifelse(to == "Z", clrs_np$red, "grey80")) %>% 
+  mutate(arrow_color = ifelse(to == "Z", clrs$red, "grey80")) %>% 
   mutate(node_color = ifelse(name == "Z", TRUE, FALSE))
@@
   geom_dag_text(data = filter(data_collider, name == "Z"), color = "white", size = 4) +
-  scale_color_manual(values = c("grey80", clrs_np$red), guide = "none") +
+  scale_color_manual(values = c("grey80", clrs$red), guide = "none") +
   coord_cartesian(ylim = c(0.95, 2.05)) +
@@
   theme_np_dag() +
-  theme(plot.title = element_text(hjust = 0.5, color = clrs_np$red),
-        plot.subtitle = element_text(hjust = 0.5, color = clrs_np$red))
+  theme(plot.title = element_text(hjust = 0.5, color = clrs$red),
+        plot.subtitle = element_text(hjust = 0.5, color = clrs$red))
 
@@
 
-This stuff in Figure \@ref(fig:dag-associations): (@Elwert:2013 for forks, chains, inverted forks terminology)
+This stuff in @fig-dag-associations: (@Elwert:2013 for forks, chains, inverted forks terminology)
 
@@
 The language of causal graphs, identification, and adjustment provide a universal grammar for discussing causal effects.
-Commonly used approaches in econometrics and other social science disciplines can be written as causal graphs (see Figure \@ref(fig:design-based-dags))
+Commonly used approaches in econometrics and other social science disciplines can be written as causal graphs (see @fig-design-based-dags))
 
-```{r design-based-dags, fig.width=9, fig.height=6, fig.asp=NULL, fig.cap="Possible DAGs for common design-based experimental and quasi-experimental approaches to causal inference.
+```{r}
+#| label: fig-design-based-dags
+#| fig-width: 9
+#| fig-height: 6
+#| fig-cap: "Possible DAGs for common design-based experimental and quasi-experimental approaches to causal inference.
 Red arrows represent the identified and isolated relationship between treatment $x$ and outcome $y$.
-Square nodes represent statistical adjustment.", warning=FALSE}
+Square nodes represent statistical adjustment."
+#| include: true
+
 data_rct <- dagify(
@@
   node_status() %>% 
-  mutate(arrow_color = ifelse(name == "X", clrs_np$red, "grey80"))
+  mutate(arrow_color = ifelse(name == "X", clrs$red, "grey80"))
 
@@
   annotate(geom = "text", x = 1, y = 1.2, label = "X = x") +
-  scale_color_manual(values = c(clrs_np$green, clrs_np$blue), 
-                     na.value = clrs_np$yellow, guide = "none") +
+  scale_color_manual(values = c(clrs$green, clrs$blue), 
+                     na.value = clrs$yellow, guide = "none") +
   coord_cartesian(ylim = c(0.95, 2.05)) +
@@
   geom_dag_text(size = 4) +
-  scale_color_manual(values = c(clrs_np$green, "grey80", clrs_np$blue), 
-                     na.value = clrs_np$yellow, guide = "none") +
+  scale_color_manual(values = c(clrs$green, "grey80", clrs$blue), 
+                     na.value = clrs$yellow, guide = "none") +
   scale_shape_manual(values = c(15, 19), guide = "none") +
@@
                 nudge_y = -0.37, size = 4) +
-  scale_color_manual(values = c(clrs_np$green, clrs_np$blue), 
-                     na.value = clrs_np$yellow, guide = "none") +
+  scale_color_manual(values = c(clrs$green, clrs$blue), 
+                     na.value = clrs$yellow, guide = "none") +
   scale_shape_manual(values = c(15, 19), guide = "none") +
@@
   node_status() %>% 
-  mutate(arrow_color = ifelse(name == "X", clrs_np$red, "grey80"),
+  mutate(arrow_color = ifelse(name == "X", clrs$red, "grey80"),
          name = recode(name, "Running" = "Running\nvariable"),
@@
                 nudge_y = -0.35, size = 4) +
-  scale_color_manual(values = c(clrs_np$green, "grey80", clrs_np$blue), 
-                     na.value = clrs_np$yellow, guide = "none") +
+  scale_color_manual(values = c(clrs$green, "grey80", clrs$blue), 
+                     na.value = clrs$yellow, guide = "none") +
   scale_shape_manual(values = c(15, 19), guide = "none") +

### 2023-12-15 R/funs_graphics.R

@@ -1,0 +1,148 @@
+# Windows fonts
+register_fonts <- function() {
+  if (Sys.info()["sysname"] == "Windows") {
+    # Register these fonts if running R on Windows
+    windowsFonts(`Assistant` = windowsFont("Assistant"))
+    windowsFonts(`Assistant Light` = windowsFont("Assistant Light"))
+    windowsFonts(`Assistant SemiBold` = windowsFont("Assistant SemiBold"))
+  }
+}
+
+
+# ggplot themes -----------------------------------------------------------
+set_annotation_fonts <- function() {
+  ggplot2::update_geom_defaults(
+    "label",
+    list(family = "Assistant", fontface = "bold", color = "black")
+  )
+  ggplot2::update_geom_defaults(
+    "text",
+    list(family = "Assistant", fontface = "bold", color = "black")
+  )
+
+  if (isNamespaceLoaded("ggdag")) {
+    ggplot2::update_geom_defaults(
+      ggdag:::GeomDagText,
+      list(family = "Assistant", fontface = "bold", color = "black")
+    )
+  }
+}
+
+#' theme_np
+#'
+#' A custom ggplot2 theme used throughout this project
+#'
+#' @param base_size base font size (default is 11)
+#' @param base_family base font family (default is Assistant)
+#'
+#' @export
+#'
+#' @examples
+#' library(ggplot2)
+#'
+#' ggplot(mpg, aes(x = displ, y = hwy, color = drv)) +
+#'   geom_point() +
+#'   labs(x = "Engine displacement", y = "Highway miles per gallon",
+#'        title = "A standard graph about cars",
+#'        subtitle = "Heavier cars get lower mileage",
+#'        caption = "Source: A dataset included in ggplot2",
+#'        color = "Drive") +
+#'   theme_np()
+theme_np <- function(base_size = 11, base_family = "Assistant") {
+  theme_bw(base_size, base_family) + 
+    theme(plot.title = element_text(size = rel(1.4), face = "bold",
+                                    family = "Assistant"),
+          plot.subtitle = element_text(size = rel(1), face = "plain",
+                                       family = "Assistant Light"),
+          plot.caption = element_text(size = rel(0.8), color = "grey50", face = "plain",
+                                      family = "Assistant Light",
+                                      margin = margin(t = 10), hjust = 0),
+          plot.tag = element_text(size = rel(1.4), face = "bold",
+                                  family = "Assistant"),
[…]

### 2023-12-15 R/funs_knitting.R

@@ -1,221 +1,0 @@
-suppressPackageStartupMessages(library(dplyr))
-library(purrr)
-library(stringr)
-library(bib2df)
-library(rvest)
-library(xml2)
-library(stringi)
-
-render_html <- function(input, output, csl, support_folder, ...) {
-  # Add CSS file as a dependency so it goes into the _files directory
-  dep <- htmltools::htmlDependency(
-    name = "ath",
-    version = "1.0.0",
-    "pandoc/css",
-    stylesheet = "ath-clean.css"
-  )
-  extra_dependencies <- list(dep)
-
-  # IMPORTANT
-  # When knitting to docx, bookdown deletes the _files directory for whatever
-  # reason, so if you knit to HTML and then docx, you get a nice *_files
-  # directly that then disappears when the Word file is done. One way around
-  # this is to specify lib_dir here in rmarkdown::render()
-
-  out <- rmarkdown::render(
-    input = input,
-    output_file = output,
-    bookdown::html_document2(
-      template = "pandoc/templates/html.html",
-      pandoc_args = c("--metadata", "link-citations=true",
-                      "--metadata", "linkReferences=true",
-                      paste0("--csl=", csl)),
-      md_extensions = "+raw_tex+smart-autolink_bare_uris+ascii_identifiers",
-      toc = TRUE,
-      number_sections = FALSE,
-      self_contained = FALSE,
-      theme = NULL,
-      extra_dependencies = extra_dependencies,
-      lib_dir = support_folder,
-      includes = list(before_body = "pandoc/colors-annotations.html")
-    ),
-    encoding = "UTF-8"
-  )
-
-  return(fs::path_rel(out))
-}
-
-render_pdf <- function(input, output, bibstyle, ...) {
-  out <- rmarkdown::render(
-    input = input,
-    output_file = output,
-    bookdown::pdf_document2(
-      template = "pandoc/templates/xelatex.tex",
-      latex_engine = "xelatex",
-      dev = "cairo_pdf",
-      pandoc_args = c("--top-level-division=section",
-                      "--shift-heading-level-by=0",
-                      "-V", bibstyle,
-                      "-V", "chapterstyle=hikma-article"),
-      md_extensions = "+raw_tex+smart-autolink_bare_uris",
-      toc = FALSE,
-      keep_tex = TRUE,
-      citation_package = "biblatex",
-      includes = list(in_header = "pandoc/colors-annotations.tex")
-    ),
[…]

### 2023-12-15 R/graphics.R

@@ -1,178 +1,0 @@
-# Windows fonts
-if (Sys.info()["sysname"] == "Windows") {
-  # Register these fonts if running R on Windows
-  windowsFonts(`Assistant` = windowsFont("Assistant"))
-  windowsFonts(`Assistant Light` = windowsFont("Assistant Light"))
-  windowsFonts(`Assistant SemiBold` = windowsFont("Assistant SemiBold"))
-}
-
-
-# ggplot themes -----------------------------------------------------------
-# Set default fonts for text-based geoms
-update_geom_defaults("text",
-                     list(family = "Assistant", 
-                          fontface = "bold",
-                          color = "black"))
-
-update_geom_defaults(ggdag:::GeomDagText,
-                     list(family = "Assistant", 
-                          fontface = "bold",
-                          color = "black"))
-
-#' theme_np
-#'
-#' A custom ggplot2 theme used throughout this project
-#'
-#' @param base_size base font size (default is 11)
-#' @param base_family base font family (default is Assistant)
-#'
-#' @export
-#'
-#' @examples
-#' library(ggplot2)
-#'
-#' ggplot(mpg, aes(x = displ, y = hwy, color = drv)) +
-#'   geom_point() +
-#'   labs(x = "Engine displacement", y = "Highway miles per gallon",
-#'        title = "A standard graph about cars",
-#'        subtitle = "Heavier cars get lower mileage",
-#'        caption = "Source: A dataset included in ggplot2",
-#'        color = "Drive") +
-#'   theme_np()
-theme_np <- function(base_size = 11, base_family = "Assistant") {
-  theme_bw(base_size, base_family) + 
-    theme(plot.title = element_text(size = rel(1.4), face = "bold",
-                                    family = "Assistant"),
-          plot.subtitle = element_text(size = rel(1), face = "plain",
-                                       family = "Assistant Light"),
-          plot.caption = element_text(size = rel(0.8), color = "grey50", face = "plain",
-                                      family = "Assistant Light",
-                                      margin = margin(t = 10), hjust = 0),
-          plot.tag = element_text(size = rel(1.4), face = "bold",
-                                  family = "Assistant"),
-          panel.border = element_rect(color = "grey50", fill = NA, size = 0.15),
-          panel.spacing = unit(1, "lines"),
[…]

### 2023-12-15 _targets.R

@@ -2,19 +2,22 @@ library(targets)
 library(tarchetypes)
 
 # General pipeline settings -----------------------------------------------
-csl <- "pandoc/csl/apa.csl"
-bibstyle <- "bibstyle-apa"
-
-options(tidyverse.quiet = TRUE,
-        dplyr.summarise.inform = FALSE)
+options(
+  tidyverse.quiet = TRUE,
+  dplyr.summarise.inform = FALSE
+)
 
 set.seed(4393)  # From random.org
 
+# Global target options
+tar_option_set(
+  packages = c("tidyverse"),  # Packages available to all targets
+  format = "qs",  # Storage format
+  workspace_on_error = TRUE  # Automatically create a debug workspace on errors
+)
+
 
 # R functions -------------------------------------------------------------
-
-# source("R/whatever.R")
-source("R/funs_knitting.R")
 
 # here::here() returns an absolute path, which then gets stored in tar_meta and
 # becomes computer-specific (i.e. /Users/andrew/Research/blah/thing.Rmd).
@@ -23,72 +26,27 @@ source("R/funs_knitting.R")
 # https://github.com/r-lib/here/issues/36#issuecomment-530894167)
 here_rel <- function(...) {fs::path_rel(here::here(...))}
 
+# Load all the scripts in the R/ folder that contain the functions to be used in
+# the pipeline
+tar_source()
+
 
 # Pipeline ----------------------------------------------------------------
 
 list(
   
-  # Define helper functions
-  tar_target(plot_funs, here_rel("R", "graphics.R"), format = "file"),
+  ## Helper functions ----
+  tar_target(graphic_functions, lst(
+    register_fonts, set_annotation_fonts,  
+    theme_np, theme_np_dag, 
+    clrs, get_latex_rgb, print.color_table)
+  ),
 
-  ## Manuscript targets ------------------------------------------------------
+  ## Manuscript targets ----
 
-  # tarchetypes::tar_render() automatically detects target dependencies in Rmd
-  # files and knits them, but there's no easy way to pass a custom rendering
-  # script like bookdown::html_document2(), so two things happen here:
-  #   1. Set a file-based target with tar_target_raw() and use tar_knitr_deps()
-  #      to detect the target dependencies in the Rmd file
-  #   2. Use a bunch of other file-based targets to actually render the document
-  #      through different custom functions
-  tar_target(bib_file,
-             here_rel("manuscript", "bibliography.bib"),
[…]

### 2023-12-15 _targets.R
Create targets for manuscript

@@ -44,8 +44,8 @@ list(
 
   ## Manuscript targets ----
 
-  # tar_quarto(manuscript_nice, path = "manuscript", quiet = FALSE, profile = "nice"),
-  # tar_quarto(manuscript_manuscripty, path = "manuscript", quiet = FALSE, profile = "ms"),
+  tar_quarto(manuscript_nice, path = "manuscript", quiet = FALSE, profile = "nice"),
+  tar_quarto(manuscript_manuscripty, path = "manuscript", quiet = FALSE, profile = "ms"),
   
   ## Render the README ----
   tar_quarto(readme, here_rel("README.qmd"))

### 2023-12-15 manuscript/manuscript.qmd
Metadata changes

@@
 author:
-- name: Meng Ye
-  affiliation: Georgia State University
-  email: mye2@gsu.edu
-  url: https://aysps.gsu.edu/phd-student/ye-meng/
-  orcid: 0000-0002-6552-8338
 - name: Andrew Heiss
@@
     corresponding: true
+- name: Meng Ye
+  affiliation: Georgia State University
+  email: mye2@gsu.edu
+  url: https://aysps.gsu.edu/phd-student/ye-meng/
+  orcid: 0000-0002-6552-8338
 correspondence-prefix: "Correspondence concerning this article should be addressed to"

### 2023-12-16 manuscript/manuscript.qmd
Use new Ma et al citekey

@@
 
-Similar to recent work by @BaBerrettCoupet:2021 and @Ma-etal:2021 and @Rohrer:2018, in this paper, we provide a introductory primer / framework for approaching causal questions with observational data - accessible and approachable guide to nonprofit researchers and practitioners.
+Similar to recent work by @BaBerrettCoupet:2021 and @MaEbeiddeWit:2023 and @Rohrer:2018, in this paper, we provide a introductory primer / framework for approaching causal questions with observational data - accessible and approachable guide to nonprofit researchers and practitioners.
 The framework here is crucial for general academic research, policy analysis, and program evaluation, among other common types of nonprofit research.

### 2023-12-16 manuscript/manuscript.qmd
Fix underbrackets in math when using xelatex and unicode-math
https://github.com/latex3/mathtools/issues/29

@@
 invisible(list2env(tar_read(graphic_functions), .GlobalEnv))
+```
+
+```{=latex}
+% xelatex + unicode-math breaks over/underbrackets so we need to save the 
+% bracket environments and then restore them after \begin{document}. 
+% See https://github.com/latex3/mathtools/issues/29
+
+\makeatletter
+\@ifpackageloaded{unicode-math}{
+% Restore mathtools' brackets
+\let\underbracket=\normalunderbracket
+\let\overbracket=\normaloverbracket
+}{}
+\makeatother
 ```
@@
 
-With observational data, we'd like to measure $\mathbf{E}(y \mid \operatorname{do}(x))$ but we can only actually see and measure $\mathbf{E}(y \mid x)$, and as shown in Equation \@ref(eq:correlation-not-causation), these two expressions are not the same.
+With observational data, we'd like to measure $\mathbf{E}(y \mid \operatorname{do}(x))$ but we can only actually see and measure $\mathbf{E}(y \mid x)$, and as shown in @eq-correlation-not-causation, these two expressions are not the same.
 This is a formal statement of the phrase "correlation isn't causation":
 
-\begin{equation}
-{\color{gray} \overbrace{{\color{orange} \underbracket[0.25pt]{{\color{black} \mathbf{E}(y \mid \operatorname{do}(x)) \vphantom{\frac{1}{2}}}}_{\color{orange} \text{Causation}}}}^{\color{gray} \mathclap{\substack{\text{The average} \\ \text{population-level} \\ \text{change in $y$ when} \\ \textit{directly intervening} \\ \text{(or doing) $x$}}}}}
+$$
+{\color{gray} \overbrace{{\color{orange} \underbracket[0.25pt]{{\color{black} \mathbf{E}(y \mid \operatorname{do}(x)) \vphantom{\frac{1}{2}}}}_{\color{orange} \text{``Causation"}}}}^{\color{gray} \mathclap{\substack{\text{The average} \\ \text{population-level} \\ \text{change in $y$ when} \\ \textit{directly intervening} \\ \text{(or doing) $x$}}}}}
 \quad \neq \quad
-{\color{gray} \overbrace{\color{purple} \underbracket[0.25pt]{{\color{black} \mathbf{E}(y \mid x)} \vphantom{\frac{1}{2}}}_{\color{purple} \text{Correlation}}}^{\color{gray} \mathclap{\substack{\text{The average} \\ \text{population-level} \\ \text{change in $y$ when} \\ \text{accounting for} \\ \textit{observed } x}}}}
-(\#eq:correlation-not-causation)
-\end{equation}
+{\color{gray} \overbrace{\color{purple} \underbracket[0.25pt]{{\color{black} \mathbf{E}(y \mid x)} \vphantom{\frac{1}{2}}}_{\color{purple} \text{``Correlation"}}}^{\color{gray} \mathclap{\substack{\text{The average} \\ \text{population-level} \\ \text{change in $y$ when} \\ \text{accounting for} \\ \textit{observed } x}}}}
+$$ {#eq-correlation-not-causation}
 
@@
 By adjusting or controlling for nodes that open up backdoor paths between the treatment and outcome nodes, we can mathematically transform a $\operatorname{do}(\cdot)$ expression into something based solely on observational data.
-Formally, the backdoor adjustment formula is defined in Equation \@ref(eq:backdoor):
+Formally, the backdoor adjustment formula is defined in @eq-backdoor:
 
-\begin{equation}
+$$
 {\color{gray} \overbrace{\color{black} \mathbf{E}(y \mid \operatorname{do}(x)) \vphantom{\frac{1}{2}}}^{\substack{\text{Causal effect} \\ \text{of $x$ on $y$}}}}
-=
+\quad=\quad
 {\color{gray} \underbrace{{\color{black} \sum_z}}_{\mathclap{\substack{\text{Sum across} \\ \text{all values of $z$}}}}}
-{\color{gray} \overbrace{\color{black} \mathbf{E} (y \mid x, z) \vphantom{\frac{1}{2}}}^{\mathclap{\substack{\text{Conditional} \\ \text{mean of $y$,} \\ \text{given $x$ and $z$$\dots$}}}}}
-\times 
+{\color{gray} \overbrace{\color{black} \mathbf{E} (y \mid x, z) \vphantom{\frac{1}{2}}}^{\mathclap{\substack{\text{Conditional} \\ \text{mean of $y$,} \\ \text{given $x$ and $z\dots$}}}}}
+\enspace\times\enspace
 {\color{gray} \overbrace{\color{black} \mathbf{P}(z) \vphantom{\frac{1}{2}}}^{\mathclap{\substack{\text{$\dots$ weighted} \\ \text{by $z$}}}}}
-(\#eq:backdoor)
-\end{equation}
+$$ {#eq-backdoor}
 

### 2024-11-08 _targets.R
Stop using quarto profiles

@@ -44,8 +44,7 @@ list(
 
   ## Manuscript targets ----
 
-  tar_quarto(manuscript_nice, path = "manuscript", quiet = FALSE, profile = "nice"),
-  tar_quarto(manuscript_manuscripty, path = "manuscript", quiet = FALSE, profile = "ms"),
+  tar_quarto(manuscript, path = "manuscript", working_directory = "manuscript", quiet = FALSE),
   
   ## Render the README ----
   tar_quarto(readme, here_rel("README.qmd"))

### 2024-11-08 manuscript/manuscript.qmd
Add outline for ARNOVA paper

@@
 short-title: "Clarifying Correlation and Causation"
-date: "December 14, 2023"
+date: 2024-11-08
 author:
@@
 options(
-  digits = 3, width = 120,
-  dplyr.summarise.inform = FALSE,
-  knitr.kable.NA = ""
+  digits = 3, width = 120
 )
@@
 ```
+
+::: {.content-visible when-format="html"}
+# Introduction {.hide}
+:::
+
+::: {.callout-note}
+# ARNOVA outline
+
+Cleaned up, simplified outline that I'll expand and integrate with the rest of the outline/draft below
+:::
+
+Causal inference important in social sciences, but hard. Nobel prize for Angrist and Imbens and gang...
+
+Answering causal questions is an important endeavor for nonprofit researchers and practitioners.
+Nonprofits serve social goals and create and implement a range of programs to accomplish those goals.
+Measuring the impact of nonprofit programming is essential both for grantors that require evidence of program impact and for nonprofit program managers who are interested in whether their programs lead to the desired outcomes.
+
+Causal work in the nonprofit policy and management literature has increased in the past decade (see fig1), but it remains a less explored topic.
+The nonprofit literature is full of careful quantitative and statistical work, explicitly descriptive, implicitly causal (with weasel words), and explicitly causal.
+
+In this paper, we provide a practical reference, guide, and review of modern causal inference methods.
+For the sake of accessibility, we provide sample data, R and Stata code, minimal—and extensively annotated—mathematical notation.
+We offer a unified, cross-disciplinary framework for thinking about causal identification, spanning econometrics and epidemiology, providing a toolbox for nonprofit researchers to think more carefully about causally-oriented research design and analysis, leading to more effective evaluations of programs, policies, and interventions.
+
+Roadmap—in this paper we blah blah blah
+
+# The fundamental problem of casual inference
+
+Explain that here…
+
+# Visualizing, identifying, and isolating causal relationships with DAGs
+
+General explanation of DAGs and confounding/colliding/mediating and d-separation and do-calculus, etc.
+
+# Isolating causal relationships
+
+To identify a causal relationship in a DAG, we have to ensure that the arrow between the treatment and outcome nodes is isolated and that the treatment and outcome nodes are not linked through any other pathways
+
+There are two general approaches for doing this:
+
+- Circumstantial identification
+- Adjustment-based identification
+
+# Situation-based approaches
+
+## RCTs
+
+- Explanation + DAG
+- Illustration of how to use it
+- Review of existing nonprofit studies that use it
+- Possible nonprofit research questions that could use it
+
+## Diff-in-diff + TWFE
+
+- Explanation + DAG
+- Illustration of how to use it
+- Review of existing nonprofit studies that use it
+- Possible nonprofit research questions that could use it
+
+## RDD
+
+- Explanation + DAG
+- Illustration of how to use it
+- Review of existing nonprofit studies that use it
+- Possible nonprofit research questions that could use it
+
+## IV
+
+- Explanation + DAG
+- Illustration of how to use it
+- Review of existing nonprofit studies that use it
+- Possible nonprofit research questions that could use it
+
+# Adjustment-based approaches
+
+## Regression adjustment
+
+- Explanation + DAG
+- Illustration of how to use it
+- Review of existing nonprofit studies that use it
+- Possible nonprofit research questions that could use it
+
+## IPW and g-computation
+
+- Explanation + DAG
+- Illustration of how to use it
+- Review of existing nonprofit studies that use it
+- Possible nonprofit research questions that could use it
+
+# Hybrid approaches
+
+Briefly describe combinations of the two approaches, like synthetic controls, DiD with matching, RDD with covariate adjustment, etc.
+
+# Tools for researchers and practitioners
+
+Code in R and Stata? Point to other resources like The Mixtape, The Effect, Pearl's stuff, Morgan and Winship?
+
+# Conclusion
+
+A call for more causation / more careful thinking around causation. Prediction is fine. But don't automatically run away from causal work.
+
+---
 

### 2024-11-08 _targets.R

@@ -9,6 +9,14 @@ options(
 
 set.seed(4393)  # From random.org
 
+# This hardcodes the absolute path in _targets.yaml, so to make this more
+# portable, we rewrite it every time this pipeline is run (and we don't track
+# _targets.yaml with git)
+tar_config_set(
+  store = here::here("_targets"),
+  script = here::here("_targets.R")
+)
+
 # Global target options
 tar_option_set(
   packages = c("tidyverse"),  # Packages available to all targets

### 2024-11-08 manuscript/manuscript.qmd
%>% → |>

@@
                 y = c(X = 1, Y = 1, Z = 2))
-) %>% 
-  tidy_dagitty() %>% 
-  mutate(arrow_color = ifelse(name == "Z", clrs$orange, "grey80")) %>% 
+) |> 
+  tidy_dagitty() |> 
+  mutate(arrow_color = ifelse(name == "Z", clrs$orange, "grey80")) |> 
   mutate(node_color = ifelse(name == "Z", TRUE, FALSE))
@@
                 y = c(X = 1, Y = 1, Z = 2))
-) %>% 
-  tidy_dagitty() %>% 
+) |> 
+  tidy_dagitty() |> 
   mutate(arrow_color = case_when(
@@
     name == "X" & to == "Z" ~ clrs$purple,
-    TRUE ~ "grey80")) %>% 
+    TRUE ~ "grey80")) |> 
   mutate(node_color = ifelse(name == "Z", TRUE, FALSE))
@@
                 y = c(X = 1, Y = 1, Z = 2))
-) %>% 
-  tidy_dagitty() %>% 
-  mutate(arrow_color = ifelse(to == "Z", clrs$red, "grey80")) %>% 
+) |> 
+  tidy_dagitty() |> 
+  mutate(arrow_color = ifelse(to == "Z", clrs$red, "grey80")) |> 
   mutate(node_color = ifelse(name == "Z", TRUE, FALSE))
@@
   outcome = "Y"
-) %>% 
-  tidy_dagitty() %>% 
-  node_status() %>% 
+) |> 
+  tidy_dagitty() |> 
+  node_status() |> 
   mutate(arrow_color = ifelse(name == "X", clrs$red, "grey80"))
@@
                   coords = list(x = c(X = 1, Y = 3, U = 2, I = 0),
-                                y = c(X = 1, Y = 1, U = 2, I = 1))) %>% 
-  tidy_dagitty() %>% 
-  adjust_for(var = "I") %>% 
-  node_status() %>% 
+                                y = c(X = 1, Y = 1, U = 2, I = 1))) |> 
+  tidy_dagitty() |> 
+  adjust_for(var = "I") |> 
+  node_status() |> 
   mutate(arrow_color = ifelse(name == "X", clrs$red, "grey80"))
@@
                          coords = list(x = c(X = 1, Y = 3, Time = 2, Location = 2),
-                                       y = c(X = 1, Y = 1, Time = 2, Location = 0))) %>% 
-  tidy_dagitty() %>% 
-  adjust_for(var = c("Time", "Location")) %>% 
-  node_status() %>% 
+                                       y = c(X = 1, Y = 1, Time = 2, Location = 0))) |> 
+  tidy_dagitty() |> 
+  adjust_for(var = c("Time", "Location")) |> 
+  node_status() |> 
   mutate(arrow_color = ifelse(name == "X", clrs$red, "grey80"))
@@
                    coords = list(x = c(X = 1, Y = 4, Running = 3, Threshold = 2, U = 3),
-                                 y = c(X = 1, Y = 1, Running = 2, Threshold = 1.75, U = 3))) %>% 
-  tidy_dagitty() %>% 
-  adjust_for(var = c("Threshold", "Running")) %>% 
-  node_status() %>% 
+                                 y = c(X = 1, Y = 1, Running = 2, Threshold = 1.75, U = 3))) |> 
+  tidy_dagitty() |> 
+  adjust_for(var = c("Threshold", "Running")) |> 
+  node_status() |> 
   mutate(arrow_color = ifelse(name == "X", clrs$red, "grey80"),

### 2024-11-08 analysis/wcnpd.R

@@ -8,18 +8,18 @@ articles <- read_csv(here("data", "manual_data", "Skimming notes_coding.csv"))
 
 nvsq_counts <- read_csv(here("data", "manual_data", "nvsq_counts.csv"))
 
-nvsq_counts_year <- nvsq_counts %>% 
-  filter(Year <= 2020) %>% 
-  # filter(Year <= 2021 & !(Year == 2021 & Issue > 3)) %>% 
-  # mutate(Year = recode(Year, "2021" = 2020)) %>% 
-  group_by(Year) %>% 
+nvsq_counts_year <- nvsq_counts |> 
+  filter(Year <= 2020) |> 
+  # filter(Year <= 2021 & !(Year == 2021 & Issue > 3)) |> 
+  # mutate(Year = recode(Year, "2021" = 2020)) |> 
+  group_by(Year) |> 
   summarize(total = sum(Count))
 
-plot_data <- articles %>% 
-  filter(year <= 2020) %>% 
-  # mutate(year = recode(year, "2021" = 2020)) %>% 
-  count(year) %>% 
-  left_join(nvsq_counts_year, by = c("year" = "Year")) %>% 
+plot_data <- articles |> 
+  filter(year <= 2020) |> 
+  # mutate(year = recode(year, "2021" = 2020)) |> 
+  count(year) |> 
+  left_join(nvsq_counts_year, by = c("year" = "Year")) |> 
   mutate(pct = n / total)
 
 nvsq_count <- ggplot(plot_data, aes(x = factor(year), y = n)) +

### 2024-11-09 manuscript/manuscript.qmd
Mega additions and edits and reorganization

@@
 link-citations: true
+
+params:
+  wordcount: |
+    <strong>{{< words-sum body-ref >}} total words</strong>: {{< words-body >}} in the body • {{< words-ref >}} in the references
 ---
+
+::: {.content-visible when-meta="wordcount-banner"}
+{{< include _extensions/andrewheiss/wordcount/banner.html >}}
+:::
 
@@
 
-::: {.callout-note}
-# ARNOVA outline
-
-Cleaned up, simplified outline that I'll expand and integrate with the rest of the outline/draft below
-:::
-
 Causal inference important in social sciences, but hard. Nobel prize for Angrist and Imbens and gang...
@@
 
-# The fundamental problem of casual inference
+---
 
-Explain that here…
-
-# Visualizing, identifying, and isolating causal relationships with DAGs
-
-General explanation of DAGs and confounding/colliding/mediating and d-separation and do-calculus, etc.
-
-# Isolating causal relationships
-
-To identify a causal relationship in a DAG, we have to ensure that the arrow between the treatment and outcome nodes is isolated and that the treatment and outcome nodes are not linked through any other pathways
-
-There are two general approaches for doing this:
-
-- Circumstantial identification
-- Adjustment-based identification
-
-# Situation-based approaches
-
-## RCTs
-
-- Explanation + DAG
-- Illustration of how to use it
-- Review of existing nonprofit studies that use it
-- Possible nonprofit research questions that could use it
-
-## Diff-in-diff + TWFE
-
-- Explanation + DAG
-- Illustration of how to use it
-- Review of existing nonprofit studies that use it
-- Possible nonprofit research questions that could use it
-
-## RDD
-
-- Explanation + DAG
-- Illustration of how to use it
-- Review of existing nonprofit studies that use it
-- Possible nonprofit research questions that could use it
-
-## IV
-
-- Explanation + DAG
-- Illustration of how to use it
-- Review of existing nonprofit studies that use it
-- Possible nonprofit research questions that could use it
-
-# Adjustment-based approaches
-
-## Regression adjustment
-
-- Explanation + DAG
-- Illustration of how to use it
-- Review of existing nonprofit studies that use it
-- Possible nonprofit research questions that could use it
-
-## IPW and g-computation
-
-- Explanation + DAG
-- Illustration of how to use it
-- Review of existing nonprofit studies that use it
-- Possible nonprofit research questions that could use it
-
-# Hybrid approaches
-
-Briefly describe combinations of the two approaches, like synthetic controls, DiD with matching, RDD with covariate adjustment, etc.
-
-# Tools for researchers and practitioners
-
-Code in R and Stata? Point to other resources like The Mixtape, The Effect, Pearl's stuff, Morgan and Winship?
-
-# Conclusion
-
-A call for more causation / more careful thinking around causation. Prediction is fine. But don't automatically run away from causal work.
-
----
+Stuff from first version of introduction:
 
@@
 
+---
+
 # Why care about causation in nonprofit studies? (MENG)
@@
 
-# Why is causal inference so difficult?
+---
+
+Textbook treatments of causal inference approahces are typically abstract and generic, explaining how treatment $X$ causes outcome $Y$ after adjusting for confounders $Z$.
+To make these principles as concrete as possible, we use a practical running example relevant to nonprofit education and practice.
+Experiential philanthropy - Newark, BYU Grantwell and Be a Philanthropist.
+These programs provide students interested in the nonprofit sector with hands on, practical experience with philanthropy.
+Beyond their pedagogical aims, these programs are designed to instill a long-term interest and participation in philanthropy.
+
+> Do experiential philanthropy programs cause long-term charitable or prosocial behavior?
+
+We do not provide any evidence to answer this question directly.
+Rather, we use simulated data to explore multiple hypothetical research designs that could be used to quantititavely measure the causal impact of these programs.
+These causal inference approaches can be applied to a wide range of nonprofit-focued questions, though some methods are better suited than others, depending on the context of the research question.
+TODO We provide examples of possible nonprofit-related causal inquiries that are amenable to each method
+
+# A visual vocabulary for causal inference
+
+## Causation through listening
+
+Judea Pearl, computer science, and epidemiology blah blah blah provide us with a specialized framework for describing causal theories and making causal inferences.
+
+At the core of causal inquiry is the notion of "causation," or what it means when we say that an intervention causes an outcome.
+In Pearl's framework, causation can be defined using a metaphor of listening and responding:
+
+> A variable $X$ is a cause of a variable $Y$ if $Y$ in any way relies on $X$ for its value.… $X$ is a cause of $Y$ if $Y$ listens to $X$ and decides its value in response to what it hears.
+[@PearlGlymourJewell:2016, 5--6].
+
+In the context of our example causal question, we posit that prosocial behavior listens to, responds to, or is caused by experiential philanthropy.
+We have a theoretical reason to believe that the intervention and the outcome are associated with each other in a sequential, nonspurious way—it is reasonable to assume that having hands-on experience with philanthropy today would influence future attitudes toward charitable giving.
+Specific educational interventions are not the only cause of prosocial behavior.
+This behavior listens and responds to other factors, including family background, income, education, and personal attitudes and opinions towards volunteerism and charity—experiential philanthropy is just one of many causes.
+
+## Directed acyclic graphs (DAGs)
+
+We can formally represent our theory of how an intervention ($X$) causes an outcome ($Y$) using a directed acyclic graph, or DAG [@PearlMackenzie:2018; @PearlGlymourJewell:2016; @MorganWinship:2014].
+DAGs encode our understanding of the data generating process, or which phenomena cause the treatment, the outcome, and both simultaneously.
+These causal graphs are a philosophical model of the associations between different phenomena and our assumptions behind those relationships.
+
+- **Nodes** - nodes can be unmeasurable, or even unobserved
+- **Edges** - Arrows indicate a relationship, or the passing of *statistical association* between nodes - intervening on one node leads to changes in another node - When two nodes are connected by an arrow, we are stating that there is an assumed causal relationship between those nodes; when there is no arrow between nodes, we are explicitly stating that there is no relationship between the two.
+Unlike other graphical approaches like structural equation models (SEMs) that assume all relationships are additive and linear, DAG arrows are nonparametric and can represent any kind of functional form (linear, multiplicative, polynomial, exponential, etc.).
+Accordingly, when we say that $X \rightarrow Y$, we mean that $X$ causally affects $Y$ in some general way, not necessarily linearly [@Rohrer:2018].
+- **Acyclicity** - can't get back to a node - represents flow of time, or temporal ordering - if things are cyclical, like hiring new staff $\leftrightarrow$ increased capacity, you can make these nodes time-based: hiring new staff_t → increased capacity_t → hiring new staff_t+1 → increased capacity_t+1
+
+```{r}
+#| label: fig-dag-basic-real
+#| fig-width: 4.5
+#| fig-height: 2.75
+#| out-width: 60%
+#| fig-cap: "An example directed acyclic graph that shows how experiential philanthropy causes charitable behavior, while incomes causes both"
+#| include: true
+
+data_confounder_example <- dagify(
+  Y ~ Z + X,
+  X ~ Z,
+  coords = list(x = c(X = 1, Y = 3, Z = 2),
+                y = c(X = 1, Y = 1, Z = 2)),
+  labels = c(X = "Experiential\nphilanthropy", Y = "Charitable\nbehavior", Z = "Income")
+) |> 
+  tidy_dagitty()
+
+ggplot(data_confounder_example, aes(x = x, y = y, xend = xend, yend = yend)) +
+  geom_dag_edges(edge_width = 1) +
+  geom_dag_point(aes(color = name), size = 12) +
+  geom_dag_text(aes(label = name)) +
+  geom_dag_text(
+    data = filter(data_confounder_example, name != "Z"), aes(label = label), 
+    size = 4, nudge_y = -0.32, color = "black", lineheight = 1
+  ) +
+  geom_dag_text(
+    data = filter(data_confounder_example, name == "Z"), aes(label = label), 
+    size = 4, nudge_y = 0.25, color = "black", lineheight = 1
+  ) +
+  scale_color_manual(values = c("X" = clrs$green, "Y" = clrs$blue, "Z" = clrs$yellow), guide = "none") +
+  coord_cartesian(xlim = c(0.75, 3.25), ylim = c(0.6, 2.2)) +
+  theme_np_dag()
+```
+
+Importantly, DAGs do more than simply represent the relationships and associations between variables.
+They provide us with ways to isolate specific relationships of interest.
+TODO: Rohrer, 28: Even if a researcher does not rely fully on a DAG, mapping out the underlying data-generating process
+
+## Statistical associations and causal structures
+
+General explanation of DAGs and confounding/colliding/mediating and d-separation and do-calculus, etc.
+
+```{r}
+#| label: fig-dag-associations
+#| fig-width: 9
+#| fig-height: 2.75
+#| fig-cap: "Three types of relationships in DAGs"
+#| include: true
+
+data_confounder <- dagify(
+  Y ~ Z + X,
+  X ~ Z,
+  coords = list(x = c(X = 1, Y = 3, Z = 2),
+                y = c(X = 1, Y = 1, Z = 2))
+) |> 
+  tidy_dagitty() |> 
+  mutate(arrow_color = ifelse(name == "Z", clrs$orange, "grey80")) |> 
+  mutate(node_color = ifelse(name == "Z", TRUE, FALSE))
+
+confounder <- ggplot(data_confounder, aes(x = x, y = y, xend = xend, yend = yend)) +
+  geom_dag_edges(aes(edge_colour = arrow_color), edge_width = 1) +
+  geom_dag_point(aes(color = node_color), size = 12) +
+  geom_dag_text(data = filter(data_confounder, name != "Z"), size = 4) +
+  geom_dag_text(data = filter(data_confounder, name == "Z"), color = "white", size = 4) +
+  scale_color_manual(values = c("grey80", clrs$orange), guide = "none") +
+  coord_cartesian(ylim = c(0.95, 2.05)) +
+  labs(title = "Confounder") +
+  theme_np_dag() +
+  theme(plot.title = element_text(hjust = 0.5, color = clrs$orange))
+
+data_mediator <- dagify(
+  Y ~ X + Z,
+  Z ~ X,
+  coords = list(x = c(X = 1, Y = 3, Z = 2),
+                y = c(X = 1, Y = 1, Z = 2))
+) |> 
+  tidy_dagitty() |> 
+  mutate(arrow_color = case_when(
+    name == "Z" & to == "Y" ~ clrs$purple, 
+    name == "X" & to == "Z" ~ clrs$purple,
+    TRUE ~ "grey80")) |> 
+  mutate(node_color = ifelse(name == "Z", TRUE, FALSE))
+
+mediator <- ggplot(data_mediator, aes(x = x, y = y, xend = xend, yend = yend)) +
+  geom_dag_edges(aes(edge_colour = arrow_color), edge_width = 1) +
+  geom_dag_point(aes(color = node_color), size = 12) +
+  geom_dag_text(data = filter(data_mediator, name != "Z"), size = 4) +
+  geom_dag_text(data = filter(data_mediator, name == "Z"), color = "white", size = 4) +
+  scale_color_manual(values = c("grey80", clrs$purple), guide = "none") +
+  coord_cartesian(ylim = c(0.95, 2.05)) +
+  labs(title = "Mediator") +
+  theme_np_dag() +
+  theme(plot.title = element_text(hjust = 0.5, color = clrs$purple))
+
+data_collider <- dagify(
+  Y ~ X,
+  Z ~ X + Y,
+  coords = list(x = c(X = 1, Y = 3, Z = 2),
+                y = c(X = 1, Y = 1, Z = 2))
+) |> 
+  tidy_dagitty() |> 
+  mutate(arrow_color = ifelse(to == "Z", clrs$red, "grey80")) |> 
+  mutate(node_color = ifelse(name == "Z", TRUE, FALSE))
+
+collider <- ggplot(data_collider, aes(x = x, y = y, xend = xend, yend = yend)) +
+  geom_dag_edges(aes(edge_colour = arrow_color), edge_width = 1) +
+  geom_dag_point(aes(color = node_color), size = 12) +
+  geom_dag_text(data = filter(data_collider, name != "Z"), size = 4) +
+  geom_dag_text(data = filter(data_collider, name == "Z"), color = "white", size = 4) +
+  scale_color_manual(values = c("grey80", clrs$red), guide = "none") +
+  coord_cartesian(ylim = c(0.95, 2.05)) +
+  labs(title = "Collider") +
+  theme_np_dag() +
+  theme(plot.title = element_text(hjust = 0.5, color = clrs$red))
+
+combined_associations <- (confounder | plot_spacer() | mediator | plot_spacer() | collider) +
+  plot_layout(widths = c(0.32, 0.02, 0.32, 0.02, 0.32))
+combined_associations
+```
+
+This stuff in @fig-dag-associations: (@Elwert:2013 for forks, chains, inverted forks terminology)
+
+- Confounding - forks
+- Mediation - chains
+- Collision - inverted forks [@ElwertWinship:2014; @KnoxLoweMummolo:2020]
+
+Confounding is a major scary issue though - especially if it's unmeasured or unobservable…
+
+## Bad controls
+
+Even if we do not rely on a DAG for complete identification of causal effects, mapping out the relationships between nodes in the data-generating process is a useful exercise in any kind of quantitative, regression-based analysis, including non-causal descriptive work.
+Researchers often take a "kitchen sink" approach to statistical control, including all potential covariates in a model.
+Understanding the underlying DAG provides clearer guidance about what to control for, helping researchers avoid "bad controls" [@cite].
+For instance, if a potential covariate is only associated with the outcome and has no relationship to the treatment, it is not a confounder and need not be included as a control variable in a model, since it can reduce the precision of the model estimates.
+If a potential covariate is a collider, caused by both the treatment and the outcome, the model results will be biased or distorted.
+
+In addition to identification, DAGs provide additional statistical insight and guidance regarding control variables or covariates in regression models
+
+Bad controls and colliders
+
+Good/bad controls: <https://papers.ssrn.com/sol3/papers.cfm?abstract_id=3689437> - <https://twitter.com/analisereal/status/1512596580632707078>
+
+Table 2 fallacy + @KeeleStevensonElwert:2020 thing
+
+Post-treatment control bias - cite that one paper by Brendan Nyhan - DAGs make it obvious which nodes are post-treatment, since they appear after the treatment node in the causal chain in the graph
+
+# Identifying and isolating causal relationships
 
@@
 
-# A vocabulary for observational causal inference (ANDREW)
+## Causal identification and *d*-separation
 
-In the absence of random assignment, finding causal effects in observational data where units of analysis can self select their treatments and outcomes thus appears impossible.
-
-Goal of much research in recent years has been to create methods that allow researchers to make plausible causal arguments using observed, non-experimental data.
-Judea Pearl, computer science, and epidemiology provide us with a special framework for observational causal inference.
-We introduce the vocabulary and grammar of this framework below.
-
-## Causal diagrams
-
-[@PearlMackenzie:2018; @MorganWinship:2014; @PearlGlymourJewell:2016]
-
-DAGs encode our understanding of the data generating process - what in the natural and social world leads to the treatment/exposure, the outcome, and both simultaneously.
-It represents a type of philosophical model of how the observed world works.
-Importantly, causal graphs also encode our assumptions of how the world works.
-
-- Nodes - nodes can be unmeasurable, or even unobserved
-- Edges - Arrows indicate a relationship, or the passing of *statistical association* between nodes - intervening on one node leads to changes in another node - When two nodes are connected by an arrow, we are stating that there is an assumed causal relationship between those nodes; when there is no arrow between nodes, we are explicitly stating that there is no relationship between the two.
-- Acyclicity - can't get back to a node - represents flow of time - if things are cyclical, like hiring new staff $\leftrightarrow$ increased capacity, you can make these nodes time-based: hiring new staff_t → increased capacity_t → hiring new staff_t+1 → increased capacity_t+1
-
-```{r}
-#| label: fig-dag-associations
-#| fig-width: 9
-#| fig-height: 2.75
-#| fig-cap: "Three types of relationships in DAGs"
-#| include: true
-
-data_confounder <- dagify(
-  Y ~ Z + X,
-  X ~ Z,
-  coords = list(x = c(X = 1, Y = 3, Z = 2),
-                y = c(X = 1, Y = 1, Z = 2))
-) |> 
-  tidy_dagitty() |> 
-  mutate(arrow_color = ifelse(name == "Z", clrs$orange, "grey80")) |> 
-  mutate(node_color = ifelse(name == "Z", TRUE, FALSE))
-
-confounder <- ggplot(data_confounder, aes(x = x, y = y, xend = xend, yend = yend)) +
-  geom_dag_edges(aes(edge_colour = arrow_color), edge_width = 1) +
-  geom_dag_point(aes(color = node_color), size = 12) +
-  geom_dag_text(data = filter(data_confounder, name != "Z"), size = 4) +
-  geom_dag_text(data = filter(data_confounder, name == "Z"), color = "white", size = 4) +
-  scale_color_manual(values = c("grey80", clrs$orange), guide = "none") +
-  coord_cartesian(ylim = c(0.95, 2.05)) +
-  labs(title = "Confounder", subtitle = "(Fork)") +
-  theme_np_dag() +
-  theme(plot.title = element_text(hjust = 0.5, color = clrs$orange),
-        plot.subtitle = element_text(hjust = 0.5, color = clrs$orange))
-
-data_mediator <- dagify(
-  Y ~ X + Z,
-  Z ~ X,
-  coords = list(x = c(X = 1, Y = 3, Z = 2),
-                y = c(X = 1, Y = 1, Z = 2))
-) |> 
-  tidy_dagitty() |> 
-  mutate(arrow_color = case_when(
-    name == "Z" & to == "Y" ~ clrs$purple, 
-    name == "X" & to == "Z" ~ clrs$purple,
-    TRUE ~ "grey80")) |> 
-  mutate(node_color = ifelse(name == "Z", TRUE, FALSE))
-
-mediator <- ggplot(data_mediator, aes(x = x, y = y, xend = xend, yend = yend)) +
-  geom_dag_edges(aes(edge_colour = arrow_color), edge_width = 1) +
-  geom_dag_point(aes(color = node_color), size = 12) +
-  geom_dag_text(data = filter(data_mediator, name != "Z"), size = 4) +
-  geom_dag_text(data = filter(data_mediator, name == "Z"), color = "white", size = 4) +
-  scale_color_manual(values = c("grey80", clrs$purple), guide = "none") +
-  coord_cartesian(ylim = c(0.95, 2.05)) +
-  labs(title = "Mediator", subtitle = "(Chain)") +
-  theme_np_dag() +
-  theme(plot.title = element_text(hjust = 0.5, color = clrs$purple),
-        plot.subtitle = element_text(hjust = 0.5, color = clrs$purple))
-
-data_collider <- dagify(
-  Y ~ X,
-  Z ~ X + Y,
-  coords = list(x = c(X = 1, Y = 3, Z = 2),
-                y = c(X = 1, Y = 1, Z = 2))
-) |> 
-  tidy_dagitty() |> 
-  mutate(arrow_color = ifelse(to == "Z", clrs$red, "grey80")) |> 
-  mutate(node_color = ifelse(name == "Z", TRUE, FALSE))
-
-collider <- ggplot(data_collider, aes(x = x, y = y, xend = xend, yend = yend)) +
-  geom_dag_edges(aes(edge_colour = arrow_color), edge_width = 1) +
-  geom_dag_point(aes(color = node_color), size = 12) +
-  geom_dag_text(data = filter(data_collider, name != "Z"), size = 4) +
-  geom_dag_text(data = filter(data_collider, name == "Z"), color = "white", size = 4) +
-  scale_color_manual(values = c("grey80", clrs$red), guide = "none") +
-  coord_cartesian(ylim = c(0.95, 2.05)) +
-  labs(title = "Collider", subtitle = "(Inverted fork)") +
-  theme_np_dag() +
-  theme(plot.title = element_text(hjust = 0.5, color = clrs$red),
-        plot.subtitle = element_text(hjust = 0.5, color = clrs$red))
-
-combined_associations <- (confounder | plot_spacer() | mediator | plot_spacer() | collider) +
-  plot_layout(widths = c(0.32, 0.02, 0.32, 0.02, 0.32))
-combined_associations
-```
-
-This stuff in @fig-dag-associations: (@Elwert:2013 for forks, chains, inverted forks terminology)
-
-- Confounding - forks
-- Mediation - chains
-- Collision - inverted forks [@ElwertWinship:2014; @KnoxLoweMummolo:2020]
-
-Confounding is a major scary issue though - especially if it's unmeasured or unobservable…
-
-## Causal identification
+To identify a causal relationship in a DAG, we have to ensure that the arrow between the treatment and outcome nodes is isolated and that the treatment and outcome nodes are not linked through any other pathways
 
@@
 
-## Bad controls
-
-In addition to identification, DAGs provide additional statistical insight and guidance regarding control variables or covariates in regression models
-
-Bad controls and colliders
-
-Good/bad controls: <https://papers.ssrn.com/sol3/papers.cfm?abstract_id=3689437> - <https://twitter.com/analisereal/status/1512596580632707078>
-
-Table 2 fallacy + @KeeleStevensonElwert:2020 thing
-
-Post-treatment control bias - cite that one paper by Brendan Nyhan - DAGs make it obvious which nodes are post-treatment, since they appear after the treatment node in the causal chain in the graph
-
-# Adjustment-based identification (MENG AND ANDREW)
-
-TODO: Use simulated data based on a DAG that has a mediator, collider, and a bunch of confounders
-
-TODO: Replicate a study + draw a DAG for it + use IPW to close back doors
-
-## Backdoor adjustment
+---
 
@@
 
-There are many different ways—regression, matching, IPW—each with different nuances.
+There are two general approaches for doing this:
 
-### Regression
+- Circumstantial identification
+- Adjustment-based identification
 
-### Matching
-
-Matching creates entirely new treatment/control populations
-
-### IPW
-
-IPW creates comparable pseudo populations
-
-MAYBE: Make an image with little shaded people showing how matching and IPW work, similar to Torres:2020 (maybe with https://github.com/propublica/weepeople ?)
-
-## Illustration
-
-### Simulated example
-
-### Real example
-
-## Frontdoor adjustment
-
-A less common approach to removing confounding is frontdoor adjustment
-
-## Other methods?
-
-Briefly mention other methods, like marginal structural models [@BlackwellGlynn:2018]?
-
-# Design-based identification (ANDREW AND MENG)
+## Circumstantial identification
 
@@
 The language of causal graphs, identification, and adjustment provide a universal grammar for discussing causal effects.
-Commonly used approaches in econometrics and other social science disciplines can be written as causal graphs (see @fig-design-based-dags))
+Commonly used approaches in econometrics and other social science disciplines can be written as causal graphs [see @fig-design-based-dags]
 
@@
   coord_cartesian(ylim = c(0.95, 2.05)) +
-  labs(title = "Randomized trial", 
+  labs(title = "Randomization", 
        subtitle = "Randomization deletes all arrows into X") +
@@
 
-For each of these:
-
-1. Sort explanation of what they require, how they work
-2. Explanation of how that connects to the DAG and how the graph reveals the identification strategy
-3. Point to some nonprofit-related papers that use these
-
-Quasi-experiments (design-based—special situations)
-
-- Diff-in-diff
-- RDD
-- IV
-
-## Experiments
+### RCTs
 
@@
 
-## Diff-in-diff
+- Explanation + DAG
+- Illustration of how to use it
+- Review of existing nonprofit studies that use it - survey vignette experiments, conjoint experiments, field experiments, other kinds of RCTs
+- Possible nonprofit research questions that could use it
+
+### Diff-in-diff + TWFE
 
@@
 
-## Regression discontinuity
+- Explanation + DAG
+- Illustration of how to use it
+- Review of existing nonprofit studies that use it
+- Possible nonprofit research questions that could use it
+
+### RDD
 
@@
 
-## Instrumental variables
+- Explanation + DAG
+- Illustration of how to use it
+- Review of existing nonprofit studies that use it
+- Possible nonprofit research questions that could use it
+
+### IV
 
@@
 
-# Tools for researchers and practitioners (BOTH)
+These used to be common, but have become less popular because of the difficulty in finding a valid instrument that meets all the criteria for clean identification.
+@rainfall-instrument-paper.
+Randomized promotion works well, though - @worldbank-book
+
+- Explanation + DAG
+- Illustration of how to use it
+- Review of existing nonprofit studies that use it
+- Possible nonprofit research questions that could use it
+
+## Adjustment-based identification
+
+### Regression adjustment
+
+- Explanation + DAG
+- Illustration of how to use it
+- Review of existing nonprofit studies that use it
+- Possible nonprofit research questions that could use it
+
+### IPW and g-computation
+
+- Explanation + DAG
+- Illustration of how to use it
+- Review of existing nonprofit studies that use it
+- Possible nonprofit research questions that could use it
+
+Matching creates entirely new treatment/control populations
+
+IPW creates comparable pseudo populations
+
+MAYBE: Make an image with little shaded people showing how matching and IPW work, similar to Torres:2020 (maybe with https://github.com/propublica/weepeople ?)
+
+---
+
+Despite its popularity in epidemiology and public health, to our knowledge, there are no studies in NVSQ, Voluntas, or NML that employ inverse probability weighting for covariate adjustment for causal inference.
+There are X that use matching, and there are many that use regression-based adjustment in explicitly non-causal ways (but Haber et al paper finds that these things imply causation anyway).
+For instance, Huafang et al.
+paper explicity talks about associations after controlling for ostensible confounders, but also uses causal language like "program impacts" - C-word article and not shying away from causation
+
+## Hybrid approaches
+
+Briefly describe combinations of the two approaches, like synthetic controls, DiD with matching, RDD with covariate adjustment, etc.
+
+Briefly mention other methods, like marginal structural models [@BlackwellGlynn:2018]?
+
+# Tools for researchers and practitioners
 
@@
 
-## Dealing with time
-
-- Epidemiology and polsci and marginal structural models
-- Economics and TWFE and synthetic controls and event studies - fancier diff-in-diff
-
-# Conclusion (BOTH)
+# Conclusion
 
 A call for more causation / more careful thinking around causation. Prediction is fine. But don't automatically run away from causal work.
+
+We believe that adjustment-based approaches like regression adjustment, matching, and IPW are highly amenable to nonprofit administrative data.
+Finding specific situations like diff-in-diff is popular in public policy analysis and econometrics because the scale of the data is much larger—researchers can analyze the effect of state- or national-level policies in difference-in-difference designs, or observe the behavior of millions of participants in programs like Medicaid or the Affordable Care Act for regression discontinuity designs.
+A smaller nonprofit interested in evaluating the impact of a new community program is most likely unable to consider these types of situation-specific quasi-experimental designs to isolate the causal effect.
+However, with careful theoretical thinking; the development of a robust DAG that describes potential confounders, mediators, and colliders; the use of high quality administrative data; and intentional sensitivity analysis that probes the strength of possible causal relationships when faced with unmeasured confounding [@that-sensmakr-paper; @lucys-tipr-paper]; nonprofit researchers can tell plausible causal stories.
+
+# References

### 2024-11-21 R/funs_data.R
Simulate observational and RCT DAGs

@@ -1,0 +1,18 @@
+# When using a file-based target, {targets} requires that the function that
+# saves the file returns a path to the file. write_csv() and write_dta() both
+# invisibly return the data frame being written, and saveRDS() returns NULL, so
+# we need some wrapper functions to save the files and return the paths.
+save_csv <- function(df, path) {
+  readr::write_csv(df, path)
+  return(path)
+}
+
+save_r <- function(df, path) {
+  saveRDS(df, path)
+  return(path)
+}
+
+save_dta <- function(df, path) {
+  haven::write_dta(df, path)
+  return(path)
+}

### 2024-11-21 R/generate-observational.R

@@ -1,0 +1,68 @@
+make_observational <- function(n, seed) {
+  withr::with_seed(seed, {
+    simulated_data_raw <- tibble(
+      id = as.integer(1:n + 999)
+    ) |>
+
+    # Generate confounders
+    mutate(
+      social_awareness = 100 * rbeta(n, shape1 = 7, shape2 = 3),
+      age = rnorm(n, mean = 25, sd = 2.5),
+      gpa = pmin(pmax(rnorm(n, mean = 3.25, sd = 0.6), 2.0), 4.0),
+      income = pmax(rnorm(n, mean = 40000, sd = 15000), 15000)
+    ) |>
+
+    # Generate treatment variable
+    mutate(
+      treatment_logit = 
+        -3 +  # Logit intercept
+        0.04 * social_awareness +
+        0.15 * (age - 25) +
+        0.8 * (gpa - 3.25) +
+        0.000032 * (income - 40000),
+      treatment_prob = plogis(treatment_logit),
+      treatment = rbinom(n, 1, treatment_prob)
+    ) |>
+
+    # Generate mediator
+    mutate(community_connections = rpois(n, lambda = 3 + 2.75 * treatment)) |> 
+
+    # Generate outcome
+    mutate(
+      prosocial_base = 
+        50 +  # OLS intercept
+        0.45 * social_awareness +
+        0.6 * (age - 25) +
+        3.3 * (gpa - 3.25) +
+        0.0002 * (income - 40000) +
+        9.2 * treatment +
+        1.3 * community_connections,  # Mediator
+      prosocial_index = pmin(pmax(prosocial_base + rnorm(n, 0, 6), 0), 100)
+    ) |> 
+
+    # Generate collider
+    mutate(
+      employed_logit = 
+        -23 +
+        1.3 * treatment +
+        0.25 * prosocial_index,
+      employed_prob = plogis(employed_logit),
+      employed_nonprofit = rbinom(n, 1, employed_prob)
+    )
+  })
+
+  simulated_data <- simulated_data_raw |> 
+    mutate(
+      across(c(prosocial_index, social_awareness, age), 
+      \(x) as.integer(round(x, 0)))
+    ) |> 
+    mutate(
+      gpa = round(gpa, 2),
+      income = as.integer(signif(income, 3))
+    ) |> 
+    select(
+      id, experiential_learning = treatment, prosocial_index, 
+      social_awareness, age, gpa, income, 
+      community_connections, employed_nonprofit
+    )
+}

### 2024-11-21 R/generate-rct.R

@@ -1,0 +1,60 @@
+make_rct <- function(n, seed) {
+  withr::with_seed(seed, {
+    simulated_data_raw <- tibble(
+      id = as.integer(1:n + 999)
+    ) |>
+
+    # Generate confounders
+    mutate(
+      social_awareness = 100 * rbeta(n, shape1 = 7, shape2 = 3),
+      age = rnorm(n, mean = 25, sd = 2.5),
+      gpa = pmin(pmax(rnorm(n, mean = 3.25, sd = 0.6), 2.0), 4.0),
+      income = pmax(rnorm(n, mean = 40000, sd = 15000), 15000)
+    ) |>
+
+    # Generate treatment variable
+    # lol this is so easy compared to the full observational DAG
+    mutate(treatment = rbinom(n, 1, 0.5)) |> 
+
+    # Generate mediator
+    mutate(community_connections = rpois(n, lambda = 3 + 2.75 * treatment)) |> 
+
+    # Generate outcome
+    mutate(
+      prosocial_base = 
+        50 +  # OLS intercept
+        0.45 * social_awareness +
+        0.6 * (age - 25) +
+        3.3 * (gpa - 3.25) +
+        0.0002 * (income - 40000) +
+        9.7 * treatment +
+        1.3 * community_connections,  # Mediator
+      prosocial_index = pmin(pmax(prosocial_base + rnorm(n, 0, 6), 0), 100)
+    ) |> 
+
+    # Generate collider
+    mutate(
+      employed_logit = 
+        -23 +
+        1.3 * treatment +
+        0.25 * prosocial_index,
+      employed_prob = plogis(employed_logit),
+      employed_nonprofit = rbinom(n, 1, employed_prob)
+    )
+  })
+
+  simulated_data <- simulated_data_raw |> 
+    mutate(
+      across(c(prosocial_index, social_awareness, age), 
+      \(x) as.integer(round(x, 0)))
+    ) |> 
+    mutate(
+      gpa = round(gpa, 2),
+      income = as.integer(signif(income, 3))
+    ) |> 
+    select(
+      id, experiential_learning = treatment, prosocial_index, 
+      social_awareness, age, gpa, income, 
+      community_connections, employed_nonprofit
+    )
+}

### 2024-11-21 _targets.R

@@ -50,10 +50,45 @@ list(
     clrs, get_latex_rgb, print.color_table)
   ),
 
-  ## Manuscript targets ----
+  ## Data generation ----
+  tar_target(data_observational, make_observational(n = 1135, seed = 536406)),
+  tar_target(data_rct, make_rct(n = 1328, seed = 180295)),
 
+  ## Save data -----
+  tar_target(
+    data_observational_rds, 
+    save_r(data_observational, here_rel("data", "simulated_data", "observational.rds")),
+    format = "file"
+  ),
+  tar_target(
+    data_observational_csv, 
+    save_csv(data_observational, here_rel("data", "simulated_data", "observational.csv")),
+    format = "file"
+  ),
+  tar_target(
+    data_observational_dta, 
+    save_dta(data_observational, here_rel("data", "simulated_data", "observational.dta")),
+    format = "file"
+  ),
+  tar_target(
+    data_rct_rds, 
+    save_r(data_rct, here_rel("data", "simulated_data", "rct.rds")),
+    format = "file"
+  ),
+  tar_target(
+    data_rct_csv, 
+    save_csv(data_rct, here_rel("data", "simulated_data", "rct.csv")),
+    format = "file"
+  ),
+  tar_target(
+    data_rct_dta, 
+    save_dta(data_rct, here_rel("data", "simulated_data", "rct.dta")),
+    format = "file"
+  ),
+
+  ## Manuscript targets ----
   tar_quarto(manuscript, path = "manuscript", working_directory = "manuscript", quiet = FALSE),
-  
+
   ## Render the README ----
   tar_quarto(readme, here_rel("README.qmd"))
 )

### 2024-11-21 R/funs_graphics.R
size → linewidth

@@ -59,7 +59,7 @@ theme_np <- function(base_size = 11, base_family = "Assistant") {
                                       margin = margin(t = 10), hjust = 0),
           plot.tag = element_text(size = rel(1.4), face = "bold",
                                   family = "Assistant"),
-          panel.border = element_rect(color = "grey50", fill = NA, size = 0.15),
+          panel.border = element_rect(color = "grey50", fill = NA, linewidth = 0.15),
           panel.spacing = unit(1, "lines"),
           panel.grid.minor = element_blank(),
           strip.text = element_text(size = rel(0.9), hjust = 0,

### 2024-11-21 R/funs_graphics.R
Better colors

@@ -109,9 +109,15 @@ theme_np_dag <- function(...) {
 
 # Colors ------------------------------------------------------------------
 
-clrs <- viridisLite::turbo(6, begin = 0, end = 1) |> 
-  purrr::set_names(c("purple", "blue", "green", "yellow", "orange", "red")) |> 
-  as.list()
+# clrs <- viridisLite::turbo(6, begin = 0, end = 1) |> 
+#   purrr::set_names(c("purple", "blue", "green", "yellow", "orange", "red")) |> 
+#   as.list()
+
+# https://carto.com/carto-colors/
+clrs <- list(
+  Prism = rcartocolor::carto_pal(n = 12, "Prism"),
+  Sunset = rcartocolor::carto_pal(7, "Sunset")
+)
 
 #' Convert colors to LaTeX RGB definitions
 #'

### 2024-11-21 methods/observational.qmd
Analyze RCT and observational simulations

---
title: "Observational"
---

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  fig.width = 6,
  fig.height = 6 * 0.618,
  fig.retina = 3,
  dev = "ragg_png",
  fig.align = "center",
  out.width = "95%",
  warning = FALSE,
  collapse = TRUE,
  cache.extra = 1234  # Change number to invalidate cache
)

options(
  digits = 3, width = 120,
  tinytable_tt_digits = 2
)
```

```{r}
#| label: packages-data
#| warning: false
#| message: false

library(tidyverse)
library(targets)
library(modelsummary)
library(tinytable)
library(marginaleffects)
library(ggdist)
library(scales)
library(WeightIt)
library(ggdag)
library(dagitty)
library(parameters)

tar_config_set(
  store = here::here('_targets'),
  script = here::here('_targets.R')
)

observed <- tar_read(data_observational)

# Plotting functions
invisible(list2env(tar_read(graphic_functions), .GlobalEnv))
set_annotation_fonts()

gof_map <- tribble(
  ~raw,            ~clean,               ~fmt, ~omit,
  "nobs",          "N",                  0,    FALSE,
  "r.squared",     "$R^2$",          2,    FALSE,
  "adj.r.squared", "\\(R^2\\) adjusted", 2,    FALSE
)
```

## DAG

```{r}
#| label: plot-dag-observational
#| fig-width: 6
#| fig-height: 6

coords <- tribble(
  ~name, ~x,   ~y,   ~type,         ~color,         ~label,
  "X",   1,    3,    "Treatment",   clrs$Prism[4],  "Experiential\nphilanthropy",
  "Y",   3,    3,    "Outcome",     clrs$Prism[2],  "Charitable\nbehavior",
  "Z1",  5/3,  4.2,  "Confounder",  clrs$Prism[6],  "Social awareness",
  "Z2",  7/3,  4.2,  "Confounder",  clrs$Prism[6],  "Age",
  "Z3",  5/3,  5,    "Confounder",  clrs$Prism[6],  "GPA",
  "Z4",  7/3,  5,    "Confounder",  clrs$Prism[6],  "Income",
  "M",   2,    2,    "Mediator",    clrs$Prism[3],  "Community\nconnections",
  "C",   2,    1,    "Collider",    clrs$Prism[8],  "Employment in\nnonprofit sector"
) |> 
  mutate(type = fct_inorder(type))

data_full_dag <- dagify(
  Y ~ X + Z1 + Z2 + Z3 + Z4 + M,
  X ~ Z1 + Z2 + Z3 + Z4,
  M ~ X,
  C ~ X + Y,
  coords = coords,
  labels = pull(coords, label, name = name)
) |> 
  tidy_dagitty() |> 
  left_join(select(coords, name, type, color), by = join_by(name)) |> 
  mutate(arrow_color = case_when(
    name == "X" & to == "Y" ~ "black",
    to == "C" ~ clrs$Prism[8],
    to == "M" | (name == "M" & to == "Y") ~ clrs$Prism[3],
    .default = color
  ))

ggplot(data_full_dag, aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_edges(edge_width = 0.5) +
  geom_dag_edges(aes(edge_color = arrow_color), edge_width = 0.5) +
  geom_dag_point(aes(color = type), size = 12) +
  geom_dag_label(
    data = filter(data_full_dag, type != "Confounder"), aes(label = label), 
    size = 4, nudge_y = -0.46, color = "black", lineheight = 1
  ) +
  geom_dag_label(
    data = filter(data_full_dag, type == "Confounder"), aes(label = label), 
    size = 4, nudge_y = 0.35, color = "black", lineheight = 1
  ) +
  scale_color_manual(
    values = coords |> distinct(type, color) |> pull(color, name = type),
    guide = guide_legend(
      title = NULL, override.aes = list(size = 6), label.position = "bottom"
    )
  ) +
  coord_cartesian(xlim = c(0.98, 3.02)) +
  theme_np_dag() + 
  theme(legend.position = "top")
```

## EDA

### Confounders

```{r}
ggplot(observed, aes(x = social_awareness, y = experiential_learning)) +
  geom_dots(aes(side = ifelse(experiential_learning == 1, "bottom", "top")),
    pch = 19, color = "grey20", scale = 0.2
  ) +
  geom_smooth(
    method = "glm", method.args = list(family = binomial(link = "logit")),
    se = FALSE
  ) +
  scale_y_continuous(labels = label_percent()) +
  labs(x = "Social awareness index", y = "Experiential learning participation")

ggplot(observed, aes(x = age, y = experiential_learning)) +
  geom_dots(aes(side = ifelse(experiential_learning == 1, "bottom", "top")),
    pch = 19, color = "grey20", scale = 0.4
  ) +
  geom_smooth(
    method = "glm", method.args = list(family = binomial(link = "logit")),
    se = FALSE
  ) +
  scale_y_continuous(labels = label_percent()) +
  labs(x = "Age", y = "Experiential learning participation")

ggplot(observed, aes(x = gpa, y = experiential_learning)) +
  geom_dots(aes(side = ifelse(experiential_learning == 1, "bottom", "top")),
    pch = 19, color = "grey20", scale = 0.4
  ) +
  geom_smooth(
    method = "glm", method.args = list(family = binomial(link = "logit")),
    se = FALSE
  ) +
  scale_y_continuous(labels = label_percent()) +
  labs(x = "GPA", y = "Experiential learning participation")

ggplot(observed, aes(x = income, y = experiential_learning)) +
  geom_dots(aes(side = ifelse(experiential_learning == 1, "bottom", "top")),
    pch = 19, color = "grey20", scale = 0.4
  ) +
  geom_smooth(
    method = "glm", method.args = list(family = binomial(link = "logit")),
    se = FALSE
  ) +
  scale_x_continuous(labels = label_dollar()) +
  scale_y_continuous(labels = label_percent()) +
  labs(x = "Income", y = "Experiential learning participation")
```

```{r}
ggplot(observed, aes(x = social_awareness, y = prosocial_index)) +
  geom_point() +
  geom_smooth()

ggplot(observed, aes(x = age, y = prosocial_index)) +
  geom_point() +
  geom_smooth()

ggplot(observed, aes(x = gpa, y = prosocial_index)) +
  geom_point() +
  geom_smooth()

ggplot(observed, aes(x = income, y = prosocial_index)) +
  geom_point() +
  geom_smooth()
```

### Mediator

```{r}
observed |> 
  mutate(experiential_learning_fct = as.logical(experiential_learning)) |> 
  ggplot(aes(x = experiential_learning_fct, y = community_connections)) +
  geom_point(position = position_jitter(height = 0)) +
  geom_boxplot(aes(group = experiential_learning_fct))

ggplot(observed, aes(x = community_connections, y = prosocial_index)) +
  geom_point() +
  labs(x = "Community connections", y = "Prosocial index")
```

### Collider

```{r}
observed |> 
  mutate(experiential_learning_fct = as.logical(experiential_learning)) |> 
  mutate(employed_nonprofit_fct = as.logical(employed_nonprofit)) |> 
  ggplot(aes(x = experiential_learning_fct, y = employed_nonprofit_fct)) +
  geom_point(position = position_jitter())

observed |> 
  mutate(employed_nonprofit_fct = as.logical(employed_nonprofit)) |> 
  ggplot(aes(x = prosocial_index, y = employed_nonprofit_fct)) +
  geom_point(position = position_jitter(width = 0))
```

## Model comparisons

```{r}
model_naive <- lm(
  prosocial_index ~ experiential_learning, 
  data = observed
)

model_adjusted_correct <- lm(
  prosocial_index ~ experiential_learning + social_awareness + age + gpa + income,
  data = observed
)

model_with_mediator <- lm(
  prosocial_index ~ experiential_learning + social_awareness + age + gpa + income + community_connections, 
  data = observed
)

model_collider_selection <- lm(
  prosocial_index ~ experiential_learning + social_awareness + age + gpa + income, 
  data = filter(observed, employed_nonprofit == 1)
)

model_collider_post_treatment <- lm(
  prosocial_index ~ experiential_learning + social_awareness + age + gpa + income + employed_nonprofit, 
  data = observed
)

models <- list(
  "Correctly adjusted" = model_adjusted_correct,
  "Naive" = model_naive,
  "Mediator included" = model_with_mediator,
  "Collider (selection)" = model_collider_selection,
  "Collider (post-treatment covariate)" = model_collider_post_treatment
)
```

```{r}
modelsummary(
  models,
  statistic = "[{conf.low}, {conf.high}]",
  fmt = 2,
  gof_map = gof_map,
  add_columns = tibble("True effect" = c(NA, NA, 10)) |> magrittr::set_attr("position", 2)
) |> 
  style_tt(
    i = 3:4, j = 1:6, 
    background = clrs$Prism[6], color = "#ffffff", bold = TRUE
  )
```

```{r}
modelsummary(
  models,
  statistic = "[{conf.low}, {conf.high}]",
  fmt = 2,
  coef_map = c("experiential_learning"),
  gof_map = gof_map,
  add_columns = tibble("True effect" = 10) |> magrittr::set_attr("position", 2)
)
```

```{r}
treatment_effects <- enframe(models) |> 
  mutate(params = map(value, \(x) model_parameters(x))) |> 
  unnest(params) |> 
  filter(Parameter == "experiential_learning") |> 
  mutate(name = factor(name, levels = names(models)))

ggplot(treatment_effects, aes(x = Coefficient, y = fct_rev(name))) +
  geom_vline(xintercept = 10, color = clrs$Prism[2]) +
  geom_pointrange(aes(xmin = CI_low, xmax = CI_high)) +
  labs(x = "Treatment effect", y = NULL) +
  theme_np() +
  theme(panel.grid.major.y = element_blank())
```

## Fancier estimation

```{r}
weights_ate <- weightit(
  experiential_learning ~ social_awareness + age + gpa + income, 
  data = observed, 
  method = "glm", 
  estimand = "ATE"
)

observed$wts_ate <- weights_ate$weights

plot_data_weights_ate <- tibble(
  propensity = weights_ate$ps,
  weight = weights_ate$weights,
  treatment = weights_ate$treat
)

ggplot() + 
  geom_histogram(data = filter(plot_data_weights_ate, treatment == 1), 
    bins = 50, aes(x = propensity, weight = weight, fill = "Treated pseudo-population")) + 
  geom_histogram(data = filter(plot_data_weights_ate, treatment == 0), 
    bins = 50, aes(x = propensity, weight = weight, y = -after_stat(count), fill = "Untreated psuedo-population")) +
  geom_histogram(data = filter(plot_data_weights_ate, treatment == 1), 
    bins = 50, aes(x = propensity, fill = "Treated people")) + 
  geom_histogram(data = filter(plot_data_weights_ate, treatment == 0), 
    bins = 50, aes(x = propensity, y = -after_stat(count), fill = "Untreated people")) +
  geom_hline(yintercept = 0, color = "white", linewidth = 0.25) +
  scale_x_continuous(labels = scales::label_percent()) +
  scale_y_continuous(label = abs) +
  scale_fill_manual(
    values = c(
      clrs$Prism[2], colorspace::lighten(clrs$Prism[2], 0.5), 
      clrs$Prism[8], colorspace::lighten(clrs$Prism[8], 0.65)), 
    guide = guide_legend(reverse = FALSE, nrow = 2)) +
  labs(x = "Propensity", y = "Count", fill = NULL) +
  theme_np() +
  theme(
    panel.grid = element_blank(),
    legend.position = "top",
    legend.key.size = unit(0.65, "lines")
  )
```

```{r}
model_outcome_ate <- lm(
  prosocial_index ~ experiential_learning, 
  data = observed, 
  weights = wts_ate
)

avg_comparisons(model_outcome_ate, variables = "experiential_learning")
```

### 2024-11-21 methods/rct.qmd
Analyze RCT and observational simulations

---
title: "RCT"
---

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  fig.width = 6,
  fig.height = 6 * 0.618,
  fig.retina = 3,
  dev = "ragg_png",
  fig.align = "center",
  out.width = "95%",
  warning = FALSE,
  collapse = TRUE,
  cache.extra = 1234  # Change number to invalidate cache
)

options(
  digits = 3, width = 120,
  tinytable_tt_digits = 2
)
```

```{r}
#| label: packages-data
#| warning: false
#| message: false

library(tidyverse)
library(targets)
library(modelsummary)
library(tinytable)
library(ggdist)
library(scales)
library(ggdag)
library(dagitty)
library(parameters)

tar_config_set(
  store = here::here('_targets'),
  script = here::here('_targets.R')
)

rct <- tar_read(data_rct)

# Plotting functions
invisible(list2env(tar_read(graphic_functions), .GlobalEnv))
set_annotation_fonts()

gof_map <- tribble(
  ~raw,            ~clean,               ~fmt, ~omit,
  "nobs",          "N",                  0,    FALSE,
  "r.squared",     "$R^2$",          2,    FALSE,
  "adj.r.squared", "\\(R^2\\) adjusted", 2,    FALSE
)
```

## DAG

```{r}
#| label: plot-dag-rct
#| fig-width: 6
#| fig-height: 6

coords <- tribble(
  ~name, ~x,   ~y,   ~type,         ~color,         ~label,
  "X",   1,    3,    "Treatment",   clrs$Prism[4],  "Experiential\nphilanthropy",
  "Y",   3,    3,    "Outcome",     clrs$Prism[2],  "Charitable\nbehavior",
  "Z1",  5/3,  4.2,  "Confounder",  clrs$Prism[6],  "Social awareness",
  "Z2",  7/3,  4.2,  "Confounder",  clrs$Prism[6],  "Age",
  "Z3",  5/3,  5,    "Confounder",  clrs$Prism[6],  "GPA",
  "Z4",  7/3,  5,    "Confounder",  clrs$Prism[6],  "Income",
  "M",   2,    2,    "Mediator",    clrs$Prism[3],  "Community\nconnections",
  "C",   2,    1,    "Collider",    clrs$Prism[8],  "Employment in\nnonprofit sector"
) |> 
  mutate(type = fct_inorder(type))

data_full_dag <- dagify(
  Y ~ X + Z1 + Z2 + Z3 + Z4 + M,
  M ~ X,
  C ~ X + Y,
  coords = coords,
  labels = pull(coords, label, name = name)
) |> 
  tidy_dagitty() |> 
  left_join(select(coords, name, type, color), by = join_by(name)) |> 
  mutate(arrow_color = case_when(
    name == "X" & to == "Y" ~ "black",
    to == "C" ~ clrs$Prism[8],
    to == "M" | (name == "M" & to == "Y") ~ clrs$Prism[3],
    .default = color
  ))

ggplot(data_full_dag, aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_edges(edge_width = 0.5) +
  geom_dag_edges(aes(edge_color = arrow_color), edge_width = 0.5) +
  geom_dag_point(aes(color = type), size = 12) +
  geom_dag_label(
    data = filter(data_full_dag, type != "Confounder"), aes(label = label), 
    size = 4, nudge_y = -0.46, color = "black", lineheight = 1
  ) +
  geom_dag_label(
    data = filter(data_full_dag, type == "Confounder"), aes(label = label), 
    size = 4, nudge_y = 0.35, color = "black", lineheight = 1
  ) +
  scale_color_manual(
    values = coords |> distinct(type, color) |> pull(color, name = type),
    guide = guide_legend(
      title = NULL, override.aes = list(size = 6), label.position = "bottom"
    )
  ) +
  coord_cartesian(xlim = c(0.98, 3.02)) +
  theme_np_dag() + 
  theme(legend.position = "top")
```

## EDA

### Confounders

```{r}
ggplot(rct, aes(x = social_awareness, y = experiential_learning)) +
  geom_dots(aes(side = ifelse(experiential_learning == 1, "bottom", "top")),
    pch = 19, color = "grey20", scale = 0.2
  ) +
  geom_smooth(
    method = "glm", method.args = list(family = binomial(link = "logit")),
    se = FALSE
  ) +
  scale_y_continuous(labels = label_percent()) +
  labs(x = "Social awareness index", y = "Experiential learning participation")

ggplot(rct, aes(x = age, y = experiential_learning)) +
  geom_dots(aes(side = ifelse(experiential_learning == 1, "bottom", "top")),
    pch = 19, color = "grey20", scale = 0.4
  ) +
  geom_smooth(
    method = "glm", method.args = list(family = binomial(link = "logit")),
    se = FALSE
  ) +
  scale_y_continuous(labels = label_percent()) +
  labs(x = "Age", y = "Experiential learning participation")

ggplot(rct, aes(x = gpa, y = experiential_learning)) +
  geom_dots(aes(side = ifelse(experiential_learning == 1, "bottom", "top")),
    pch = 19, color = "grey20", scale = 0.4
  ) +
  geom_smooth(
    method = "glm", method.args = list(family = binomial(link = "logit")),
    se = FALSE
  ) +
  scale_y_continuous(labels = label_percent()) +
  labs(x = "GPA", y = "Experiential learning participation")

ggplot(rct, aes(x = income, y = experiential_learning)) +
  geom_dots(aes(side = ifelse(experiential_learning == 1, "bottom", "top")),
    pch = 19, color = "grey20", scale = 0.4
  ) +
  geom_smooth(
    method = "glm", method.args = list(family = binomial(link = "logit")),
    se = FALSE
  ) +
  scale_x_continuous(labels = label_dollar()) +
  scale_y_continuous(labels = label_percent()) +
  labs(x = "Income", y = "Experiential learning participation")
```

```{r}
ggplot(rct, aes(x = social_awareness, y = prosocial_index)) +
  geom_point() +
  geom_smooth()

ggplot(rct, aes(x = age, y = prosocial_index)) +
  geom_point() +
  geom_smooth()

ggplot(rct, aes(x = gpa, y = prosocial_index)) +
  geom_point() +
  geom_smooth()

ggplot(rct, aes(x = income, y = prosocial_index)) +
  geom_point() +
  geom_smooth()
```

### Mediator

```{r}
rct |> 
  mutate(experiential_learning_fct = as.logical(experiential_learning)) |> 
  ggplot(aes(x = experiential_learning_fct, y = community_connections)) +
  geom_point(position = position_jitter(height = 0)) +
  geom_boxplot(aes(group = experiential_learning_fct))

ggplot(rct, aes(x = community_connections, y = prosocial_index)) +
  geom_point() +
  labs(x = "Community connections", y = "Prosocial index")
```

### Collider

```{r}
rct |> 
  mutate(experiential_learning_fct = as.logical(experiential_learning)) |> 
  mutate(employed_nonprofit_fct = as.logical(employed_nonprofit)) |> 
  ggplot(aes(x = experiential_learning_fct, y = employed_nonprofit_fct)) +
  geom_point(position = position_jitter())

rct |> 
  mutate(employed_nonprofit_fct = as.logical(employed_nonprofit)) |> 
  ggplot(aes(x = prosocial_index, y = employed_nonprofit_fct)) +
  geom_point(position = position_jitter(width = 0))
```

## Model comparisons

```{r}
model_correct <- lm(
  prosocial_index ~ experiential_learning, 
  data = rct
)

model_adjusted <- lm(
  prosocial_index ~ experiential_learning + social_awareness + age + gpa + income,
  data = rct
)

model_with_mediator <- lm(
  prosocial_index ~ experiential_learning + social_awareness + age + gpa + income + community_connections, 
  data = rct
)

model_collider_selection <- lm(
  prosocial_index ~ experiential_learning + social_awareness + age + gpa + income, 
  data = filter(rct, employed_nonprofit == 1)
)

model_collider_post_treatment <- lm(
  prosocial_index ~ experiential_learning + social_awareness + age + gpa + income + employed_nonprofit, 
  data = rct
)

models <- list(
  "Treatment only" = model_correct,
  "Covariates included" = model_adjusted,
  "Mediator included" = model_with_mediator,
  "Collider (selection)" = model_collider_selection,
  "Collider (post-treatment covariate)" = model_collider_post_treatment
)
```

```{r}
modelsummary(
  models,
  statistic = "[{conf.low}, {conf.high}]",
  fmt = 2,
  gof_map = gof_map,
  add_columns = tibble("True effect" = c(NA, NA, 10)) |> magrittr::set_attr("position", 2)
) |> 
  style_tt(
    i = 3:4, j = 1:7, 
    background = clrs$Prism[6], color = "#ffffff", bold = TRUE
  )
```

```{r}
modelsummary(
  models,
  statistic = "[{conf.low}, {conf.high}]",
  fmt = 2,
  coef_map = c("experiential_learning"),
  gof_map = gof_map,
  add_columns = tibble("True effect" = 10) |> magrittr::set_attr("position", 2)
)
```

```{r}
treatment_effects <- enframe(models) |> 
  mutate(params = map(value, \(x) model_parameters(x))) |> 
  unnest(params) |> 
  filter(Parameter == "experiential_learning") |> 
  mutate(name = factor(name, levels = names(models)))

ggplot(treatment_effects, aes(x = Coefficient, y = fct_rev(name))) +
  geom_vline(xintercept = 10, color = clrs$Prism[2]) +
  geom_pointrange(aes(xmin = CI_low, xmax = CI_high)) +
  labs(x = "Treatment effect", y = NULL) +
  theme_np() +
  theme(panel.grid.major.y = element_blank())
```

### 2024-11-21 deploy.sh
Add deploy script

@@ -1,0 +1,7 @@
+REMOTE_HOST="ath-cloud"
+REMOTE_DIR="~/sites/stats.andrewheiss.com/public_html/corrupt-caterpillar"
+REMOTE_DEST=$REMOTE_HOST:$REMOTE_DIR
+
+echo "Uploading new changes to remote server..."
+echo
+rsync -crvP --exclude '*_cache' --delete _site/ $REMOTE_DEST

### 2024-11-21 causal-studies-over-time.qmd
Add NVSQ proportion chart

---
title: Causal work in nonnprofit studies
---

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  fig.width = 6,
  fig.height = 6 * 0.618,
  fig.retina = 3,
  dev = "ragg_png",
  fig.align = "center",
  out.width = "95%",
  warning = FALSE,
  collapse = TRUE,
  cache.extra = 1234  # Change number to invalidate cache
)

options(
  digits = 3, width = 120,
  tinytable_tt_digits = 2
)
```

```{r}
#| label: packages-data
#| warning: false
#| message: false

library(tidyverse)
library(targets)
library(here)

tar_config_set(
  store = here::here('_targets'),
  script = here::here('_targets.R')
)

invisible(list2env(tar_read(graphic_functions), .GlobalEnv))

articles <- read_csv(here("data", "manual_data", "Skimming notes_coding.csv"))
nvsq_counts <- read_csv(here("data", "manual_data", "nvsq_counts.csv"))
```

```{r}
#| fig-width: 6
#| fig-height: 4
nvsq_counts_year <- nvsq_counts |>
  filter(Year <= 2020) |>
  group_by(Year) |>
  summarize(total = sum(Count))

plot_data <- articles |>
  filter(year <= 2020) |>
  count(year) |>
  left_join(nvsq_counts_year, by = c("year" = "Year")) |>
  mutate(pct = n / total)

ggplot(plot_data, aes(x = factor(year), y = pct)) +
  geom_col(fill = clrs$Prism[9]) +
  labs(
    x = NULL,
    y = "Percent of NVSQ articles",
    title = "Proportion of causally-focused\narticles in NVSQ, 2010–2020",
    caption = "Article uses an experiment, difference-in-differences, regression discontinuity,\ninstrumental variables, or other method for statistically identifying a causal mechanism\n\n(Voluntas, NML, PAR, JPART, JPAM, and PMR forthcoming)"
  ) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  theme_np() +
  theme(panel.grid.major.x = element_blank())
```

### 2024-11-22 methods/diff-in-diff.qmd
Add diff-in-diff and RDD

---
title: "Difference-in-differences"
---

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  fig.width = 6,
  fig.height = 6 * 0.618,
  fig.retina = 3,
  dev = "ragg_png",
  fig.align = "center",
  out.width = "95%",
  warning = FALSE,
  collapse = TRUE,
  cache.extra = 1234  # Change number to invalidate cache
)

options(
  digits = 3, width = 120,
  tinytable_tt_digits = 2
)
```

```{r}
#| label: packages-data
#| warning: false
#| message: false

library(tidyverse)
library(targets)
library(modelsummary)
library(tinytable)
library(scales)
library(ggdag)
library(dagitty)
library(parameters)

tar_config_set(
  store = here::here('_targets'),
  script = here::here('_targets.R')
)

did <- tar_read(data_did)

invisible(list2env(tar_read(graphic_functions), .GlobalEnv))
set_annotation_fonts()

gof_map <- tribble(
  ~raw,            ~clean,               ~fmt, ~omit,
  "nobs",          "N",                  0,    FALSE,
  "r.squared",     "$R^2$",          2,    FALSE,
  "adj.r.squared", "\\(R^2\\) adjusted", 2,    FALSE
)
```

## Parallel trends

```{r}
#| message: false

did |> 
  group_by(school, year) |> 
  summarize(mean = ggplot2::mean_se(prosocial_index, mult = 1.96)) |> 
  unnest(mean) |> 
  ggplot(aes(x = year, y = y, color = school)) +
  geom_vline(xintercept = 2022.5, linetype = "dashed") +
  geom_line(linewidth = 0.25, show.legend = FALSE) +
  geom_pointrange(aes(ymin = ymin, ymax = ymax)) +
  scale_color_manual(values = c(clrs$Prism[2], clrs$Prism[8])) +
  labs(x = NULL, y = "Prosocial index", color = NULL) +
  theme_np()
```

## Models

```{r}
model_small <- lm(
  prosocial_index ~ school + factor(year) + school*factor(year),
  data = filter(did, year %in% c(2022, 2023))
)

model_small_controls <- lm(
  prosocial_index ~ school + factor(year) + school*factor(year) +
    social_awareness + age + gpa + income,
  data = filter(did, year %in% c(2022, 2023))
)

model_full <- lm(
  prosocial_index ~ school + factor(year) + I(school == "School B" & year >= 2023),
  data = did
)

model_full_controls <- lm(
  prosocial_index ~ school + factor(year) + I(school == "School B" & year >= 2023) +
    social_awareness + age + gpa + income,
  data = did
)

models <- list(
  "Interaction term only" = model_small,
  "Interaction term + covariates" = model_small_controls,
  "Interaction term only (all years)" = model_full,
  "Interaction term + covariates (all years)" = model_full_controls
)
```

```{r}
modelsummary(
  models,
  statistic = "[{conf.low}, {conf.high}]",
  fmt = 2,
  gof_map = gof_map,
  add_columns = tibble("True effect" = c(NA, NA, 10)) |> magrittr::set_attr("position", 2)
)# |> 
  # style_tt(
  #   i = 3:4, j = 1:6, 
  #   background = clrs$Prism[6], color = "#ffffff", bold = TRUE
  # )
```

```{r}
modelsummary(
  models,
  statistic = "[{conf.low}, {conf.high}]",
  fmt = 2,
  coef_map = c(
    `schoolSchool B:factor(year)2023` = "School B × 2023",
    `I(school == "School B" & year >= 2023)TRUE` = "School B × 2023"
  ),
  gof_map = gof_map,
  add_columns = tibble("True effect" = 10) |> magrittr::set_attr("position", 2)
)
```

```{r}
treatment_effects <- enframe(models) |> 
  mutate(params = map(value, \(x) model_parameters(x))) |> 
  unnest(params) |> 
  filter(Parameter %in% c("schoolSchool B:factor(year)2023", 'I(school == "School B" & year >= 2023)TRUE')) |> 
  mutate(name = factor(name, levels = names(models)))

ggplot(treatment_effects, aes(x = Coefficient, y = fct_rev(name))) +
  geom_vline(xintercept = 10, color = clrs$Prism[2]) +
  geom_pointrange(aes(xmin = CI_low, xmax = CI_high)) +
  labs(x = "Treatment effect", y = NULL) +
  theme_np() +
  theme(panel.grid.major.y = element_blank())
```

### 2024-11-22 methods/rdd-sharp.qmd
Add diff-in-diff and RDD

---
title: "Sharp RDD"
---

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  fig.width = 6,
  fig.height = 6 * 0.618,
  fig.retina = 3,
  dev = "ragg_png",
  fig.align = "center",
  out.width = "95%",
  warning = FALSE,
  collapse = TRUE,
  cache.extra = 1234  # Change number to invalidate cache
)

options(
  digits = 3, width = 120,
  tinytable_tt_digits = 2
)
```

```{r}
#| label: packages-data
#| warning: false
#| message: false

library(tidyverse)
library(targets)
library(modelsummary)
library(tinytable)
library(scales)
library(ggdag)
library(dagitty)
library(parameters)
library(rdrobust)
library(rddensity)

tar_config_set(
  store = here::here('_targets'),
  script = here::here('_targets.R')
)

rdd <- tar_read(data_rdd_sharp)

invisible(list2env(tar_read(graphic_functions), .GlobalEnv))
set_annotation_fonts()

gof_map <- tribble(
  ~raw,            ~clean,               ~fmt, ~omit,
  "nobs",          "N",                  0,    FALSE,
  "r.squared",     "$R^2$",          2,    FALSE,
  "adj.r.squared", "\\(R^2\\) adjusted", 2,    FALSE
)
```

```{r}
#| include: false
#' Tidy a `rdrobust` regression discontinuity design object
#'
#' @param x A `rdrobust` object
#' @param ... extra arguments (not used)
#' @examples
#' \dontrun{
#' x <- runif(1000,-1,1)
#' y <- 3 * x + 2 * (x>=0) + rnorm(1000)
#' mod <- rdrobust(y,x)
#' 
#' tidy(mod)
#' 
#' glance(mod)
#' }
#' @export
tidy.rdrobust <- function(object, ...){
    ret <- tibble(term = row.names(object$coef), 
                  estimate = object$coef[, 1], 
                  std.error = object$se[, 1], 
                  statistic = object$z[, 1],
                  p.value = object$pv[, 1], 
                  conf.low = object$ci[,1],
                  conf.high = object$ci[, 2])
    ret
}

#' Glance a `rdrobust` regression discontinuity design object
#'
#' @inheritParams tidy.rdrobust
#' @export
glance.rdrobust <- function(object, ...){
    ret <- tibble(nobs.left = object$N[1],
                  nobs.right = object$N[2],
                  nobs.effective.left = object$N_h[1],
                  nobs.effective.right = object$N_h[2],
                  cutoff = object$c,
                  order.regression = object$q,
                  order.bias = object$q,
                  kernel = object$kernel,
                  bwselect = object$bwselect)
    ret
}
```

```{r}
rdd |> 
  ggplot(aes(
    x = entrance_score, y = as.logical(experiential_learning), 
    color = factor(experiential_learning)
  )) +
  geom_point(
    size = 0.5, alpha = 0.5,
    position = position_jitter(width = 0, height = 0.25, seed = 1234)
  ) +
  geom_vline(xintercept = 75) +
  labs(
    x = "Entrance test score", 
    y = "Participated in\nexperiential learning program"
  ) +
  scale_color_manual(values = c(clrs$Prism[2], clrs$Prism[8]), guide = "none") +
  theme_np()
```

```{r}
ggplot(rdd, aes(x = entrance_score, fill = as.logical(experiential_learning))) +
  geom_histogram(binwidth = 1, color = "white", boundary = 0) +
  geom_vline(xintercept = 75) +
  labs(x = "Entrance exam score", y = "Count", fill = "In program") +
  scale_fill_manual(values = c(clrs$Prism[2], clrs$Prism[8])) +
  theme_np()
```

```{r}
test_density <- rddensity(rdd$entrance_score, c = 75)
# summary(test_density)

plot_density_test <- rdplotdensity(
  rdd = test_density,
  X = rdd$entrance_score,
  type = "both"
)
```

```{r}
ggplot(rdd, aes(x = entrance_score, y = prosocial_index, color = as.logical(experiential_learning))) +
  geom_point(size = 0.5, alpha = 0.5) +
  geom_smooth(data = filter(rdd, entrance_score <= 75, entrance_score >= 70),
              method = "lm", se = FALSE, linewidth = 2) +
  geom_smooth(data = filter(rdd, entrance_score > 75, entrance_score <= 80),
              method = "lm", se = FALSE, linewidth = 2) +
  geom_vline(xintercept = 75.5) +
  coord_cartesian(xlim = c(65, 85), ylim = c(60, 100)) +
  labs(x = "Entrance test score", y = "Prosocial index", color = "Participated") +
  scale_color_manual(values = c(clrs$Prism[2], clrs$Prism[8])) +
  theme_np()
```

```{r}
blah <- rdplot(y = rdd$prosocial_index, x = rdd$entrance_score, c = 75, hide = TRUE)
blah$rdplot +
  theme_np() +
  labs(title = NULL, x = "Entrance test score", y = "Prosocial index")
```

```{r}
#| warning: false
model_rdrobust <- rdrobust(y = rdd$prosocial_index, x = rdd$entrance_score, c = 75)

model_rdrobust |>
  summary()

tidy(model_rdrobust)
```

```{r}
model_bw5 <- lm(
  prosocial_index ~ entrance_score + experiential_learning,
  data = filter(rdd, entrance_score >= (75 - 5) & entrance_score <= (75 + 5))
)

model_bw10 <- lm(
  prosocial_index ~ entrance_score + experiential_learning,
  data = filter(rdd, entrance_score >= (75 - 10) & entrance_score <= (75 + 10))
)

models <- list(
  "BW = 5" = model_bw5,
  "BW = 10" = model_bw10,
  "rdrobust" = model_rdrobust
)

modelsummary(
  models,
  statistic = "[{conf.low}, {conf.high}]",
  fmt = 2,
  coef_map = c(
    `experiential_learning` = "Experiential learning",
    `Conventional` = "Experiential learning"
  ),
  gof_map = gof_map,
  add_columns = tibble("True effect" = 8.5) |> magrittr::set_attr("position", 2)
)
```

```{r}
treatment_effects <- enframe(models) |> 
  mutate(params = map(value, \(x) tidy(x, conf.int = TRUE))) |> 
  unnest(params) |> 
  filter(term %in% c("experiential_learning", "Conventional")) |> 
  mutate(name = factor(name, levels = names(models)))

ggplot(treatment_effects, aes(x = estimate, y = fct_rev(name))) +
  geom_vline(xintercept = 8.5, color = clrs$Prism[2]) +
  geom_pointrange(aes(xmin = conf.low, xmax = conf.high)) +
  labs(x = "Treatment effect", y = NULL) +
  theme_np() +
  theme(panel.grid.major.y = element_blank())
```

### 2024-11-22 R/generate-diff-in-diff.R

@@ -1,0 +1,86 @@
+make_diff_in_diff <- function(n_per_school_year, seed) {
+  withr::with_seed(seed, {
+    # Make a skeleton of school-year combinations
+    school_years <- expand_grid(
+      school = c("School A", "School B"),
+      # 3 pre-treatment years (2020-2022) and 2 post (2023-2024)
+      # We'll pretend the pandemic never happend
+      year = 2020:2024
+    ) |> 
+      # Add year-specific random offsets for each school
+      group_by(school, year) |>
+      mutate(year_effect = rnorm(1, mean = 0, sd = 0.2)) |> 
+      ungroup()
+
+    # Replicate each combination n times
+    base_data <- school_years |>
+      slice(rep(1:n(), each = n_per_school_year)) |>
+      mutate(id = row_number() + 999)
+
+    simulated_data <- base_data |>
+      # Create confounders that vary a little by school, for fun
+      mutate(
+        social_awareness = case_when(
+          school == "School A" ~ 100 * rbeta(n(), shape1 = 7, shape2 = 3),
+          school == "School B" ~ 100 * rbeta(n(), shape1 = 7.2, shape2 = 2.8)
+        ),
+        age = case_when(
+          school == "School A" ~ rnorm(n(), mean = 25, sd = 2.5),
+          school == "School B" ~ rnorm(n(), mean = 25.2, sd = 2.4)
+        ),
+        gpa = pmin(pmax(
+          case_when(
+            school == "School A" ~ rnorm(n(), mean = 3.25, sd = 0.6),
+            school == "School B" ~ rnorm(n(), mean = 3.28, sd = 0.58)
+          ), 2.0), 4.0
+        ),
+        income = pmax(
+          case_when(
+            school == "School A" ~ rnorm(n(), mean = 40000, sd = 15000),
+            school == "School B" ~ rnorm(n(), mean = 41000, sd = 14500)
+          ), 15000
+        )
+      ) |>
+
+      # Generate outcome
+      mutate(
+        prosocial_base = 
+          30 +  # Intercept
+          0.45 * social_awareness +
+          0.6 * (age - 25) +
+          3.3 * (gpa - 3.25) +
+          0.0002 * (income - 40000) +
+
+          # School fixed effect
+          case_when(
+            school == "School A" ~ 0,
+            school == "School B" ~ 2
+          ) +
+
+          # Time trend
+          1.5 * (year - 2020) +  # Base trend
[…]

### 2024-11-22 R/generate-observational.R

@@ -65,4 +65,6 @@ make_observational <- function(n, seed) {
       social_awareness, age, gpa, income, 
       community_connections, employed_nonprofit
     )
+
+  return(simulated_data)
 }

### 2024-11-22 R/generate-rct.R

@@ -57,4 +57,6 @@ make_rct <- function(n, seed) {
       social_awareness, age, gpa, income, 
       community_connections, employed_nonprofit
     )
+
+  return(simulated_data)
 }

### 2024-11-22 R/generate-rdd.R

@@ -1,0 +1,67 @@
+make_rdd <- function(n, seed, fuzzy = FALSE) {
+  withr::with_seed(seed, {
+    simulated_data_raw <- tibble(
+      id = as.integer(1:n + 999)
+    ) |>
+
+      # Generate confounders
+      mutate(
+        social_awareness = 100 * rbeta(n, shape1 = 7, shape2 = 3),
+        age = rnorm(n, mean = 25, sd = 2.5),
+        gpa = pmin(pmax(rnorm(n, mean = 3.25, sd = 0.6), 2.0), 4.0),
+        income = pmax(rnorm(n, mean = 40000, sd = 15000), 15000)
+      ) |>
+
+      # Generate entrance exam score (running variable)
+      mutate(
+        entrance_score_base = 
+          65 +  # Base score
+          0.15 * social_awareness +
+          0.3 * (age - 25) +
+          8 * (gpa - 3.25) +
+          0.0001 * (income - 40000),
+        entrance_score = pmin(pmax(
+          entrance_score_base + rnorm(n, mean = 0, sd = 8),
+          0), 100
+        )
+      ) |>
+
+      # Generate treatment assignment
+      mutate(
+        # For fuzzy RDD, use logistic function where probabilities shift at 75
+        # For sharp RDD, they're in if the score is ≥75
+        treatment_prob = case_when(
+          fuzzy ~ plogis(0 + 0.7 * (entrance_score - 75)),
+          !fuzzy ~ as.numeric(entrance_score >= 75)
+        ),
+        experiential_learning = rbinom(n, 1, treatment_prob)
+      ) |>
+
+      # Generate outcome
+      mutate(
+        prosocial_base = 
+          50 +  # Base level
+          0.45 * social_awareness +
+          0.6 * (age - 25) +
+          3.3 * (gpa - 3.25) +
+          0.0002 * (income - 40000) +
+          0.15 * (entrance_score - 75) +
+          8.5 * experiential_learning,  # Treatment effect!
+        prosocial_index = pmin(pmax(prosocial_base + rnorm(n, 0, 6), 0), 100)
+      )
+
+    simulated_data <- simulated_data_raw |>
+      mutate(
+        across(c(prosocial_index, social_awareness, age, entrance_score),
+               \(x) as.integer(round(x, 0))),
+        gpa = round(gpa, 2),
+        income = as.integer(signif(income, 3))
+      ) |>
+      select(
[…]

### 2024-11-22 _targets.R

@@ -53,6 +53,9 @@ list(
   ## Data generation ----
   tar_target(data_observational, make_observational(n = 1135, seed = 536406)),
   tar_target(data_rct, make_rct(n = 1328, seed = 180295)),
+  tar_target(data_did, make_diff_in_diff(n_per_school_year = 126, seed = 417724)),
+  tar_target(data_rdd_sharp, make_rdd(n = 967, seed = 635404, fuzzy = FALSE)),
+  tar_target(data_rdd_fuzzy, make_rdd(n = 992, seed = 995303, fuzzy = TRUE)),
 
   ## Save data -----
   tar_target(
@@ -83,6 +86,21 @@ list(
   tar_target(
     data_rct_dta, 
     save_dta(data_rct, here_rel("data", "simulated_data", "rct.dta")),
+    format = "file"
+  ),
+  tar_target(
+    data_did_rds, 
+    save_r(data_did, here_rel("data", "simulated_data", "did.rds")),
+    format = "file"
+  ),
+  tar_target(
+    data_did_csv, 
+    save_csv(data_did, here_rel("data", "simulated_data", "did.csv")),
+    format = "file"
+  ),
+  tar_target(
+    data_did_dta, 
+    save_dta(data_did, here_rel("data", "simulated_data", "did.dta")),
     format = "file"
   ),
 

### 2024-11-22 _targets.R
Add TODO

@@ -58,6 +58,7 @@ list(
   tar_target(data_rdd_fuzzy, make_rdd(n = 992, seed = 995303, fuzzy = TRUE)),
 
   ## Save data -----
+  # TODO: Use static/dynamic branching here instead of this gross repetitive mess
   tar_target(
     data_observational_rds, 
     save_r(data_observational, here_rel("data", "simulated_data", "observational.rds")),

### 2024-11-22 manuscript/manuscript.qmd
Update colors

@@
 invisible(list2env(tar_read(graphic_functions), .GlobalEnv))
+set_annotation_fonts()
 ```
@@
   ) +
-  scale_color_manual(values = c("X" = clrs$green, "Y" = clrs$blue, "Z" = clrs$yellow), guide = "none") +
+  scale_color_manual(values = c("X" = clrs$Prism[4], "Y" = clrs$Prism[2], "Z" = clrs$Prism[6]), guide = "none") +
   coord_cartesian(xlim = c(0.75, 3.25), ylim = c(0.6, 2.2)) +
@@
 ```
+
+DAGs force researchers to be explicit about their beliefs regarding the relationships between each variable and whether causal relationships do or do not exist (TODO UGH this is a gross sentence)
 
@@
   tidy_dagitty() |> 
-  mutate(arrow_color = ifelse(name == "Z", clrs$orange, "grey80")) |> 
+  mutate(arrow_color = ifelse(name == "Z", clrs$Prism[7], "grey80")) |> 
   mutate(node_color = ifelse(name == "Z", TRUE, FALSE))
@@
   geom_dag_text(data = filter(data_confounder, name == "Z"), color = "white", size = 4) +
-  scale_color_manual(values = c("grey80", clrs$orange), guide = "none") +
+  scale_color_manual(values = c("grey80", clrs$Prism[7]), guide = "none") +
   coord_cartesian(ylim = c(0.95, 2.05)) +
@@
   theme_np_dag() +
-  theme(plot.title = element_text(hjust = 0.5, color = clrs$orange))
+  theme(plot.title = element_text(hjust = 0.5, color = clrs$Prism[7]))
 
@@
   mutate(arrow_color = case_when(
-    name == "Z" & to == "Y" ~ clrs$purple, 
-    name == "X" & to == "Z" ~ clrs$purple,
+    name == "Z" & to == "Y" ~ clrs$Prism[1], 
+    name == "X" & to == "Z" ~ clrs$Prism[1],
     TRUE ~ "grey80")) |> 
@@
   geom_dag_text(data = filter(data_mediator, name == "Z"), color = "white", size = 4) +
-  scale_color_manual(values = c("grey80", clrs$purple), guide = "none") +
+  scale_color_manual(values = c("grey80", clrs$Prism[1]), guide = "none") +
   coord_cartesian(ylim = c(0.95, 2.05)) +
@@
   theme_np_dag() +
-  theme(plot.title = element_text(hjust = 0.5, color = clrs$purple))
+  theme(plot.title = element_text(hjust = 0.5, color = clrs$Prism[1]))
 
@@
   tidy_dagitty() |> 
-  mutate(arrow_color = ifelse(to == "Z", clrs$red, "grey80")) |> 
+  mutate(arrow_color = ifelse(to == "Z", clrs$Prism[8], "grey80")) |> 
   mutate(node_color = ifelse(name == "Z", TRUE, FALSE))
@@
   geom_dag_text(data = filter(data_collider, name == "Z"), color = "white", size = 4) +
-  scale_color_manual(values = c("grey80", clrs$red), guide = "none") +
+  scale_color_manual(values = c("grey80", clrs$Prism[8]), guide = "none") +
   coord_cartesian(ylim = c(0.95, 2.05)) +
@@
   theme_np_dag() +
-  theme(plot.title = element_text(hjust = 0.5, color = clrs$red))
+  theme(plot.title = element_text(hjust = 0.5, color = clrs$Prism[8]))
 
@@
 Confounding is a major scary issue though - especially if it's unmeasured or unobservable…
+
+Experiential learning → community connections → long-term pro-social outcomes
+
+Experiential learning → MPA students ← Pro-social behaviors
+
+Collider bias is often seen as a form of selection bias, leading to results that are not generalizable to other contexts.
+Colliders can lead to bias beyond issues of selection, though.
+KnoxLoweMummolo use a DAG to argue that police stops, race, administrative records, etc.
 
@@
   node_status() |> 
-  mutate(arrow_color = ifelse(name == "X", clrs$red, "grey80"))
+  mutate(arrow_color = ifelse(name == "X", clrs$Prism[8], "grey80"))
 
@@
   annotate(geom = "text", x = 1, y = 1.2, label = "X = x") +
-  scale_color_manual(values = c(clrs$green, clrs$blue), 
-                     na.value = clrs$yellow, guide = "none") +
+  scale_color_manual(values = c(clrs$Prism[4], clrs$Prism[2]), 
+                     na.value = clrs$Prism[6], guide = "none") +
   coord_cartesian(ylim = c(0.95, 2.05)) +
@@
   geom_dag_text(size = 4) +
-  scale_color_manual(values = c(clrs$green, "grey80", clrs$blue), 
-                     na.value = clrs$yellow, guide = "none") +
+  scale_color_manual(values = c(clrs$Prism[4], "grey80", clrs$Prism[2]), 
+                     na.value = clrs$Prism[6], guide = "none") +
   scale_shape_manual(values = c(15, 19), guide = "none") +
@@
                 nudge_y = -0.37, size = 4) +
-  scale_color_manual(values = c(clrs$green, clrs$blue), 
-                     na.value = clrs$yellow, guide = "none") +
+  scale_color_manual(values = c(clrs$Prism[4], clrs$Prism[2]), 
+                     na.value = clrs$Prism[6], guide = "none") +
   scale_shape_manual(values = c(15, 19), guide = "none") +
@@
   node_status() |> 
-  mutate(arrow_color = ifelse(name == "X", clrs$red, "grey80"),
+  mutate(arrow_color = ifelse(name == "X", clrs$Prism[8], "grey80"),
          name = recode(name, "Running" = "Running\nvariable"),
@@
                 nudge_y = -0.35, size = 4) +
-  scale_color_manual(values = c(clrs$green, "grey80", clrs$blue), 
-                     na.value = clrs$yellow, guide = "none") +
+  scale_color_manual(values = c(clrs$Prism[4], "grey80", clrs$Prism[2]), 
+                     na.value = clrs$Prism[6], guide = "none") +
   scale_shape_manual(values = c(15, 19), guide = "none") +

### 2024-11-22 methods/diff-in-diff.qmd
Bigger font for slides

@@
   labs(x = NULL, y = "Prosocial index", color = NULL) +
-  theme_np()
+  theme_np(base_size = 16)
 ```
@@
   labs(x = "Treatment effect", y = NULL) +
-  theme_np() +
+  theme_np(base_size = 16) +
   theme(panel.grid.major.y = element_blank())

### 2024-11-22 methods/rdd-sharp.qmd
Bigger font for slides

@@
   scale_color_manual(values = c(clrs$Prism[2], clrs$Prism[8])) +
-  theme_np()
+  theme_np(base_size = 16)
 ```

### 2024-11-22 presentations/2024-11-22_arnova/index.qmd
Add embedded google presentation

---
title: "ARNOVA"
date: 2024-11-22
toc: false
resources: slides
abstract: |
  Annual conference of the Association for Research on Nonprofit Organizations and Voluntary Action (ARNOVA); Washington, D.C.
abstract-title: "Location"
published-title: "Presented"
image: first-slide.png
page-layout: full
---

<div class="ratio ratio-16x9">
<iframe class="slide-deck" src="https://docs.google.com/presentation/d/e/2PACX-1vRTS3Uve-PLfIOJjtIFIQgMFz1Zs3HhArdtNvgFd6VGnA_kuxqrRelNHHE2Tv4CFRty77ZB5EdkNRon/embed?start=false&loop=false&delayms=3000" allowfullscreen="true" mozallowfullscreen="true" webkitallowfullscreen="true"></iframe>
</div>

### 2024-11-22 methods/iv.qmd
Add IV randomized promotion data

--- 2024-11-22 methods/rct.qmd
@@
 ---
-title: "RCT"
+title: "Instrumental variables"
 ---
@@
 library(tinytable)
-library(ggdist)
 library(scales)
@@
 library(parameters)
+library(estimatr)
 
@@
 
-rct <- tar_read(data_rct)
+iv <- tar_read(data_iv)
 
-# Plotting functions
 invisible(list2env(tar_read(graphic_functions), .GlobalEnv))
@@
 
-## DAG
-
 ```{r}
-#| label: plot-dag-rct
-#| fig-width: 6
-#| fig-height: 6
-
-coords <- tribble(
-  ~name, ~x,   ~y,   ~type,         ~color,         ~label,
-  "X",   1,    3,    "Treatment",   clrs$Prism[4],  "Experiential\nphilanthropy",
-  "Y",   3,    3,    "Outcome",     clrs$Prism[2],  "Charitable\nbehavior",
-  "Z1",  5/3,  4.2,  "Confounder",  clrs$Prism[6],  "Social awareness",
-  "Z2",  7/3,  4.2,  "Confounder",  clrs$Prism[6],  "Age",
-  "Z3",  5/3,  5,    "Confounder",  clrs$Prism[6],  "GPA",
-  "Z4",  7/3,  5,    "Confounder",  clrs$Prism[6],  "Income",
-  "M",   2,    2,    "Mediator",    clrs$Prism[3],  "Community\nconnections",
-  "C",   2,    1,    "Collider",    clrs$Prism[8],  "Employment in\nnonprofit sector"
-) |> 
-  mutate(type = fct_inorder(type))
-
-data_full_dag <- dagify(
-  Y ~ X + Z1 + Z2 + Z3 + Z4 + M,
-  M ~ X,
-  C ~ X + Y,
-  coords = coords,
-  labels = pull(coords, label, name = name)
-) |> 
-  tidy_dagitty() |> 
-  left_join(select(coords, name, type, color), by = join_by(name)) |> 
-  mutate(arrow_color = case_when(
-    name == "X" & to == "Y" ~ "black",
-    to == "C" ~ clrs$Prism[8],
-    to == "M" | (name == "M" & to == "Y") ~ clrs$Prism[3],
-    .default = color
-  ))
-
-ggplot(data_full_dag, aes(x = x, y = y, xend = xend, yend = yend)) +
-  geom_dag_edges(edge_width = 0.5) +
-  geom_dag_edges(aes(edge_color = arrow_color), edge_width = 0.5) +
-  geom_dag_point(aes(color = type), size = 12) +
-  geom_dag_label(
-    data = filter(data_full_dag, type != "Confounder"), aes(label = label), 
-    size = 4, nudge_y = -0.46, color = "black", lineheight = 1
-  ) +
-  geom_dag_label(
-    data = filter(data_full_dag, type == "Confounder"), aes(label = label), 
-    size = 4, nudge_y = 0.35, color = "black", lineheight = 1
-  ) +
-  scale_color_manual(
-    values = coords |> distinct(type, color) |> pull(color, name = type),
-    guide = guide_legend(
-      title = NULL, override.aes = list(size = 6), label.position = "bottom"
-    )
-  ) +
-  coord_cartesian(xlim = c(0.98, 3.02)) +
-  theme_np_dag() + 
-  theme(legend.position = "top")
-```
-
-## EDA
-
-### Confounders
-
-```{r}
-ggplot(rct, aes(x = social_awareness, y = experiential_learning)) +
-  geom_dots(aes(side = ifelse(experiential_learning == 1, "bottom", "top")),
-    pch = 19, color = "grey20", scale = 0.2
-  ) +
-  geom_smooth(
-    method = "glm", method.args = list(family = binomial(link = "logit")),
-    se = FALSE
-  ) +
-  scale_y_continuous(labels = label_percent()) +
-  labs(x = "Social awareness index", y = "Experiential learning participation")
-
-ggplot(rct, aes(x = age, y = experiential_learning)) +
-  geom_dots(aes(side = ifelse(experiential_learning == 1, "bottom", "top")),
-    pch = 19, color = "grey20", scale = 0.4
-  ) +
-  geom_smooth(
-    method = "glm", method.args = list(family = binomial(link = "logit")),
-    se = FALSE
-  ) +
-  scale_y_continuous(labels = label_percent()) +
-  labs(x = "Age", y = "Experiential learning participation")
-
-ggplot(rct, aes(x = gpa, y = experiential_learning)) +
-  geom_dots(aes(side = ifelse(experiential_learning == 1, "bottom", "top")),
-    pch = 19, color = "grey20", scale = 0.4
-  ) +
-  geom_smooth(
-    method = "glm", method.args = list(family = binomial(link = "logit")),
-    se = FALSE
-  ) +
-  scale_y_continuous(labels = label_percent()) +
-  labs(x = "GPA", y = "Experiential learning participation")
-
-ggplot(rct, aes(x = income, y = experiential_learning)) +
-  geom_dots(aes(side = ifelse(experiential_learning == 1, "bottom", "top")),
-    pch = 19, color = "grey20", scale = 0.4
-  ) +
-  geom_smooth(
-    method = "glm", method.args = list(family = binomial(link = "logit")),
-    se = FALSE
-  ) +
-  scale_x_continuous(labels = label_dollar()) +
-  scale_y_continuous(labels = label_percent()) +
-  labs(x = "Income", y = "Experiential learning participation")
-```
-
-```{r}
-ggplot(rct, aes(x = social_awareness, y = prosocial_index)) +
-  geom_point() +
-  geom_smooth()
-
-ggplot(rct, aes(x = age, y = prosocial_index)) +
-  geom_point() +
-  geom_smooth()
-
-ggplot(rct, aes(x = gpa, y = prosocial_index)) +
-  geom_point() +
-  geom_smooth()
-
-ggplot(rct, aes(x = income, y = prosocial_index)) +
-  geom_point() +
-  geom_smooth()
-```
-
-### Mediator
-
-```{r}
-rct |> 
-  mutate(experiential_learning_fct = as.logical(experiential_learning)) |> 
-  ggplot(aes(x = experiential_learning_fct, y = community_connections)) +
-  geom_point(position = position_jitter(height = 0)) +
-  geom_boxplot(aes(group = experiential_learning_fct))
-
-ggplot(rct, aes(x = community_connections, y = prosocial_index)) +
-  geom_point() +
-  labs(x = "Community connections", y = "Prosocial index")
-```
-
-### Collider
-
-```{r}
-rct |> 
-  mutate(experiential_learning_fct = as.logical(experiential_learning)) |> 
-  mutate(employed_nonprofit_fct = as.logical(employed_nonprofit)) |> 
-  ggplot(aes(x = experiential_learning_fct, y = employed_nonprofit_fct)) +
-  geom_point(position = position_jitter())
-
-rct |> 
-  mutate(employed_nonprofit_fct = as.logical(employed_nonprofit)) |> 
-  ggplot(aes(x = prosocial_index, y = employed_nonprofit_fct)) +
-  geom_point(position = position_jitter(width = 0))
-```
-
-## Model comparisons
-
-```{r}
-model_correct <- lm(
-  prosocial_index ~ experiential_learning, 
-  data = rct
+model_iv_basic <- iv_robust(
+  prosocial_index ~ experiential_learning | promoted, 
+  data = iv
 )
 
-model_adjusted <- lm(
-  prosocial_index ~ experiential_learning + social_awareness + age + gpa + income,
-  data = rct
+model_iv_controls <- iv_robust(
+  prosocial_index ~ experiential_learning + social_awareness + age + gpa + income | 
+    promoted + social_awareness + age + gpa + income, 
+  data = iv
 )
 
-model_with_mediator <- lm(
-  prosocial_index ~ experiential_learning + social_awareness + age + gpa + income + community_connections, 
-  data = rct
+model_naive <- lm(
+  prosocial_index ~ experiential_learning, 
+  data = iv
 )
 
-model_collider_selection <- lm(
+model_naive_controls <- lm(
   prosocial_index ~ experiential_learning + social_awareness + age + gpa + income, 
-  data = filter(rct, employed_nonprofit == 1)
-)
-
-model_collider_post_treatment <- lm(
-  prosocial_index ~ experiential_learning + social_awareness + age + gpa + income + employed_nonprofit, 
-  data = rct
+  data = iv
 )
@@
 models <- list(
-  "Treatment only" = model_correct,
-  "Covariates included" = model_adjusted,
-  "Mediator included" = model_with_mediator,
-  "Collider (selection)" = model_collider_selection,
-  "Collider (post-treatment covariate)" = model_collider_post_treatment
-)
-```
-
-```{r}
-modelsummary(
-  models,
-  statistic = "[{conf.low}, {conf.high}]",
-  fmt = 2,
-  gof_map = gof_map,
-  add_columns = tibble("True effect" = c(NA, NA, 10)) |> magrittr::set_attr("position", 2)
-) |> 
-  style_tt(
-    i = 3:4, j = 1:7, 
-    background = clrs$Prism[6], color = "#ffffff", bold = TRUE
+  "IV" = model_iv_basic,
+  "IV with controls" = model_iv_controls,
+  "Naive OLS" = model_naive,
+  "OLS with controls" = model_naive_controls
 )

### 2024-11-22 R/generate-iv.R

@@ -1,0 +1,61 @@
+make_iv <- function(n, seed) {
+  withr::with_seed(seed, {
+    simulated_data_raw <- tibble(
+      id = as.integer(1:n + 999)
+    ) |>
+
+      mutate(
+        social_awareness = 100 * rbeta(n, shape1 = 7, shape2 = 3),
+        age = rnorm(n, mean = 25, sd = 2.5),
+        gpa = pmin(pmax(rnorm(n, mean = 3.25, sd = 0.6), 2.0), 4.0),
+        income = pmax(rnorm(n, mean = 40000, sd = 15000), 15000)
+      ) |>
+
+      # Generate random promotion
+      mutate(
+        promoted = rbinom(n, 1, 0.5)
+      ) |>
+
+      # Generate treatment variable with confoudners and promotion
+      mutate(
+        treatment_logit = 
+          -1.2 +  # Logit intercept (base rate, plogis(-1.2) = 23%)
+          1.4 * promoted +  # Promotion increases the probability
+          0.015 * social_awareness +
+          0.08 * (age - 25) +
+          0.4 * (gpa - 3.25) +
+          0.00002 * (income - 40000),
+        treatment_prob = plogis(treatment_logit),
+        experiential_learning = rbinom(n, 1, treatment_prob)
+      ) |>
+
+      # Generate outcome
+      mutate(
+        prosocial_base = 
+          50 +  # OLS intercept
+          0.45 * social_awareness +
+          0.6 * (age - 25) +
+          3.3 * (gpa - 3.25) +
+          0.0002 * (income - 40000) +
+          11.25 * experiential_learning,  # Treatment effect
+        prosocial_index = pmin(pmax(
+          prosocial_base + rnorm(n, 0, 6),
+          0), 100
+        )
+      )
+
+    simulated_data <- simulated_data_raw |>
+      mutate(
+        across(c(prosocial_index, social_awareness, age),
+               \(x) as.integer(round(x, 0))),
+        gpa = round(gpa, 2),
+        income = as.integer(signif(income, 3))
+      ) |>
+      select(
+        id, promoted, experiential_learning, prosocial_index,
+        social_awareness, age, gpa, income
+      )
+
+    return(simulated_data)
+  })
+}

### 2024-11-22 _targets.R

@@ -56,6 +56,7 @@ list(
   tar_target(data_did, make_diff_in_diff(n_per_school_year = 126, seed = 417724)),
   tar_target(data_rdd_sharp, make_rdd(n = 967, seed = 635404, fuzzy = FALSE)),
   tar_target(data_rdd_fuzzy, make_rdd(n = 992, seed = 995303, fuzzy = TRUE)),
+  tar_target(data_iv, make_iv(n = 914, seed = 266416)),
 
   ## Save data -----
   # TODO: Use static/dynamic branching here instead of this gross repetitive mess

### 2024-11-23 manuscript/manuscript.qmd
Add website as target + include manuscript in website

@@
 short-title: "Clarifying Correlation and Causation"
-date: 2024-11-08
+date: 2024-11-22
 author:
@@
 
-# Why care about causation in nonprofit studies? (MENG)
+# Why care about causation in nonprofit studies?
 

### 2024-11-23 paper/index.qmd
Add website as target + include manuscript in website

---
title: Paper
toc: false
page-layout: full
---

```{r include=FALSE}
# Add the manuscript target as a dependency to the overall website
targets::tar_config_set(
  store = here::here('_targets'),
  script = here::here('_targets.R')
)

targets::tar_load(c(manuscript))
```

```{=html}
<p class="text-center">
  <a class="btn btn-success btn-sm mb-1" href="/manuscript/output/index.html" target="manuscript">
  <i class="fa-solid fa-file-lines"></i> &ensp;HTML
  </a>

  <a class="btn btn-success btn-sm mb-1" href="/manuscript/output/index.html" target="_blank">
  <i class="fa-solid fa-arrow-up-right-from-square"></i>&ensp;HTML <small>(new window)</small>
  </a>

  <a class="btn btn-danger btn-sm mb-1" href="/manuscript/output/heiss-ye-causation.pdf" target="manuscript">
  <i class="fa-solid fa-file-pdf"></i>&ensp;PDF
  </a>

  <a class="btn btn-danger btn-sm mb-1" target="_blank" href="/manuscript/output/heiss-ye-causation.pdf">
  <i class="fa-solid fa-file-arrow-down"></i>&ensp;PDF <small>(download)</small>
  </a>

  <a class="btn btn-warning btn-sm mb-1" href="/manuscript/output/heiss-ye-causation-ms.pdf" target="manuscript">
  <i class="fa-solid fa-file-pdf"></i>&ensp;Manuscripty PDF
  </a>

  <a class="btn btn-warning btn-sm mb-1" target="_blank" href="/manuscript/output/heiss-ye-causation-ms.pdf">
  <i class="fa-solid fa-file-arrow-down"></i>&ensp;Manuscripty PDF <small>(download)</small>
  </a>
</p>

<iframe src="/manuscript/output/index.html" frameborder="0" style="overflow:hidden;height:90%;width:100%;border:2px solid #ccc;" height="90%" width="100%" name="manuscript"></iframe>
```

### 2024-11-23 _targets.R

@@ -38,6 +38,9 @@ here_rel <- function(...) {fs::path_rel(here::here(...))}
 # the pipeline
 tar_source()
 
+# Set some conditional flags
+should_deploy <- identical(Sys.getenv("UPLOAD_WEBSITES"), "TRUE")
+
 
 # Pipeline ----------------------------------------------------------------
 
@@ -108,6 +111,17 @@ list(
 
   ## Manuscript targets ----
   tar_quarto(manuscript, path = "manuscript", working_directory = "manuscript", quiet = FALSE),
+
+  ## Website targets ----
+  tar_quarto(website, path = ".", quiet = FALSE),
+
+  tar_target(deploy_script, here_rel("deploy.sh"), format = "file", cue = tar_cue_skip(!should_deploy)),
+  tar_target(deploy, {
+    # Force a dependency
+    website
+    # Run the deploy script
+    if (should_deploy) processx::run(paste0("./", deploy_script))
+  }, cue = tar_cue_skip(!should_deploy)),
 
   ## Render the README ----
   tar_quarto(readme, here_rel("README.qmd"))

### 2025-12-31 manuscript/manuscript.qmd
Add epigraph + some other notes

@@
 short-title: "Clarifying Correlation and Causation"
-date: 2024-11-22
+date: 2025-04-24
 author:
@@
 published: Drafting
-abstract: |
+abstract11: |
   Discovering causal relationships and testing theoretical mechanisms is a core endeavor of social science.
@@
   Acknowledgments here if any
-epigraph:
-  - text: "Don't let us forget that the causes of human actions are usually immeasurably more complex and varied than our subsequent explanations of them."
-    source: "Fyodor Dostoevsky, *The Idiot*, part IV, chapter 2"
-keywords:
+keywords11:
   - those
@@
 
+{{< epigraph "Scientia et potentia humana in idem coincidunt, quia ignoratio causae destituit effectum.
+*(Human knowledge and human power meet in one; for where the cause is not known the effect cannot be produced.)*" source="Francis Bacon, *Novum Organum*, part I, aphorism III" >}}
+
+{{< epigraph "Don't let us forget that the causes of human actions are usually immeasurably more complex and varied than our subsequent explanations of them." source="Fyodor Dostoevsky, *The Idiot*, part IV, chapter 2" >}}
+
 Causal inference important in social sciences, but hard. Nobel prize for Angrist and Imbens and gang...
@@
 
+Qualitative case studies: Rohlfing:2012, GeorgeBennett:2005, BennettCheckel:2015
+
 Instead, we address the perennial caveat taught in introductory statistics classes: that correlation does not imply causation.
@@
 How can we use statistical tools to explore causal questions using observational data?
+
+---
+
+From thread by Stephen Wild (https://bsky.app/profile/stephenjwild.bsky.social/post/3ldhouhkcsk2z)
+
+- Causal inference is hard: https://www.nature.com/articles/s41562-024-01939-z
+- Always clearly define your estimand: https://journals.sagepub.com/doi/10.1177/00031224211004187
+- Threats to validity (with DAGs): https://journals.lww.com/epidem/fulltext/2020/05000/A_Graphical_Catalog_of_Threats_to_Validity_.11.aspx
+- Crash course in good and bad controls: https://journals.sagepub.com/doi/full/10.1177/00491241221099552
+- Need to think about causality when selecting control variables: https://journals.sagepub.com/doi/full/10.1177/25152459221095823
+- Conditioning on colliders is generally bad: https://www.annualreviews.org/content/journals/10.1146/annurev-soc-071913-043455
 
@@
 Bad controls and colliders
+
+For example, in 2022, using a large national-level observational dataset, researchers found an association between smoking behavior and reduced COVID-19 risk: people who smoked regularly saw a sigificantly reduction in the probability of contracting severe COVID-19.
+However, this paradoxical finding can be explained with collider bias [@AuYeungKwok:2022].
+The dataset contained information about hospital patients who were (1) smokers diagnosed with COVID, (2) smokers diagnosed with smoking-related respiratory disease, and (3) non-smokers diagnosed with COVID.
+The data did not contain any patients who were non-smokers diagnosed with non-COVID, smoking-related respiratory disease, as these people were not patients and needed no medical treatment.
+In DAG terms, presence at the hospital was a collider (Smoking → Hospital ← Severe COVID), and conditioning the estimate of Smoking → Severe COVID on being a patient therefore distorted and biased the measured effect.
 
